### 2024-10-29 paper.qmd
Initial commit

---
title: ""
format: 
  docx:
    reference-doc: "ref.docx"
    fig-format: pdf
    fig-asp: 0.75
    fig-width: 7
bibliography: "misc/ref.bib"
csl: "misc/frontiers-in-physiology.csl"
editor_options: 
  chunk_output_type: console
execute: 
  echo: false
  warning: false
  error: false
---

**Title**: Modeling Nonlinear R-R Interval Dynamics during Exercise: A Novel Logistic Approach to Cardiovascular Recovery.

**Authors**: Matías Castillo-Aguilar^1,2^, [...], Cristian Núñez-Espinoza^1,3^\*.

::: {custom-style="Afiliation"}
^1^ Centro Asistencial Docente e Investigación (CADI-UMAG), University of Magallanes, Punta Arenas, Chile.

^2^ Kinesiology Department, University of Magallanes, Punta Arenas, Chile.

^3^ School of Medicine, University of Magallanes, Punta Arenas, Chile.
:::

### \*Correspondence:

Cristian Núñez-Espinosa, School of Medicine, University of Magallanes, Punta Arenas, Chile. Centro Asistencial de Docencia e Investigación CADI-UMAG, Chile. e-mail: [cristian.nunez\@umag.cl](mailto:cristian.nunez@umag.cl). Address: Avenida Bulnes 01855, Box 113-D. Phone: +56 61 2201411.

{{< pagebreak >}}

## Abstract

**Objective**: [...]. **Methods**: [...]. **Results**: [...]. **Conclusion**: [...].

**Keywords**: [...].

{{< pagebreak >}}

# Introduction

Current research has delved deep into understanding the underlying mechanisms behind cardiac autonomic dynamics in response to exercise, and how this could be related to health-related quality of life and an increased risk of cardiovascular diseases. Hence, shedding light on the autonomic mechanics can provide insights into the role play between modifying parameters of exercise-induced response and how to optimize it to achieve optimal effects in both young and older individuals.

The study of R-R intervals (RRi) in response to exercise is a key research area within this context, given its significant implications for cardiovascular health, athletic performance, and overall physiological adaptation [@kristal1995heart; @thayer2010relationship; @dong2016role; @lundstrom2023practices]. Unlike measures of heart rate variability (HRV), which summarize autonomic activity over time, RRi offers a more direct and granular view of the heart's electrical activity during or immediately after an exercise test in older individuals [@mongin2022hrv; @castillo2023validity; @mabe2024associations]. Analyzing the temporal dynamics of RRi (i.e., the time between successive heartbeats) provides invaluable insights into how the cardiovascular system behaves to and recovers from physical stressors, such as fatigue after competition or an intense exercise test [@castillo2021cardiac; @eser2022exercise].

Understanding these fluctuations is particularly relevant during dynamic periods of exercise when the autonomic nervous system (ANS) shifts between parasympathetic withdrawal and sympathetic activation [@boettger2010heart]. Modeling RRi dynamics, rather than relying on broader HRV metrics, allows researchers to focus on the direct physiological markers of autonomic adaptation to stress [@hautala2003short]. This detailed perspective is critical for identifying recovery patterns and understanding cardiovascular reactivity in athletes and individuals with varied fitness levels [@mongin2022hrv].

Despite the importance of RRi analysis, modeling their nonlinear behavior during and after exercise presents significant challenges. Traditional approaches, such as linear regression and time-series analysis, often fail to capture the intricate transitions in RRi, especially during intense physical exertion and recovery. This limitation is due to the non-linear and non-stationary nature of HRV [@gronwald2019effects]. Linear models, which oversimplify these transitions, are prone to misrepresent the actual physiological processes. More advanced non-linear models have begun to address this issue, yet many focus on overall HRV parameters rather than directly modeling the RRi [@gronwald2019non].

Non-linear models have gained increasing attention for analyzing the complex dynamics of RRi, especially during exercise. These models offer a more comprehensive understanding of HRV compared to traditional linear approaches [@wu2003nonlinear]. Recent studies have proposed exponential decay models to describe RRi recovery, while others have employed logistic functions to depict the gradual return to baseline following periods of heightened physical stress [@molkkari2020online; @gronwald2019non]. These models offer an advantage over traditional HRV analyses by providing a more detailed understanding of the cardiovascular system's response to exercise. However, despite these advancements, few models explicitly target real-time RRi dynamics, and fewer still incorporate the variability introduced by individual physiological differences, such as fitness level, autonomic balance, and exercise intensity [@kanniainen2023estimation].

Given the unique characteristics of RRi, their direct relationship with cardiac electrical activity, and responsiveness to autonomic changes, there is a compelling need for a model that captures their non-linear fluctuations during exercise and recovery. Such a model would offer a more biologically relevant representation of the heart’s behavior than the broader HRV indices typically used in research [@bacopoulou2021entropy].

Hence, the primary objective of this paper is to present a novel non-linear model that characterizes the fluctuations in RRi during exercise and recovery. This model is specifically designed to reflect the complex, real-time changes in RRi, offering a more precise understanding of the physiological processes underlying cardiovascular adaptation to exercise. By focusing exclusively on RRi, the model can deliver insights directly applicable to athletic training regimens, recovery protocols, and clinical practices aimed at improving cardiovascular health. The proposed model seeks to bridge the gap between existing modeling frameworks and the physiological reality of RRi dynamics during exercise.

# Methods

```{r config}
library(data.table)
library(ggplot2)
library(brms)
```

## Model Specification

The mathematical model proposed to characterize the RRi response to exercise and recovery is defined by @eq-main-model.

$$
\begin{aligned}
\text{RRi}(t) = \alpha + \frac{\beta}{1 + e^{\lambda(t - \tau)}} + \frac{-\beta \cdot c}{1 + e^{\phi(t - \tau - \delta)}}
\end{aligned}
$${#eq-main-model}

This model includes two logistic functions representing the RRi dynamics across exercise and recovery phases. The first logistic term models the decrease in RRi during exercise, where the parameter $\beta$ denotes the magnitude of this decline. The rate of decline is governed by $\lambda$, while $\tau$ represents the onset of the RRi decrease or the time at which the physiological shift begins.

The second logistic term accounts for RRi recovery post-exercise. Here, $c$ scales the magnitude of recovery relative to the initial decline represented by $\beta$, effectively capturing the proportion of the decline regained during recovery. The rate at which RRi returns to baseline is controlled by $\phi$, and $\delta$ indicates the lag following the cessation of exercise, marking the beginning of recovery.

This logistic structure is well-suited to modeling the RRi dynamics, providing a smooth, continuous transition for the decline and recovery phases. Logistic growth functions are particularly effective in physiological modeling contexts, where transitions (e.g., rest to exercise or exercise to recovery) occur gradually and non-linearly. Compared to conditional models, which may introduce abrupt transitions, this model is designed to minimize discontinuities, thus offering a realistic representation of RRi responses without artifacts.

## Sensitivity to model parameters

To assess the sensitivity of model parameters in influencing RRi over time, we implemented a Sobol sensitivity analysis using Monte Carlo simulations. To compute Sobol indices, a total of 500 Monte Carlo simulations were conducted, with each simulation involving 500 randomly sampled parameter sets (250,000 model runs in total). For each set of parameters, RR intervals were calculated at each time point $t$ across a range from 0 to 20 minutes at intervals of 0.1 minutes. The resulting Sobol indices provided a measure of the contribution of each parameter to the variance in RR intervals at each time point.

The first-order Sobol index for each parameter was computed by isolating the variance attributable to each parameter while averaging over the others. To achieve this, we perturbed each parameter individually while holding all other parameters at their average values across the samples. The proportion of variance explained by each parameter at each time point was calculated by dividing the variance of the perturbed model outputs by the total variance, yielding time-dependent Sobol indices. The selected parameter ranges, provided in [Table 1](#tbl-sens-params), reflect a 50% variation in both directions from a reference value for each parameter.

|  | $\alpha$| $\beta$| $c$| $\lambda$| $\phi$| $\tau$| $\delta$|
|----|--------:|-------:|----:|-------:|-------:|------:|--------:|
|     x 0.5 |  400| -200| 0.5| -1.151| -0.602| 2.5| 1|
| Reference |  800| -400| 1.0| -2.303| -1.203|   5| 2|
|     x 1.5 | 1200| -600| 1.5| -3.454| -1.806| 7.5| 3|

[**Table 1**]{#tbl-sens-params}. Parameter ranges for sensitivity analysis to be used as the limits to sample from uniform distribution and designed to evaluate model response to variability. Each parameter's range reflects a ±50% variation from its reference value. It is worth noting that reference values for $\lambda$ and $\phi$ parameters were set considering a reference rate of 90% and 70% change per minute in the original scale of these parameters.

## Usage of real-world RRi data

To further assess the performance and applicability of the proposed model, real-world RRi data were analyzed in addition to the synthetic data generated through simulation. This dataset was derived from a cohort participating in the FONDECYT Project No. 11220116, funded by the Chilean National Association of Research and Development (ANID). Ethical approval was granted by the Ethics Committee of the University of Chile (ACTA No. 029-18/05/2022) and the Ethics Committee of the University of Magallanes (No. 008/SH/2022).

The dataset consisted of 272 participants who underwent a validated exercise protocol encompassing rest, exercise, and recovery phases within a single, continuous measurement session [@castillo2023validity]. Continuous heart rate data, including RRi, were collected using the Polar Team2 system (Polar®) application, which is capable of capturing dynamic fluctuations associated with varying exercise intensities and recovery.

Preprocessing steps were conducted to remove artifacts and ectopic heartbeats, with less than 3% of data excluded following established guidelines [@malik1996heart]. The preprocessed RRi data were then aggregated into time intervals to facilitate analysis, allowing the examination of acute exercise responses as well as post-exercise recovery patterns.

This real-world dataset provided a critical context for validating the model’s predictive capability against observed physiological responses, offering a robust foundation for understanding RRi dynamics under conditions of physical activity.

## Parameter Estimation

Parameter estimation was performed using Hamiltonian Monte Carlo (HMC) with the No-U-Turn Sampler (NUTS) to explore the parameter space. This method is suitable for high-dimensional spaces and utilizes gradient information for efficient sampling. The parameters $\alpha$, $\beta$, $c$, $\lambda$, $\phi$, $\tau$, and $\delta$ were estimated by sampling from the posterior distribution, which was constructed from observed RRi data and model predictions. The Bayesian framework allows the incorporation of prior distributions for parameters, enhancing the reliability of the estimates.

The priors were selected based on physiological constraints and are detailed in @eq-priors.

$$
\begin{aligned} 
\alpha &\sim \mathcal{N}(850, 50) \text{ with } \alpha \geq 0 \\
\beta &\sim \mathcal{N}(-400, 50) \text{ with } \beta \leq 0 \\
c &\sim \mathcal{N}(1, 0.2) \text{ with } c \geq 0 \\
\lambda &\sim \mathcal{N}(-3, 3) \text{ with } \lambda \leq 0 \\
\phi &\sim \mathcal{N}(-3, 3) \text{ with } \phi \leq 0 \\
\tau &\sim \mathcal{N}(5, 0.5) \text{ with } \tau \geq 0 \\
\delta &\sim \mathcal{N}(5, 0.5) \text{ with } \delta \geq 0
\end{aligned}
$$ {#eq-priors}

Simulated RRi dynamics based on the prior parameter distributions are shown in [Figure 1](#fig-prior-sim).

![](figures/prior-model.pdf)

[**Figure 1**]{#fig-prior-sim}. Predicted RRi response to exercise, illustrating uncertainty in model parameters from the prior distribution. Shaded areas represent 95%, 80%, and 60% quantile CI.

The gradient of the log-likelihood function for each parameter was computed during estimation using the brms R package, which employs the Stan probabilistic programming language. Convergence of the HMC chains was evaluated using the Gelman-Rubin diagnostic and trace plots to ensure reliable parameter estimates. This quantification of uncertainty is essential for assessing the stability of parameter estimates, particularly in the presence of noisy RRi data.

The fitting process utilized four Markov Chain Monte Carlo (MCMC) chains, each consisting of 12,500 iterations with a burn-in period of 5,000 iterations, resulting in a total of 30,000 post-warmup samples.

## Model Evaluation

The model's predictive performance was assessed using approximate leave-one-out (LOO) cross-validation for Bayesian models, utilizing Pareto-smoothed importance sampling. The primary evaluation metrics included root mean square error (RMSE) and mean absolute percentage error (MAPE). These metrics were selected for their capacity to quantify both absolute and relative errors, thereby providing a comprehensive assessment across the varying scales of RRi dynamics, including resting baseline, peak decline, and recovery.

Also, residual analysis was conducted to evaluate the model’s accuracy in capturing RRi dynamics. Residuals were defined as the difference between observed and predicted RRi values. These residuals were analyzed for temporal structure and autocorrelation to ensure that no systematic patterns remained in the errors. This indicates that the model has sufficiently captured the underlying dynamics of the RRi response to exercise.

# Results

## Non-linear model and deterministic behaviour

### RRi as a linear combination of logistic functions

According to the proposed model in @eq-main-model, the dynamics of RRi in response to physical exertion can be represented as a linear combination of a baseline RRi $\alpha$ and two logistic functions, denoted as $f_1(t)$ and $f_2(t)$. The function $f_1(t)$ models the initial decay in RRi following the initiation of exercise while $f_2(t)$ characterizes the recovery phase after exercise cessation.

The fundamental structure of both logistic functions can be expressed as:

$$
\begin{aligned}
f(t) = \frac{a_1}{1 + e^{a_2(t - a_3)}}
\end{aligned}
$${#eq-logistic-structure}

In this equation, $a_1$ represents the asymptotic value approached by the logistic function, which can be either positive (indicating an increase) or negative (indicating a decrease). For $f_1(t)$, this parameter is specified as $\beta$, indicating the absolute change in RRi at the onset of exercise. In contrast, for $f_2(t)$, $a_1$ is reparameterized as $-\beta \cdot c$, where $c$ denotes the proportion of change relative to the initial drop indicated by $\beta$. This reparameterization ensures that, after the initial decline, the second logistic function facilitates the return of RRi toward the baseline value $\alpha$.

```{r}
#| eval: false
ln_to_perc <- function(ln) 1 - exp(ln)
perc_to_ln <- function(perc) log(1 - perc)

perc_to_ln(0.9) |> # log(1 - 0.9) ==> Log-scale
  ln_to_perc()     # 1 - exp(-2.302585) ==> Percentage-change
```

The parameter $a_2$ defines the rate at which the specified increase or decrease occurs. This rate parameter is expressed on a logarithmic scale; to convert it to a percentage change per unit of time, it can be scaled as $1 - \exp(a_2)$. For instance, a 90% decrease per unit time corresponds to $a_2 = \log(1 - 0.9)$, resulting in an approximate value of -2.302585.

The parameter $a_3$ serves as an activation threshold, causing the value within the exponential function, and consequently, the value in the denominator, to increase significantly until reaching $a_3$. Beyond this point, the denominator approaches 1, allowing the logistic function to attain the asymptotic level determined by the numerator. The behavior of the model constituents is illustrated in [Figure 2](#fig-linear-constituents).

![](figures/model-constituents.pdf)

[**Figure 2**]{#fig-linear-constituents}. The RRi dynamics in response to exercise are expressed as a linear combination of model constituents based on the baseline RRi $\alpha$ and two logistic functions, denoted $f_1(t)$ and $f_2(t)$, respectively.

### Sensitivity to parameter variability

Sobol sensitivity analysis reveals that the parameter $\alpha$ exerts the most substantial influence on the model's output, followed by parameters $c$ and $\delta$. In contrast, parameters $\beta$, $\lambda$, and $\phi$ demonstrate relatively minor effects, with some values crossing zero, indicating negligible influence within the tested parameter ranges.

Individual perturbation of each parameter highlighted that RRi dynamics are sensitive to the baseline RRi parameter, $\alpha$. Conversely, the rate parameters for the initial decay during exercise, $\lambda$, and the recovery post-exercise, $\phi$, show lower sensitivity, suggesting that they are not primary sources of variation in predicted RRi trajectories when assessed in isolation. The results of the sensitivity analysis are illustrated in [Figure 3](#fig-sensitivity).

![](figures/sobol-sensitivity.pdf)

[**Figure 3**]{#fig-sensitivity}. Sensitivity analysis results illustrating the impact of parameter variability on model predictions. Shaded areas represent 95%, 80%, and 60% CI estimated from Monte Carlo samples.

Despite the limited sensitivity of $\lambda$ and $\phi$ to mean RRi, these parameters play a pivotal role in determining the rate of change in RRi in response to physical exertion. Therefore, they can significantly affect RRi dynamics over time. The influence of these rate parameters on RRi fluctuations over time is depicted in [Figure 4](#fig-data-sim).

![](figures/rate-sensitivity.pdf)

[**Figure 4**]{#fig-data-sim}. Simulated RRi dynamics during exercise with varying $\lambda$ and $\phi$ rate parameters expressed as percent change per unit of time. In this simulation, the exercise-induced RRi drop occurs at 5 minutes ($\tau$), and cardiovascular recovery begins 2 minutes after exercise initiation ($\delta$). The model assumes a 90% recovery ($c$ = 0.90) of RRi values following a 400 ms drop ($\beta$) from a baseline of 800 ms ($\alpha$).

## Model behavior to real RRi data

### Sample characteristics

The sample used to assess RRi dynamics consists of a group of 272 subjects selected from a local community of elderly individuals. The sample characteristics can be seen in [Table 2](#tbl-sample-characteristics)

<!--- TABLE 2 HERE --->
[**Table 2**]{#tbl-sample-characteristics}. Sample characteristics from which, continuous RRi monitoring data was collected during a rest-exercise-rest protocol.

```{r}
data("rri_data")
```

Initial exploration of RRi dynamics using two-dimensional density kernel estimation (see [Figure 5](#fig-2d-kernel-density)) indicates a clear drop in RRi around the 5-7 minutes mark, associated with the exercise-induced cardiovascular stress. However, greater variability across individuals in post-exercise recovery can be observed .

![](figures/2d-density-kernel.pdf)

[**Figure 5**]{#fig-2d-kernel-density}. 2D kernel density of RRi dynamics over time from a sample of individuals subjected to a rest-exercise-rest protocol. Darker colors indicate greater probability density. The contrary can be said about lighter colors.

<!---
Interpretación de los parámetros ajustados: Explica los valores estimados de los parámetros 
a, b, c, e, f, t0, y t1, y cómo se interpretan en términos fisiológicos.
--->

<!---
Comparación con otros enfoques: Si es posible, compara el rendimiento del modelo no lineal con otros enfoques más simples, como modelos lineales o mixtos.
--->

# Discussion

<!---
Interpretación de los resultados: Discute cómo los resultados obtenidos apoyan la hipótesis de que el modelo no lineal es más adecuado para describir las fluctuaciones de la HRV en respuesta al ejercicio.
--->

<!---
Implicaciones fisiológicas: Expón qué nos dice el modelo sobre los mecanismos de recuperación cardiovascular y su relevancia para la fisiología deportiva o la medicina.
--->

<!---
Limitaciones: Menciona posibles limitaciones del modelo, como la necesidad de datos de alta calidad o la dificultad de interpretar algunos parámetros en ciertos contextos.
--->

<!---
Futuras direcciones: Sugiere mejoras al modelo, posibles aplicaciones o nuevas líneas de investigación en el uso de enfoques no lineales para modelar respuestas fisiológicas.
--->

# Conclusion

<!---
Resumen de los hallazgos: Reitera brevemente los principales hallazgos y cómo el modelo puede ser útil en futuras investigaciones sobre HRV y la recuperación post-ejercicio.
--->

# Authors' Contributions

Conceptualization, MC-A; Data curation, MC-A; Investigation, MC-A; Methodology, MC-A, NMD; Supervision, CN-E; Formal analysis, MC-A; Visualization, MC-A; Writing–original draft, MC-A, CN-E, [...]; Writing–review & editing, MC-A, CN-E, [...]. All authors have read and agreed to the published version of the manuscript.

# Funding

This research did not receive any funding.

# Data Availability Statement

The data supporting the conclusions of this article will be available from the authors without reservation.

# Conflict of Interest

The authors declare that this research was conducted without any commercial or financial relationships that could be construed as potential conflicts of interest.

# References

<div id="refs"></div>

### 2024-10-29 R/1-prior-rri-model.R

@@ -1,0 +1,133 @@
+
+# Prepare workspace -------------------------------------------------------
+
+## Load libraries ----
+library(ggplot2)
+library(brms)
+library(data.table)
+
+## Custom functions ----
+
+## Main equation
+RRi_model <- function(t, alpha, beta, c, lambda, phi, tau, delta) {
+  alpha +
+    (beta / (1 + exp(lambda * (t - tau)))) +
+    ((-beta * c) / (1 + exp(phi * (t - tau - delta))))
+}
+
+## Plot posterior density of a chosen parameter
+plot_post_dens <- function(posterior, x, xlab) {
+  ggplot(posterior, aes(x = {{x}})) +
+    geom_histogram(aes(y = after_stat(density)), 
+                   col = "darkblue", fill = "lightblue", bins = 20) +
+    stat_density(outline.type = "both", fill = NA, 
+                 col = "gray20", adjust = 1.5,) +
+    scale_y_continuous(expand = c(0,0,0.05,0), breaks = NULL, name = NULL) +
+    scale_x_continuous(expand = c(0,0)) +
+    labs(x = xlab) +
+    theme_classic(base_size = 12) +
+    theme(axis.text = element_text(size = rel(.8)))
+}
+
+## Uniform sampling of a range of parameters for n subjects
+sample_params <- function(n_id, n_rep, min, max) {
+  rep(x = runif(n_id, min, max), each = n_rep)
+}
+
+
+# Define main model -------------------------------------------------------
+
+## Main model
+main_model <- bf(
+  rri ~ alpha + 
+    (beta / (1 + exp(lambda * (time - tau)))) + 
+    ((-beta * c) / (1 + exp(phi * (time - tau - delta)))), 
+  
+  alpha + beta + c ~ 1 | id, 
+  lambda + phi ~ 1 | id, 
+  tau + delta ~ 1 | id, 
+  nl = TRUE
+)
+
+# Define priors for the parameters
+priors <- c(
+  prior(normal(850, 50), nlpar = alpha, lb = 0),
+  prior(normal(-400, 50), nlpar = beta, ub = 0),
+  prior(normal(1, 0.2), nlpar = c, lb = 0),
+  prior(normal(-3, 3), nlpar = lambda, ub = 0),
[…]

### 2024-10-29 R/2-model-constituents.R

@@ -1,0 +1,37 @@
+# Prepare workspace -------------------------------------------------------
+
+## Load libraries ----
+library(ggplot2)
+library(data.table)
+
+# Simulate data -----------------------------------------------------------
+
+sim_data <- data.table(t = seq(0, 20, length.out = 80))
+
+sim_data[, `:=`(alpha = 800,
+                `italic(f)[1](t)` = round(-400 / (1 + exp(log(1 - 0.7)*(t - 5)))),
+                `italic(f)[2](t)` = round((400 * 0.8) / (1 + exp(log(1 - 0.7)*(t - 7)))))]
+
+sim_data[, `RRi == alpha + italic(f)[1](t) + italic(f)[2](t)` := 
+           alpha + `italic(f)[1](t)` + `italic(f)[2](t)`]
+
+sim_data <- melt(sim_data, id.vars = "t", value.name = "RRi", variable.name = "Parameter")
+
+
+# Visualize model constituents --------------------------------------------
+
+figure <- ggplot(sim_data, aes(t, RRi)) +
+  facet_wrap(~ Parameter, scales = "free_y", labeller = label_parsed) +
+  geom_line(aes(col = Parameter), show.legend = FALSE, linewidth = 1) +
+  geom_vline(xintercept = c(5,7), color = "gray50", linewidth = .25) +
+  scale_y_continuous(expand = c(.05, 5), n.breaks = 4) +
+  ggsci::scale_color_observable() +
+  theme_classic(base_size = 12) +
+  theme(strip.background = element_rect(fill = "grey90", colour = NA),
+        axis.text = element_text(size = rel(.8))) +
+  labs(x = "Time (minutes)", y = "R-R intervals (ms)")
+
+ggsave("figures/model-constituents.pdf", figure, "pdf", 
+       width = 7, height = 4, units = "in")
+ggsave("figures/model-constituents.png", figure, "png", 
+       width = 7, height = 4, units = "in", dpi = 500)

### 2024-10-29 R/3-sobol-sensitivity.R

@@ -1,0 +1,140 @@
+
+# Prepare workspace -------------------------------------------------------
+
+## Load libraries ----
+
+library(data.table)
+library(ggplot2)
+library(ggsci)
+
+## Custom functions ----
+
+## Define the RR interval model as a function of time
+rr_interval_model <- function(params, time) {
+  alpha <- params[1]
+  beta <- params[2]
+  c <- params[3]
+  lambda <- params[4]
+  phi <- params[5]
+  tau <- params[6]
+  delta <- params[7]
+  
+  # Apply time-dependent RR interval model with logistic functions
+  rr_interval <- alpha + beta / (1 + exp(lambda * (time - tau))) +
+    (-beta * c) / (1 + exp(phi * (time - delta)))
+  
+  return(rr_interval)
+}
+
+# Generate random uniform samples for each parameter
+sobol_sample <- function(n_samples) {
+  out <- data.frame(
+    alpha = runif(n_samples, min = 400, max = 1200), 
+    beta = runif(n_samples, min = -600, max = -200), 
+    c = runif(n_samples, min = 0.5, max = 1.5), 
+    lambda = runif(n_samples, min = -3.454, max = -1.151), 
+    phi = runif(n_samples, min = -1.806, max = -0.602),
+    tau = runif(n_samples, min = 2.5, max = 7.5), 
+    delta = runif(n_samples, min = 1, max = 3)
+  )
+  
+  return(out)
+}
+
+# Perform Sobol analysis
+sobol_variance <- function(model_function, samples, time_points) {
+  # Generate total output for each sample to get baseline variance
+  total_output <- sapply(1:nrow(samples), function(i) {
+    model_function(as.numeric(samples[i, ]), time_points)
+  })
+  
+  total_variance <- apply(total_output, 1, var) # Variance across samples for each time point
+  
+  # Initialize matrix for Sobol indices results
+  sobol_indices <- matrix(0, nrow = length(time_points), ncol = ncol(samples),
+                          dimnames = list(NULL, names(samples)))
+  
+  # Loop over each parameter to calculate first-order Sobol indices
+  for (param in names(samples)) {
+    
+    perturbed_output <- sapply(1:nrow(samples), function(i) {
+      # Perturb only the chosen parameter, while averaging the others
+      modified_params <- as.numeric(samples[i, ])
+      modified_params[names(samples) != param] <- apply(samples[names(samples) != param], 2, mean)
[…]

### 2024-10-29 R/4-rate-sensitivity.R

@@ -1,0 +1,108 @@
+# Prepare workspace -------------------------------------------------------
+
+## Load libraries ----
+library(ggplot2)
+library(brms)
+library(data.table)
+
+## Custom functions ----
+
+## Main function to simulate data
+RRi_sim <- function(t, alpha, beta, c, lambda, phi, tau, delta) {
+  hrv <-
+    alpha +
+    (beta / (1 + exp(lambda * (t - tau)))) +
+    ((-beta * c) / (1 + exp(phi * (t - tau - delta))))
+  
+  out <- data.table::data.table(time = t, hrv)
+  return(out)
+}
+
+## Function to simulate combinations of different parameters
+sim_parameters <- function(t,
+                           alpha, 
+                           beta,
+                           c,
+                           lambda,
+                           phi,
+                           tau,
+                           delta) {
+  
+  params <- expand.grid(
+    alpha = alpha, beta = beta, c = c, 
+    lambda = lambda, phi = phi, 
+    tau = tau, delta = delta)
+  len_params <- nrow(params)
+  
+  sim_data <- vector("list", length = len_params)
+  
+  for (i in seq_len(len_params)) {
+    sim_data[[i]] <- RRi_sim(
+      t = t,
+      alpha = params$alpha[i],
+      beta = params$beta[i],
+      c = params$c[i],
+      lambda = params$lambda[i],
+      phi = params$phi[i],
+      tau = params$tau[i],
+      delta = params$delta[i]
+    )
+    
+    sim_data[[i]]$alpha <- params$alpha[i]
+    sim_data[[i]]$beta <- params$beta[i]
+    sim_data[[i]]$c <- params$c[i]
+    sim_data[[i]]$lambda <- params$lambda[i]
+    sim_data[[i]]$phi <- params$phi[i]
+    sim_data[[i]]$tau <- params$tau[i]
+    sim_data[[i]]$delta <- params$delta[i]
+    # sim_data[[i]]$hrv <- sim_data[[i]]$hrv + rstudent_t(n_time, 5, 0, 50)
+  }
+  
+  sim_data <- data.table::rbindlist(sim_data, idcol = "id")
+  sim_data[, id := factor(id)][]
+  return(sim_data)
+}
+
+
+# Simulate data -----------------------------------------------------------
+
+sim_data <- sim_parameters(
+  t = seq(0, 20, length.out = 120),
[…]

### 2024-10-29 R/5-2d-density-kernel.R

@@ -1,0 +1,33 @@
+
+# Prepare workspace -------------------------------------------------------
+
+## Load libraries
+library(ggplot2)
+library(data.table)
+
+## Load data
+data("rri_data")
+
+
+# Generate figure ---------------------------------------------------------
+
+rri_data[, time_cat := cut(time, breaks = seq(0, 20, by = 1), 
+               labels = seq(0, 20, by = 1)[-1]) |> 
+           as.character() |> as.numeric()]
+
+estimate <- rri_data[, list(mean = mean(rr_denoised, na.rm = T),
+                            se = sd(rr_denoised, na.rm = T)), 
+                     list(time = as.numeric(as.character(time_cat)))]
+
+figure <- ggplot(rri_data, aes(time, rr_denoised)) +
+  stat_density_2d_filled(show.legend = FALSE, adjust = 1/3, na.rm = TRUE) +
+  scale_x_continuous(expand = c(0,0), name = "Time (minutes)") +
+  scale_y_continuous(expand = c(0,0), name = "RRi (ms)") +
+  scale_fill_brewer() +
+  theme_classic(base_size = 12) +
+  theme(axis.text = element_text(size = rel(0.8)))
+
+ggsave("figures/2d-density-kernel.pdf", figure, "pdf", 
+       width = 7, height = 4, units = "in")
+ggsave("figures/2d-density-kernel.png", figure, "png", 
+       width = 7, height = 4, units = "in", dpi = 500)

### 2024-10-29 data-raw/process-data.R

@@ -1,0 +1,93 @@
+
+# Prepare workspace -------------------------------------------------------
+
+## Load packages
+library(data.table)
+library(brms)
+library(ggplot2)
+
+## Selection of txt files
+files <- list.files(path = "data-raw/hrv_data", full.names = TRUE, pattern = "\\.txt$")
+
+## Data import
+rri_data <- lapply(files, \(i) {
+  rr_interval <- as.double(x = readLines(i))
+  not_na <- !is.na(rr_interval)
+  rr_interval <- rr_interval[not_na & between(rr_interval, 400, 1200)]
+  time <- cumsum(rr_interval)/60000
+  data.table(rr_interval = rr_interval[time <= 20],
+             time = time[time <= 20],
+             file = gsub("data-raw/", "", i, fixed = TRUE))
+}) |> 
+  rbindlist(idcol = "id")
+
+## Data filtering
+id_to_keep <- rri_data[, max(time) > 10, id][V1 == TRUE, id]
+id_to_move <- rri_data[, max(time) > 10, id][V1 == FALSE, id]
+files_to_move <- rri_data[id %in% id_to_keep, files[!files %in% paste0("data-raw/", file)]]
+
+
+if (length(id_to_move) > 0) {
+  for (i in id_to_move) {
+    file <- rri_data[id == i, unique(file)]
+    from <- paste0("data-raw/", file)
+    to <- paste0("data-raw/hrv_data/unused/", gsub("hrv_data/", "", file))
+    file.copy(from = from, to = to, overwrite = TRUE)
+    file.remove(from)
+  }
+}
+
+if (length(files_to_move) > 0) {
+  for (i in files_to_move) {
+    to <- paste0("data-raw/hrv_data/unused/", gsub("data-raw/hrv_data/", "", i))
+    file.copy(from = i, to = to, overwrite = TRUE)
+    file.remove(i)
+  }
+}
+
+rm(files, files_to_move, id_to_move)
+
+rri_data <- rri_data[id %in% id_to_keep]
+
+
+# Processing signal data --------------------------------------------------
+
+## Denoising the signal data
+denoise <- function(x, w = 0.1, n = 3) {
+  signal::filtfilt(filt = signal::butter(n, w), x)
+}
+
+## Applying Butterworth 10 Hz low-pass filter
+rri_data[, j = rr_denoised := denoise(rr_interval), id]
+
+trim <- function(x, abs) {
+  x_len <- length(x)
[…]

### 2024-10-30 paper.qmd
update on paper methods and prior model code
- more samples and chains used, given the computational power available

@@
 
-Despite the importance of RRi analysis, modeling their nonlinear behavior during and after exercise presents significant challenges.
+Despite the importance of RRi analysis, modeling their nonlinear behavior during and after exercise in one continuous measurement it is not without significant challenges.
 Traditional approaches, such as linear regression and time-series analysis, often fail to capture the intricate transitions in RRi, especially during intense physical exertion and recovery.
@@
 Hence, the primary objective of this paper is to present a novel non-linear model that characterizes the fluctuations in RRi during exercise and recovery.
-This model is specifically designed to reflect the complex, real-time changes in RRi, offering a more precise understanding of the physiological processes underlying cardiovascular adaptation to exercise.
+This model is specifically designed to reflect the complex, real-time changes in RRi, offering meaningful model parameters with the potential to explore a precise understanding of the physiological processes underlying cardiovascular adaptation to exercise.
 By focusing exclusively on RRi, the model can deliver insights directly applicable to athletic training regimens, recovery protocols, and clinical practices aimed at improving cardiovascular health.
@@
 
-The priors were selected based on physiological constraints and are detailed in @eq-priors.
+The priors were selected based on physiological constraints and are detailed in @eq-priors, ensuring in this way, identifiability of model parameters by constraining the parameter space to physiologically plausible values and ultimately, improving model convergence.
 
@@
 
-![](figures/prior-model.pdf)
+![](figures/prior-model.png)
 
@@
 
-The fitting process utilized four Markov Chain Monte Carlo (MCMC) chains, each consisting of 12,500 iterations with a burn-in period of 5,000 iterations, resulting in a total of 30,000 post-warmup samples.
+The fitting process utilized five Markov Chain Monte Carlo (MCMC) chains, each consisting of 15,000 iterations with a burn-in period of 5,000 iterations, resulting in a total of 50,000 post-warmup samples.
 
@@
 
-![](figures/model-constituents.pdf)
+![](figures/model-constituents.png)
 
@@
 
-![](figures/sobol-sensitivity.pdf)
+![](figures/sobol-sensitivity.png)
 
 [**Figure 3**]{#fig-sensitivity}.
-Sensitivity analysis results illustrating the impact of parameter variability on model predictions.
+Sensitivity analysis results illustrating the impact of parameter variability on model predictions and the percentage of explained variance accounted for each model parameter.
 Shaded areas represent 95%, 80%, and 60% CI estimated from Monte Carlo samples.
@@
 
-![](figures/rate-sensitivity.pdf)
+![](figures/rate-sensitivity.png)
 
@@
 Initial exploration of RRi dynamics using two-dimensional density kernel estimation (see [Figure 5](#fig-2d-kernel-density)) indicates a clear drop in RRi around the 5-7 minutes mark, associated with the exercise-induced cardiovascular stress.
-However, greater variability across individuals in post-exercise recovery can be observed .
+However, greater variability across individuals in post-exercise recovery can be observed.
 
-![](figures/2d-density-kernel.pdf)
+![](figures/2d-density-kernel.png)
 

### 2024-10-30 R/1-prior-rri-model.R

@@ -5,6 +5,7 @@
 library(ggplot2)
 library(brms)
 library(data.table)
+library(ggpubr)
 
 ## Custom functions ----
 
@@ -86,8 +87,8 @@ m_prior_only <- brm(
   family = gaussian(),
   prior = priors,
   seed = 1234,
-  iter = 12500, warmup = 5000,
-  chains = 4, cores = 4,
+  iter = 15000, warmup = 5000,
+  chains = 5, cores = 5,
   sample_prior = "only",
   file = "models/m_prior_only.RDS"
 )

### 2024-10-30 R/2-model-constituents.R

@@ -3,6 +3,7 @@
 ## Load libraries ----
 library(ggplot2)
 library(data.table)
+library(ggsci)
 
 # Simulate data -----------------------------------------------------------
 

### 2024-10-30 R/4-rate-sensitivity.R

@@ -4,6 +4,7 @@
 library(ggplot2)
 library(brms)
 library(data.table)
+library(ggsci)
 
 ## Custom functions ----
 

### 2024-11-01 paper.qmd
update on manuscript, figures and sim data testing

@@
     reference-doc: "ref.docx"
-    fig-format: pdf
-    fig-asp: 0.75
-    fig-width: 7
 bibliography: "misc/ref.bib"
@@
 
-Current research has delved deep into understanding the underlying mechanisms behind cardiac autonomic dynamics in response to exercise, and how this could be related to health-related quality of life and an increased risk of cardiovascular diseases.
-Hence, shedding light on the autonomic mechanics can provide insights into the role play between modifying parameters of exercise-induced response and how to optimize it to achieve optimal effects in both young and older individuals.
+Current research has extensively examined the mechanisms underlying cardiac autonomic dynamics in response to exercise and their links to health-related quality of life and cardiovascular disease risk.
+Understanding these autonomic processes offers valuable insights into optimizing exercise-induced adaptations, with implications for both younger and older individuals.
 
-The study of R-R intervals (RRi) in response to exercise is a key research area within this context, given its significant implications for cardiovascular health, athletic performance, and overall physiological adaptation [@kristal1995heart; @thayer2010relationship; @dong2016role; @lundstrom2023practices].
-Unlike measures of heart rate variability (HRV), which summarize autonomic activity over time, RRi offers a more direct and granular view of the heart's electrical activity during or immediately after an exercise test in older individuals [@mongin2022hrv; @castillo2023validity; @mabe2024associations].
-Analyzing the temporal dynamics of RRi (i.e., the time between successive heartbeats) provides invaluable insights into how the cardiovascular system behaves to and recovers from physical stressors, such as fatigue after competition or an intense exercise test [@castillo2021cardiac; @eser2022exercise].
+In this context, the study of R-R intervals (RRi) in response to exercise has emerged as an important research area, given its relevance to cardiovascular health, athletic performance, and physiological adaptation [@kristal1995heart; @thayer2010relationship; @dong2016role; @lundstrom2023practices].
+Unlike heart rate variability (HRV), which aggregates autonomic responses over time, RRi analysis provides a more granular, direct view of cardiac electrical activity during or immediately following exercise, particularly in older adults [@mongin2022hrv; @castillo2023validity; @mabe2024associations].
+Analyzing the temporal dynamics of RRi (i.e., the time between successive heartbeats) provides invaluable insights into how the cardiovascular system behaves to and recovers from physical stressors such as exercise-induced fatigue and competition-related strain [@castillo2021cardiac; @eser2022exercise].
 
-Understanding these fluctuations is particularly relevant during dynamic periods of exercise when the autonomic nervous system (ANS) shifts between parasympathetic withdrawal and sympathetic activation [@boettger2010heart].
-Modeling RRi dynamics, rather than relying on broader HRV metrics, allows researchers to focus on the direct physiological markers of autonomic adaptation to stress [@hautala2003short].
-This detailed perspective is critical for identifying recovery patterns and understanding cardiovascular reactivity in athletes and individuals with varied fitness levels [@mongin2022hrv].
+Understanding these fluctuations is particularly relevant during dynamic exercise periods, where the autonomic nervous system (ANS) shifts between parasympathetic withdrawal and sympathetic activation [@boettger2010heart].
+Modeling RRi dynamics, rather than relying on broader HRV metrics, allows for a direct assessment of physiological markers of autonomic adaptation to stress [@hautala2003short].
+This approach is valuable for identifying recovery patterns and understanding cardiovascular reactivity across individuals with various fitness levels [@mongin2022hrv].
 
-Despite the importance of RRi analysis, modeling their nonlinear behavior during and after exercise in one continuous measurement it is not without significant challenges.
-Traditional approaches, such as linear regression and time-series analysis, often fail to capture the intricate transitions in RRi, especially during intense physical exertion and recovery.
-This limitation is due to the non-linear and non-stationary nature of HRV [@gronwald2019effects].
-Linear models, which oversimplify these transitions, are prone to misrepresent the actual physiological processes.
-More advanced non-linear models have begun to address this issue, yet many focus on overall HRV parameters rather than directly modeling the RRi [@gronwald2019non].
+Despite its importance, modeling RRi behavior during and after exercise in a continuous measurement poses significant challenges.
+Traditional approaches, such as linear regression and time-series analysis, often fail to capture the intricate transitions in RRi, especially under intense exertion and recovery phases.
+This difficulty arises due to the inherently non-linear and non-stationary nature of HRV [@gronwald2019effects].
+While linear models oversimplify these dynamics, advanced non-linear approaches have been developed to address the limitations of linear analysis.
+However, many focus on HRV summaries rather than direct RRi modeling [@gronwald2019non].
 
-Non-linear models have gained increasing attention for analyzing the complex dynamics of RRi, especially during exercise.
-These models offer a more comprehensive understanding of HRV compared to traditional linear approaches [@wu2003nonlinear].
-Recent studies have proposed exponential decay models to describe RRi recovery, while others have employed logistic functions to depict the gradual return to baseline following periods of heightened physical stress [@molkkari2020online; @gronwald2019non].
-These models offer an advantage over traditional HRV analyses by providing a more detailed understanding of the cardiovascular system's response to exercise.
-However, despite these advancements, few models explicitly target real-time RRi dynamics, and fewer still incorporate the variability introduced by individual physiological differences, such as fitness level, autonomic balance, and exercise intensity [@kanniainen2023estimation].
+Recent studies have begun exploring non-linear models for RRi dynamics, recognizing their potential to capture the complexity of cardiovascular response to exercise.
+Exponential decay models, for example, have been proposed to describe RRi recovery, while logistic functions have been used to model the gradual return to baseline after high-intensity exercise [@molkkari2020online; @gronwald2019non].
+These models offer advantages over traditional HRV metrics by providing a more detailed understanding of the cardiovascular system's response to exercise [@wu2003nonlinear].
+However, despite these advancements, few models are specifically designed to capture real-time RRi fluctuations, and even fewer consider individual variability factors such as fitness level, autonomic balance, and exercise intensity [@kanniainen2023estimation].
 
-Given the unique characteristics of RRi, their direct relationship with cardiac electrical activity, and responsiveness to autonomic changes, there is a compelling need for a model that captures their non-linear fluctuations during exercise and recovery.
-Such a model would offer a more biologically relevant representation of the heart’s behavior than the broader HRV indices typically used in research [@bacopoulou2021entropy].
+Given the unique characteristics of RRi, their direct relationship with cardiac electrical activity, and responsiveness to autonomic changes, there is a compelling need for a model that accurately represents RRi's non-linear fluctuations during exercise and recovery.
+Such a model would offer a more physiologically relevant representation of the heart's behavior compared to the broader HRV indices commonly used in research [@bacopoulou2021entropy].
 
-Hence, the primary objective of this paper is to present a novel non-linear model that characterizes the fluctuations in RRi during exercise and recovery.
-This model is specifically designed to reflect the complex, real-time changes in RRi, offering meaningful model parameters with the potential to explore a precise understanding of the physiological processes underlying cardiovascular adaptation to exercise.
+Hence, the primary objective of this paper is to present a novel non-linear model that characterizes RRi fluctuations during exercise and recovery.
+This model is designed to capture the complex, real-time changes in RRi, providing meaningful parameters that can enhance our understanding of the physiological processes underlying cardiovascular adaptation to exercise.
 By focusing exclusively on RRi, the model can deliver insights directly applicable to athletic training regimens, recovery protocols, and clinical practices aimed at improving cardiovascular health.
@@
 To assess the sensitivity of model parameters in influencing RRi over time, we implemented a Sobol sensitivity analysis using Monte Carlo simulations.
-To compute Sobol indices, a total of 500 Monte Carlo simulations were conducted, with each simulation involving 500 randomly sampled parameter sets (250,000 model runs in total).
-For each set of parameters, RR intervals were calculated at each time point $t$ across a range from 0 to 20 minutes at intervals of 0.1 minutes.
-The resulting Sobol indices provided a measure of the contribution of each parameter to the variance in RR intervals at each time point.
+Sobol analysis was selected for its robustness in handling non-linear and non-monotonic relationships, which are intrinsic to RRi dynamics in response to exercise.
+To compute Sobol indices, a total of 1000 Monte Carlo simulations were conducted, with each simulation involving 1000 randomly sampled parameter sets (1,000,000 model runs in total).
+For each set of parameters, RRi were calculated at each time point $t$ across a range from 0 to 20 minutes at intervals of 0.1 minutes (i.e., 6 seconds).
+The resulting Sobol indices provided a measure of the contribution of each parameter to the variance in RRi at each time point.
 
@@
 To achieve this, we perturbed each parameter individually while holding all other parameters at their average values across the samples.
-The proportion of variance explained by each parameter at each time point was calculated by dividing the variance of the perturbed model outputs by the total variance, yielding time-dependent Sobol indices.
+The proportion of variance explained by each parameter at each time point was calculated by dividing the variance of the estimated RRi of the perturbed parameter by the total variance, which accounts for the variation in all model parameters simultaneously, yielding time-dependent Sobol indices.
 The selected parameter ranges, provided in [Table 1](#tbl-sens-params), reflect a 50% variation in both directions from a reference value for each parameter.
@@
 Shaded areas represent 95%, 80%, and 60% CI estimated from Monte Carlo samples.
+Time in x-axis is truncated at 12 minutes, given that around this time, the proportion of the total variance explained by the respective model parameter stabilizes respective of time.
 
@@
 
-<!---
-Interpretación de los parámetros ajustados: Explica los valores estimados de los parámetros 
-a, b, c, e, f, t0, y t1, y cómo se interpretan en términos fisiológicos.
---->
-
-<!---
-Comparación con otros enfoques: Si es posible, compara el rendimiento del modelo no lineal con otros enfoques más simples, como modelos lineales o mixtos.
---->
-
 # Discussion
 
-<!---
-Interpretación de los resultados: Discute cómo los resultados obtenidos apoyan la hipótesis de que el modelo no lineal es más adecuado para describir las fluctuaciones de la HRV en respuesta al ejercicio.
---->
+In this study, we present a non-linear model designed to capture R-R interval (RRi) fluctuations during exercise and recovery, providing an in-depth view of cardiac autonomic dynamics.
+By focusing on RRi rather than aggregate HRV metrics, our approach enables real-time tracking of cardiovascular responses to physical stress, which has implications for both clinical and athletic applications.
+This model marks a step forward in computational physiology, bridging the gap between theoretical frameworks and practical monitoring applications.
 
-<!---
-Implicaciones fisiológicas: Expón qué nos dice el modelo sobre los mecanismos de recuperación cardiovascular y su relevancia para la fisiología deportiva o la medicina.
---->
+<!--- Sensitivity Analysis and Model Robustness --->
+The Sobol sensitivity analysis revealed baseline RRi ($\alpha$) and recovery proportion ($c$) as key drivers of model output variance, consistent with physiological expectations.
+Baseline RRi reflects intrinsic cardiac properties, while recovery proportion captures autonomic re-engagement post-exercise, both central to cardiovascular adaptation.
+The relatively low sensitivity of decay and recovery rate parameters ($\lambda$ and $\phi$) suggests the model is robust to moderate fluctuations in these rates, which enhances its reliability for both individualized and group-based applications.
 
-<!---
-Limitaciones: Menciona posibles limitaciones del modelo, como la necesidad de datos de alta calidad o la dificultad de interpretar algunos parámetros en ciertos contextos.
---->
+Sobol analysis was selected for its capacity to handle non-linear, time-varying relationships, but it assumes parameter independence and demands substantial computational resources.
+This assumption may overlook interdependencies typical in biological systems, where parameters such as decay rates and recovery kinetics often interact.
+For future work, alternative sensitivity methods, such as Bayesian sensitivity analysis, could address interdependencies and potentially improve interpretive accuracy in complex biological contexts.
+Additionally, real-time applications in resource-limited environments may benefit from computational optimizations, such as surrogate modeling or sparse-grid approximations, to reduce processing time while maintaining analytical depth.
 
-<!---
-Futuras direcciones: Sugiere mejoras al modelo, posibles aplicaciones o nuevas líneas de investigación en el uso de enfoques no lineales para modelar respuestas fisiológicas.
---->
+<!--- Validation and Practical Applications --->
+The practical application of this model extends to clinical monitoring and athletic training, but further validation is essential to confirm its robustness in real-world conditions.
+To support clinical decision-making, future studies should validate the model against established physiological benchmarks, such as VO₂ max and lactate thresholds, which would facilitate its integration into routine health assessments.
+For instance, this model could be valuable in cardiovascular rehabilitation programs, where tracking real-time autonomic responses can help personalize exercise regimens, ensuring adequate recovery without overstressing the cardiovascular system.
+Furthermore, for at-risk populations, the model could assist in detecting autonomic irregularities indicative of early cardiovascular dysfunction.
+
+For athletic applications, the model has potential in guiding interval training, where it could help identify optimal recovery points between intense exercise bouts.
+In this setting, real-time tracking of RRi dynamics may help prevent overtraining and optimize performance by aligning training load with individual recovery capacity.
+However, extending this model beyond controlled, single-bout settings requires further validation, particularly in dynamic environments where external factors like temperature, altitude, and psychological stress can influence RRi.
+Integrating the model with wearable devices would be essential for field applications, but these devices must capture RRi with sufficient accuracy to maintain model precision.
+Future studies should focus on adapting the model for wearable technology and testing it in diverse settings to evaluate its reliability across environmental and situational variations.
+
+<!--- Advances Over Existing Models --->
+Traditional models of RRi often rely on linear assumptions or simple exponential functions, which fail to capture the transient, non-stationary nature of RRi during exercise and recovery.
+By using logistic functions to model both the decay and recovery phases, our non-linear approach accommodates gradual autonomic shifts more effectively than exponential functions, offering a physiologically relevant representation of RRi dynamics.
+This model also enhances interpretability by isolating time-dependent contributions of each parameter, as demonstrated in the Sobol analysis, thereby enabling applications across varied fitness levels and exercise intensities.
+
+While recent studies have developed non-linear models, many focus on HRV aggregates that summarize rather than directly model RRi changes.
+By focusing on RRi, our model provides a high-resolution view of cardiac adaptation, which is particularly valuable for both clinical and athletic settings.
+Furthermore, the ability to adapt the model parameters to real-time changes in RRi has the potential to enhance sensitivity in detecting subtle shifts in autonomic state, which aggregated HRV indices may obscure.
+
+<!--- Limitations and Future Directions --->
+This model has limitations that warrant careful consideration.
+First, the uniform parameter sampling used in sensitivity analysis, though practical, may not fully capture individual variability in populations with extreme autonomic profiles or chronic health conditions.
+Moving forward, incorporating empirical distributions or Bayesian priors based on population data could refine parameter estimates, enhancing individualization for diverse clinical populations, such as older adults or individuals with cardiovascular disease.
+
+Another area for future research is multi-session modeling to account for cumulative adaptations over repeated exercise sessions.
+Real-world training and rehabilitation often involve multiple bouts of exercise that can introduce cumulative autonomic fatigue or adaptation.
+Extending the model to track such longitudinal effects could help identify early signs of overtraining or autonomic exhaustion, thereby providing a preventive tool for both clinical and athletic populations.
+Additionally, considering environmental and psychological factors—such as temperature, stress, or sleep quality—would add robustness, ensuring that model predictions remain accurate across varied real-world contexts.
 
@@
 
-<!---
-Resumen de los hallazgos: Reitera brevemente los principales hallazgos y cómo el modelo puede ser útil en futuras investigaciones sobre HRV y la recuperación post-ejercicio.
---->
+In summary, this study introduces a non-linear model of RRi dynamics that captures complex, real-time fluctuations reflecting the autonomic nervous system’s response to exercise and recovery.
+By focusing on RRi, the model offers enhanced sensitivity to physiological changes, making it well-suited for personalized applications in clinical and athletic settings.
+The use of Sobol sensitivity analysis enabled a time-dependent interpretation of parameter effects, underscoring the model’s adaptability to diverse physiological states and exercise intensities.
+Nevertheless, further validation, particularly with multi-session and wearable-based applications, is essential to establish the model’s robustness and applicability in dynamic, real-world environments.
+By advancing precision in cardiovascular monitoring, this model holds promise for supporting individualized health and performance interventions, contributing meaningfully to the fields of computational physiology and applied health sciences.
 

### 2024-11-01 R/0-test-sim-data.R

@@ -1,0 +1,92 @@
+
+# Prepare workspace -------------------------------------------------------
+
+## Load libraries
+library(data.table)
+library(ggplot2)
+library(brms)
+
+## Main equation
+RRi_model <- function(t, alpha, beta, c, lambda, phi, tau, delta) {
+  alpha +
+    (beta / (1 + exp(lambda * (t - tau)))) +
+    ((-beta * c) / (1 + exp(phi * (t - tau - delta))))
+}
+
+## Simulating some data
+n_id <- 10
+n_time <- 100
+
+set.seed(1234)
+sim_data <- within(
+  data = data.table(), 
+  expr = {
+    alpha = rep(runif(n_id, 800, 1000), each = n_time)
+    beta = rep(runif(n_id, -500, -300), each = n_time)
+    c = rep(runif(n_id, .7, 1.1), each = n_time)
+    lambda = rep(runif(n_id, log(1-.9), log(1-.7)), each = n_time)
+    phi = rep(runif(n_id, log(1-.7), log(1-.5)), each = n_time)
+    tau = rep(runif(n_id, 3, 7), each = n_time)
+    delta = rep(runif(n_id, 1, 3), each = n_time)
+    time = rep(x = seq(0, 20, length.out = n_time), times = n_id)
+    rri_true = RRi_model(time, alpha, beta, c, lambda, phi, tau, delta)
+    rri_hat = rri_true + brms::rstudent_t(n_id * n_time, df = 5, sigma = 20)
+    id = rep(1:n_id, each = n_time)
+  }
+)
+
+## Plot the simulated data
+ggplot(sim_data, aes(time, rri_true, group = as.factor(id))) +
+  facet_grid(~ id) +
+  geom_line() +
+  geom_line(aes(y = rri_hat)) +
+  theme_linedraw()
+
+
+# Simple model, no priors -------------------------------------------------
+
+m_prior <- 
+  c(
+    ## Fix effects
+    prior(normal(0, 1), nlpar = alpha), 
+    prior(normal(-1, 1), nlpar = beta, ub = 0),
+    prior(normal(1, 0.2), nlpar = c, lb = 0),
+    prior(normal(-2, 0.5), nlpar = lambda, ub = 0),
+    prior(normal(-2, 0.5), nlpar = phi, ub = 0),
+    prior(normal(5, 0.5), nlpar = tau, lb = 0),
[…]

### 2024-11-01 R/1-prior-rri-model.R

@@ -129,6 +129,6 @@ figure <- ggpubr::ggarrange(
 )
 
 ggsave("figures/prior-model.pdf", figure, "pdf", 
-       width = 7, height = 4, units = "in")
+       width = 7, height = 5, units = "in")
 ggsave("figures/prior-model.png", figure, "png", 
-       width = 7, height = 4, units = "in", dpi = 500)
+       width = 7, height = 5, units = "in", dpi = 500)

### 2024-11-01 R/2-model-constituents.R

@@ -33,6 +33,6 @@ figure <- ggplot(sim_data, aes(t, RRi)) +
   labs(x = "Time (minutes)", y = "R-R intervals (ms)")
 
 ggsave("figures/model-constituents.pdf", figure, "pdf", 
-       width = 7, height = 4, units = "in")
+       width = 7, height = 5, units = "in")
 ggsave("figures/model-constituents.png", figure, "png", 
-       width = 7, height = 4, units = "in", dpi = 500)
+       width = 7, height = 5, units = "in", dpi = 500)

### 2024-11-01 R/3-sobol-sensitivity.R

@@ -41,7 +41,7 @@ sobol_sample <- function(n_samples) {
   return(out)
 }
 
-# Perform Sobol analysis
+# Sobol analysis main function
 sobol_variance <- function(model_function, samples, time_points) {
   # Generate total output for each sample to get baseline variance
   total_output <- sapply(1:nrow(samples), function(i) {
@@ -76,8 +76,8 @@ sobol_variance <- function(model_function, samples, time_points) {
 
 # Perform Bootstrapped Sobol analysis -------------------------------------
 
-n_boot <- 500                       ## Number of Monte Carlo simulations
-n_samples <- 500                    ## Number of samples per MC sims
+n_boot <- 1000                       ## Number of Monte Carlo simulations
+n_samples <- 1000                    ## Number of samples per MC sims
 time_points <- seq(0, 20, by = 0.1) ## Time points for RR interval calculations
 
 boot <- vector(mode = "list", length = n_boot) ## Pre allocate list for results
@@ -101,40 +101,58 @@ if (!file.exists("R/sobol-sensitivity.RDS")) { ## Omit computation if already ex
   
   ## Save results to prevent future computation time
   saveRDS(object = sobol_results, file = "R/sobol-sensitivity.RDS")
+} else {
+  sobol_results <- readRDS("R/sobol-sensitivity.RDS")
 }
 
 
-figure <- ggplot(sobol_results[time <= 10], aes(time, mean)) +
-  facet_wrap(~ variable, nrow = 1,
-             labeller = label_parsed,
-             scales = "free_y") +
-  geom_line(aes(col = variable), show.legend = FALSE) +
-  geom_ribbon(aes(fill = variable, 
-                  ymin = mean - se * qnorm(.975),
-                  ymax = mean + se * qnorm(.975)), 
-              show.legend = FALSE, alpha = .3) +
-  geom_ribbon(aes(fill = variable, 
-                  ymin = mean - se * qnorm(.9),
-                  ymax = mean + se * qnorm(.9)), 
-              show.legend = FALSE, alpha = .3) +
-  geom_ribbon(aes(fill = variable, 
-                  ymin = mean - se * qnorm(.8),
-                  ymax = mean + se * qnorm(.8)), 
-              show.legend = FALSE, alpha = .3) +
-  scale_y_continuous(labels = scales::label_percent(accuracy = 1),
-                     name = "Sensitivity",
[…]

### 2024-11-01 R/4-rate-sensitivity.R

@@ -91,7 +91,7 @@ sim_data[, phi := factor(phi,
 
 figure <- ggplot(sim_data, aes(time, hrv)) +
   facet_wrap(~ phi, labeller = labeller(phi = label_parsed)) +
-  geom_line(aes(col = as.factor(x = lambda), group = id)) +
+  geom_line(aes(col = as.factor(x = lambda), group = id), linewidth = 1) +
   geom_vline(xintercept = c(5,7), color = "gray50", linewidth = .25) +
   scale_y_continuous(breaks = c(500, 600, 700, 800)) +
   scale_alpha_manual(values = c(1, .85, .7, .55)) +

### 2024-11-01 R/0-test-sim-data.R
model testing

@@ -13,80 +13,135 @@ RRi_model <- function(t, alpha, beta, c, lambda, phi, tau, delta) {
     ((-beta * c) / (1 + exp(phi * (t - tau - delta))))
 }
 
-## Simulating some data
-n_id <- 10
-n_time <- 100
+simulate_data <- function(n_id, n_time, noise = 10, seed = 1234) {
+  set.seed(seed)
+  within(
+    data = data.table::data.table(), 
+    expr = {
+      alpha = rep(runif(n_id, 800, 1000), each = n_time)
+      beta = rep(runif(n_id, -500, -300), each = n_time)
+      c = rep(runif(n_id, .7, 1.1), each = n_time)
+      lambda = rep(runif(n_id, log(1-.9), log(1-.7)), each = n_time)
+      phi = rep(runif(n_id, log(1-.7), log(1-.5)), each = n_time)
+      tau = rep(runif(n_id, 3, 7), each = n_time)
+      delta = rep(runif(n_id, 1, 3), each = n_time)
+      time = rep(x = seq(0, 20, length.out = n_time), times = n_id)
+      rri_true = RRi_model(time, alpha, beta, c, lambda, phi, tau, delta)
+      rri_hat = rri_true + brms::rstudent_t(n_id * n_time, df = 5, sigma = noise)
+      id = rep(1:n_id, each = n_time)
+    }
+  )
+}
 
-set.seed(1234)
-sim_data <- within(
-  data = data.table(), 
-  expr = {
-    alpha = rep(runif(n_id, 800, 1000), each = n_time)
-    beta = rep(runif(n_id, -500, -300), each = n_time)
-    c = rep(runif(n_id, .7, 1.1), each = n_time)
-    lambda = rep(runif(n_id, log(1-.9), log(1-.7)), each = n_time)
-    phi = rep(runif(n_id, log(1-.7), log(1-.5)), each = n_time)
-    tau = rep(runif(n_id, 3, 7), each = n_time)
-    delta = rep(runif(n_id, 1, 3), each = n_time)
-    time = rep(x = seq(0, 20, length.out = n_time), times = n_id)
-    rri_true = RRi_model(time, alpha, beta, c, lambda, phi, tau, delta)
-    rri_hat = rri_true + brms::rstudent_t(n_id * n_time, df = 5, sigma = 20)
-    id = rep(1:n_id, each = n_time)
-  }
+## Load the data
+data("rri_data")
+
[…]

### 2024-11-01 R/6-fit-model-to-data.R

@@ -1,0 +1,116 @@
+# Prepare workspace -------------------------------------------------------
+
+## Load libraries ----
+library(ggplot2)
+library(brms)
+library(data.table)
+library(ggpubr)
+
+## Load datasets ----
+data("rri_data")
+
+## Standardize as with the simulated dataset
+rri_data[, rri_mean := mean(rr_denoised, na.rm = TRUE), by = id]
+rri_data[, rri_sd := sd(rr_denoised, na.rm = TRUE), by = id]
+rri_data[, rri_std := (rr_denoised - rri_mean) / rri_sd, by = id]
+
+model_data <- rri_data[!is.na(rri_std), .SD[sort(sample.int(.N, 200))], id]
+
+## Custom functions ----
+
+## Main equation
+RRi_model <- function(t, alpha, beta, c, lambda, phi, tau, delta) {
+  alpha +
+    (beta / (1 + exp(lambda * (t - tau)))) +
+    ((-beta * c) / (1 + exp(phi * (t - tau - delta))))
+}
+
+
+# Build models ------------------------------------------------------------
+
+## Model without Random intercepts ----
+
+### Model formula
+m1_formula <- bf(
+  rri_std ~ alpha + 
+    (beta / (1 + exp(lambda * (time - tau)))) + 
+    ((-beta * c) / (1 + exp(phi * (time - tau - delta)))), 
+  
+  alpha + beta + c ~ 1, 
+  lambda + phi ~ 1, 
+  tau + delta ~ 1, 
+  nl = TRUE
+)
+
+m1_prior <- c(
+  ## Fix effects
+  prior(normal(1, 0.5), nlpar = alpha), 
+  prior(normal(-2.5, 0.5), nlpar = beta, ub = 0),
+  prior(normal(0.8, 0.2), nlpar = c, lb = 0),
+  prior(normal(-2, 0.2), nlpar = lambda, ub = 0),
+  prior(normal(-2, 0.2), nlpar = phi, ub = 0),
+  prior(normal(5, 0.5), nlpar = tau, lb = 0),
+  prior(normal(5, 0.5), nlpar = delta, lb = 0),
+  ## Sigma
+  prior(student_t(3, 0, 1), class = sigma, lb = 0)
+)
+
+m1_model <- brm(
+  formula = m1_formula,
+  data = model_data,
+  family = gaussian(),
+  prior = m1_prior,
[…]

### 2024-11-02 R/0-test-sim-data.R
add sample function

@@ -33,6 +33,12 @@ simulate_data <- function(n_id, n_time, noise = 10, seed = 1234) {
   )
 }
 
+sample_data <- function(data, n, id_var, na.rm = FALSE) {
+  na_fun <- function(x) return(x)
+  if (isTRUE(na.rm)) na_fun <- na.omit
+  na_fun(data)[, .SD[sort(sample.int(.N, n))], by = id_var]
+}
+
 ## Load the data
 data("rri_data")
 
@@ -41,21 +47,28 @@ rri_data[, rri_mean := mean(rr_denoised, na.rm = TRUE), by = id]
 rri_data[, rri_sd := sd(rr_denoised, na.rm = TRUE), by = id]
 rri_data[, rri_std := (rr_denoised - rri_mean) / rri_sd, by = id]
 
+m_data <- sample_data(rri_data, 200, "id", na.rm = TRUE)
+
 ## Check the 2d-density kernel of time-rri_std
-ggplot(rri_data, aes(time, rri_std)) +
-  stat_density_2d_filled() + 
+ggplot(m_data, aes(time, rri_std)) +
+  stat_density_2d_filled(show.legend = FALSE) + 
   scale_y_continuous(expand = c(0,0)) +
   scale_x_continuous(expand = c(0,0))
 ## Much more aligned across subjects. This will definitely improve model
 ## convergence
 
+ggplot(m_data[id %in% 1:5], aes(time, rri_std, col = ordered(id))) +
+  geom_line(show.legend = FALSE) +
+  scale_y_continuous(expand = c(0,0)) +
+  scale_x_continuous(expand = c(0,0)) +
+  theme_bw(base_line_size = 0)
 
 # Model with fixed effects only -------------------------------------------
 
 m_simple_prior <- c(
   ## Fix effects
   prior(normal(1, 0.5), nlpar = alpha), 
-  prior(normal(-2, 0.5), nlpar = beta, ub = 0),
+  prior(normal(-2.5, 0.5), nlpar = beta, ub = 0),
   prior(normal(0.8, 0.2), nlpar = c, lb = 0),
   prior(normal(-2, 0.2), nlpar = lambda, ub = 0),
   prior(normal(-2, 0.2), nlpar = phi, ub = 0),
@@ -76,7 +89,7 @@ m_simple_formula <- bf(
 
 m_simple <- brm(
   formula = m_simple_formula,
[…]

### 2024-11-02 R/6-fit-model-to-data.R

@@ -6,7 +6,13 @@ library(brms)
 library(data.table)
 library(ggpubr)
 
-## Load datasets ----
+sample_data <- function(data, n, id_var, na.rm = FALSE) {
+  na_fun <- function(x) return(x)
+  if (isTRUE(na.rm)) na_fun <- na.omit
+  na_fun(data)[, .SD[sort(sample.int(.N, n))], by = id_var]
+}
+
+## Load the data
 data("rri_data")
 
 ## Standardize as with the simulated dataset
@@ -14,17 +20,7 @@ rri_data[, rri_mean := mean(rr_denoised, na.rm = TRUE), by = id]
 rri_data[, rri_sd := sd(rr_denoised, na.rm = TRUE), by = id]
 rri_data[, rri_std := (rr_denoised - rri_mean) / rri_sd, by = id]
 
-model_data <- rri_data[!is.na(rri_std), .SD[sort(sample.int(.N, 200))], id]
-
-## Custom functions ----
-
-## Main equation
-RRi_model <- function(t, alpha, beta, c, lambda, phi, tau, delta) {
-  alpha +
-    (beta / (1 + exp(lambda * (t - tau)))) +
-    ((-beta * c) / (1 + exp(phi * (t - tau - delta))))
-}
-
+m_data <- sample_data(rri_data, 200, "id", na.rm = TRUE)
 
 # Build models ------------------------------------------------------------
 
@@ -57,7 +53,7 @@ m1_prior <- c(
 
 m1_model <- brm(
   formula = m1_formula,
-  data = model_data,
+  data = m_data,
   family = gaussian(),
   prior = m1_prior,
   seed = 1234,
@@ -104,7 +100,7 @@ m2_prior <- c(
 
 m2_model <- brm(
   formula = m2_formula,
-  data = model_data,
+  data = m_data,
   family = gaussian(),
   prior = m2_prior,
   seed = 1234,

### 2024-11-02 paper.qmd
new approach, a model for each subject

@@
 
-The priors were selected based on physiological constraints and are detailed in @eq-priors, ensuring in this way, identifiability of model parameters by constraining the parameter space to physiologically plausible values and ultimately, improving model convergence.
+To enhance the exploration of parameter space, each subject´s RRi data was standardize against his own mean and standard deviation, improving convergence and an efficient exploration of the posterior distribution in this high dimensional parameter space.
+
+The priors were chosen based on physiological constraints and graphical visualization of standardized RRi data.
+These priors are detailed in @eq-priors, ensuring in this way, identifiability of model parameters by constraining the parameter space to physiologically plausible values and ultimately, improving model convergence.
 
@@
 \begin{aligned} 
-\alpha &\sim \mathcal{N}(850, 50) \text{ with } \alpha \geq 0 \\
-\beta &\sim \mathcal{N}(-400, 50) \text{ with } \beta \leq 0 \\
-c &\sim \mathcal{N}(1, 0.2) \text{ with } c \geq 0 \\
-\lambda &\sim \mathcal{N}(-3, 3) \text{ with } \lambda \leq 0 \\
-\phi &\sim \mathcal{N}(-3, 3) \text{ with } \phi \leq 0 \\
+\alpha &\sim \mathcal{N}(1, 0.5) \\
+\beta &\sim \mathcal{N}(-2.5, 0.5) \text{ with } \beta \leq 0 \\
+c &\sim \mathcal{N}(0.8, 0.2) \text{ with } c \geq 0 \\
+\lambda &\sim \mathcal{N}(-2, 0.5) \text{ with } \lambda \leq 0 \\
+\phi &\sim \mathcal{N}(-2, 0.5) \text{ with } \phi \leq 0 \\
 \tau &\sim \mathcal{N}(5, 0.5) \text{ with } \tau \geq 0 \\
@@
 
-Simulated RRi dynamics based on the prior parameter distributions are shown in [Figure 1](#fig-prior-sim).
+Simulated standardized RRi dynamics based on prior parameter distributions are shown in [Figure 1](#fig-prior-sim).
 
@@
 
-The fitting process utilized five Markov Chain Monte Carlo (MCMC) chains, each consisting of 15,000 iterations with a burn-in period of 5,000 iterations, resulting in a total of 50,000 post-warmup samples.
+The fitting process utilized five Markov Chain Monte Carlo (MCMC) chains, each consisting of 10,000 iterations with a burn-in period of 5,000 iterations, resulting in a total of 25,000 post-warmup samples.
+The model was fitted for each subject´s standardized RRi data, allowing the estimation of the model parameters associated with individual RRi dynamics for each subject.
+Afterwards, we transformed back to the RRi scale the $\alpha$ and $\beta$ parameters, ensuring a physiologically meaningful interpretation of model estimation.
 

### 2024-11-02 R/1-prior-rri-model.R

@@ -40,26 +40,28 @@ sample_params <- function(n_id, n_rep, min, max) {
 
 ## Main model
 main_model <- bf(
-  rri ~ alpha + 
+  rri_std ~ alpha + 
     (beta / (1 + exp(lambda * (time - tau)))) + 
     ((-beta * c) / (1 + exp(phi * (time - tau - delta)))), 
   
-  alpha + beta + c ~ 1 | id, 
-  lambda + phi ~ 1 | id, 
-  tau + delta ~ 1 | id, 
+  alpha + beta + c ~ 1, 
+  lambda + phi ~ 1, 
+  tau + delta ~ 1, 
   nl = TRUE
 )
 
 # Define priors for the parameters
 priors <- c(
-  prior(normal(850, 50), nlpar = alpha, lb = 0),
-  prior(normal(-400, 50), nlpar = beta, ub = 0),
-  prior(normal(1, 0.2), nlpar = c, lb = 0),
-  prior(normal(-3, 3), nlpar = lambda, ub = 0),
-  prior(normal(-3, 3), nlpar = phi, ub = 0),
+  ## Fix effects
+  prior(normal(1, 0.5), nlpar = alpha), 
+  prior(normal(-2.5, 0.5), nlpar = beta, ub = 0),
+  prior(normal(0.8, 0.2), nlpar = c, lb = 0),
+  prior(normal(-2, 0.5), nlpar = lambda, ub = 0),
+  prior(normal(-2, 0.5), nlpar = phi, ub = 0),
   prior(normal(5, 0.5), nlpar = tau, lb = 0),
   prior(normal(5, 0.5), nlpar = delta, lb = 0),
-  prior(student_t(3, 0, 50), class = sigma, lb = 0)
+  ## Sigma
+  prior(student_t(3, 0, 1), class = sigma, lb = 0)
 )
 
 len_t <- 100
@@ -67,19 +69,9 @@ len_n <- 5
 
 new_data <- data.table(
   time = seq(0, 20, length.out = len_t),
-  alpha = sample_params(len_n, len_t, 500, 1500),
-  beta = sample_params(len_n, len_t, -800, -200),
-  c = sample_params(len_n, len_t, .5, 1.5),
-  lambda = sample_params(len_n, len_t, log(1-.9), log(1-.5)),
-  phi = sample_params(len_n, len_t, log(1-.8), log(1-.4)),
-  tau = sample_params(len_n, len_t, 4, 6),
-  delta = sample_params(len_n, len_t, 4, 6),
[…]

### 2024-11-02 R/5-2d-density-kernel.R

@@ -11,16 +11,8 @@ data("rri_data")
 
 # Generate figure ---------------------------------------------------------
 
-rri_data[, time_cat := cut(time, breaks = seq(0, 20, by = 1), 
-               labels = seq(0, 20, by = 1)[-1]) |> 
-           as.character() |> as.numeric()]
-
-estimate <- rri_data[, list(mean = mean(rr_denoised, na.rm = T),
-                            se = sd(rr_denoised, na.rm = T)), 
-                     list(time = as.numeric(as.character(time_cat)))]
-
-figure <- ggplot(rri_data, aes(time, rr_denoised)) +
-  stat_density_2d_filled(show.legend = FALSE, adjust = 1/3, na.rm = TRUE) +
+figure <- ggplot(rri_data, aes(time, rri_std)) +
+  stat_density_2d_filled(show.legend = FALSE, adjust = 0.75, na.rm = TRUE) +
   scale_x_continuous(expand = c(0,0), name = "Time (minutes)") +
   scale_y_continuous(expand = c(0,0), name = "RRi (ms)") +
   scale_fill_brewer() +

### 2024-11-02 R/6-fit-model-to-data.R

@@ -4,109 +4,35 @@
 library(ggplot2)
 library(brms)
 library(data.table)
-library(ggpubr)
-
-sample_data <- function(data, n, id_var, na.rm = FALSE) {
-  na_fun <- function(x) return(x)
-  if (isTRUE(na.rm)) na_fun <- na.omit
-  na_fun(data)[, .SD[sort(sample.int(.N, n))], by = id_var]
-}
 
 ## Load the data
 data("rri_data")
 
-## Standardize as with the simulated dataset
-rri_data[, rri_mean := mean(rr_denoised, na.rm = TRUE), by = id]
-rri_data[, rri_sd := sd(rr_denoised, na.rm = TRUE), by = id]
-rri_data[, rri_std := (rr_denoised - rri_mean) / rri_sd, by = id]
+m_data <- rri_data[!is.na(rri_std)]
 
-m_data <- sample_data(rri_data, 200, "id", na.rm = TRUE)
+# Load model --------------------------------------------------------------
 
-# Build models ------------------------------------------------------------
+m_model <- readRDS(file = "models/m_prior_only.RDS")
 
-## Model without Random intercepts ----
+# Loop all ID -------------------------------------------------------------
 
-### Model formula
-m1_formula <- bf(
-  rri_std ~ alpha + 
-    (beta / (1 + exp(lambda * (time - tau)))) + 
-    ((-beta * c) / (1 + exp(phi * (time - tau - delta)))), 
-  
-  alpha + beta + c ~ 1, 
-  lambda + phi ~ 1, 
-  tau + delta ~ 1, 
-  nl = TRUE
-)
+unique_ids <- m_data$id |> unique()
+list_models <- vector(mode = "list", length = length(unique_ids))
 
-m1_prior <- c(
-  ## Fix effects
-  prior(normal(1, 0.5), nlpar = alpha), 
-  prior(normal(-2.5, 0.5), nlpar = beta, ub = 0),
-  prior(normal(0.8, 0.2), nlpar = c, lb = 0),
-  prior(normal(-2, 0.2), nlpar = lambda, ub = 0),
-  prior(normal(-2, 0.2), nlpar = phi, ub = 0),
-  prior(normal(5, 0.5), nlpar = tau, lb = 0),
-  prior(normal(5, 0.5), nlpar = delta, lb = 0),
-  ## Sigma
[…]

### 2024-11-02 data-raw/process-data.R

@@ -89,5 +89,10 @@ if (FALSE) {
 
 rri_data[, file := NULL][]
 
+## Standardize as with the simulated dataset
+rri_data[, rri_mean := mean(rr_denoised, na.rm = TRUE), by = id][]
+rri_data[, rri_sd := sd(rr_denoised, na.rm = TRUE), by = id][]
+rri_data[, rri_std := (rr_denoised - rri_mean) / rri_sd, by = id][]
+
 save(rri_data, file = "data/rri_data.RData")
 fwrite(rri_data, file = "data/rri_data.csv")

### 2024-11-05 R/6-fit-model-to-data.R
models fitted

@@ -27,6 +27,8 @@ for (i in unique_ids) {
     seed = 1234,
     iter = 10000, warmup = 5000,
     chains = 5, cores = 5,
+    control = list(adapt_delta = 0.99,
+                   max_treedepth = 50),
     sample_prior = "no",
     file = paste0("models/id_level/m_id_",i,".RDS")
   )

### 2024-11-05 paper.qmd
update on paper methods for two-stage modeling

@@
 
-To enhance the exploration of parameter space, each subject´s RRi data was standardize against his own mean and standard deviation, improving convergence and an efficient exploration of the posterior distribution in this high dimensional parameter space.
+The gradient of the log-likelihood function for each parameter was computed during estimation using the brms R package, which employs the Stan probabilistic programming language.
+Convergence of the HMC chains was evaluated using the Gelman-Rubin diagnostic and trace plots to ensure reliable parameter estimates.
+This quantification of uncertainty is essential for assessing the stability of parameter estimates, particularly in the presence of noisy RRi data.
 
-The priors were chosen based on physiological constraints and graphical visualization of standardized RRi data.
-These priors are detailed in @eq-priors, ensuring in this way, identifiability of model parameters by constraining the parameter space to physiologically plausible values and ultimately, improving model convergence.
+The fitting process utilized five Markov Chain Monte Carlo (MCMC) chains, each consisting of 10,000 iterations with a burn-in period of 5,000 iterations, resulting in a total of 25,000 post-warmup samples.
+
+To enhance the exploration of parameter space we performed a two-stage analysis: individual-level and group-level estimates.
+
+### Individual-level analysis
+
+Firstly, each subject's RRi data $\text{RRi}_{i,t}$ was standardize against his own mean $\bar{\text{RRi}_i}$ and standard deviation $\sigma_{\text{RRi}_i}$ to improve convergence and exploration of the posterior distribution.
+The standardized RRi data $y_{i,t}$ for each time point $t$ was computed as:
+
+$$
+y_{i,t} = \frac{\text{RRi}_{i,t} - \bar{\text{RRi}_i}}{\sigma_{\text{RRi}_i}}
+$$
+This standardization allowed the model to focus on relative changes in RRi dynamics, independent of individual baseline differences.
+
+The model for each subject $i$ was then specified in terms of standardized RRi data $y_{i,t}$:
+
+$$
+y_{i,t} = \alpha_i + \frac{\beta_i}{1 + e^{\lambda_i \cdot (t - \tau_i)}} + \frac{-\beta_i \cdot c}{1 + e^{\phi_i \cdot (t - \tau_i - \delta_i)}} + \epsilon_{i,t}
+$$
+where $\alpha_i$, $\beta_i$, $c_i$, $\lambda_i$, $\phi_i$, $\tau_i$, $\delta_i$ are the individual-specific model parameters and $\epsilon_{i,t} \sim \mathcal{N}(0, \sigma^2)$ is the residual error term at each time point $t$.
+
+Afterwards, we transformed the estimated $\alpha$ and $\beta$ parameters back to the original RRi scale, ensuring a physiologically meaningful interpretation.
+The transformation for each subject $i$ is given by:
+
+$$
+\begin{aligned}
+\alpha_i^{\text{RRi}} &= \alpha_i \cdot \sigma_{\text{RRi}_i} + \bar{\text{RRi}}_i \\
+\beta_i^{\text{RRi}} &= \beta_i \cdot \sigma_{\text{RRi}_i}
+\end{aligned}
+$$
+
+Priors were chosen based on physiological constraints and graphical visualization of standardized RRi data, ensuring identifiability of model parameters by constraining the parameter space to plausible values, which improves model convergence.
+The prior distributions are defined as follows:
 
@@
 \end{aligned}
-$$ {#eq-priors}
+$$
 
@@
 
-[**Figure 1**]{#fig-prior-sim}. Predicted RRi response to exercise, illustrating uncertainty in model parameters from the prior distribution. Shaded areas represent 95%, 80%, and 60% quantile CI.
+[**Figure 1**]{#fig-prior-sim}.
+Prior predicted RRi response to exercise, illustrating uncertainty in model parameters from the prior distribution.
+Shaded areas represent 95%, 80%, and 60% quantile CI.
 
-The gradient of the log-likelihood function for each parameter was computed during estimation using the brms R package, which employs the Stan probabilistic programming language.
-Convergence of the HMC chains was evaluated using the Gelman-Rubin diagnostic and trace plots to ensure reliable parameter estimates.
-This quantification of uncertainty is essential for assessing the stability of parameter estimates, particularly in the presence of noisy RRi data.
+### Group-level analysis
 
-The fitting process utilized five Markov Chain Monte Carlo (MCMC) chains, each consisting of 10,000 iterations with a burn-in period of 5,000 iterations, resulting in a total of 25,000 post-warmup samples.
-The model was fitted for each subject´s standardized RRi data, allowing the estimation of the model parameters associated with individual RRi dynamics for each subject.
-Afterwards, we transformed back to the RRi scale the $\alpha$ and $\beta$ parameters, ensuring a physiologically meaningful interpretation of model estimation.
+After obtaining the posterior distribution for each subject's parameters, the mean and standard deviation for each parameter were calculated.
+These estimates were then used as input data to create a multivariate hierarchical model, capturing variability at both the subject and group levels.
+The model is described as follows:
+
+For each subject $i$, the parameter vector $\theta_i = (\alpha_i, \beta_i, c_i, \lambda_i, \phi_i, \tau_i, \delta_i)^\top$ is modeled as:
+
+$$
+\theta_{i,j} \sim \mathcal{N}(\mu_{j} + b_{i,j}, \sigma_j^2)
+$$
+
+where $\mu_{j}$ is the group-level mean for parameter $j$, $b_{i,j}$ represents the subject-level random effect for subject $i$ on parameter $j$, and $\sigma_j^2$ is the residual variance for parameter $j$.
+
+The subject-level effects $b_{i,j}$ are assumed to follow a multivariate normal distribution:
+
+$$
+\mathbf{b}_i = (b_{i, \alpha}, b_{i, \beta}, b_{i, c}, b_{i, \lambda}, b_{i, \phi}, b_{i, \tau}, b_{i, \delta})^\top \sim \mathcal{N}(\mathbf{0}, \Sigma_b)
+$$
+
+where $\Sigma_b$ is the covariance matrix that captures correlations between the random effects for different parameters across subjects.
+
+This hierarchical structure enables us to capture individual variability through subject-level random effects while also estimating group-level effects across all parameters, thus providing estimates into both subject-level and population-level model parameters.
 
@@
 [**Figure 5**]{#fig-2d-kernel-density}.
-2D kernel density of RRi dynamics over time from a sample of individuals subjected to a rest-exercise-rest protocol.
+Upper panel shows the mean and standard deviation (SD) from each of the subject's RRi recordings, used for the standardization process.
+Lower panel indicate the 2D kernel density of standardized RRi dynamics over time from a sample of individuals subjected to a rest-exercise-rest protocol.
 Darker colors indicate greater probability density.
@@
 
-In this study, we present a non-linear model designed to capture R-R interval (RRi) fluctuations during exercise and recovery, providing an in-depth view of cardiac autonomic dynamics.
+In this study, we present a non-linear model designed to RRi fluctuations during exercise and recovery, providing an in-depth view of cardiac autonomic dynamics.
 By focusing on RRi rather than aggregate HRV metrics, our approach enables real-time tracking of cardiovascular responses to physical stress, which has implications for both clinical and athletic applications.

### 2024-11-05 R/4-rate-sensitivity.R
change linewidth in rate sensitivity plot

@@ -91,7 +91,7 @@ sim_data[, phi := factor(phi,
 
 figure <- ggplot(sim_data, aes(time, hrv)) +
   facet_wrap(~ phi, labeller = labeller(phi = label_parsed)) +
-  geom_line(aes(col = as.factor(x = lambda), group = id), linewidth = 1) +
+  geom_line(aes(col = as.factor(x = lambda), group = id), linewidth = 0.7) +
   geom_vline(xintercept = c(5,7), color = "gray50", linewidth = .25) +
   scale_y_continuous(breaks = c(500, 600, 700, 800)) +
   scale_alpha_manual(values = c(1, .85, .7, .55)) +

### 2024-11-05 R/5-2d-density-kernel.R
add mean and sd distributions in 2d kernel plot
- to better understand the transformation made

@@ -11,15 +11,49 @@ data("rri_data")
 
 # Generate figure ---------------------------------------------------------
 
-figure <- ggplot(rri_data, aes(time, rri_std)) +
+figure_a <- ggplot(rri_data, aes(time, rri_std)) +
   stat_density_2d_filled(show.legend = FALSE, adjust = 0.75, na.rm = TRUE) +
   scale_x_continuous(expand = c(0,0), name = "Time (minutes)") +
-  scale_y_continuous(expand = c(0,0), name = "RRi (ms)") +
+  scale_y_continuous(expand = c(0,0), name = "Mean Standardized RRi",
+                     breaks = -3:2) +
   scale_fill_brewer() +
   theme_classic(base_size = 12) +
   theme(axis.text = element_text(size = rel(0.8)))
 
+figure_b <- 
+  rri_data |> 
+  melt.data.table(
+    id.vars = "id", 
+    measure.vars = c("rri_mean","rri_sd")
+  ) |> within({
+    levels(variable) <- c("Mean", "SD")
+  }) |> 
+  ggplot(aes(value)) +
+  facet_grid(~ variable, scales = "free_x") +
+  geom_histogram(aes(y = after_stat(density), 
+                     fill = variable), 
+                 bins = 15, show.legend = FALSE,
+                 col = "gray10") +
+  stat_density(geom = "line",
+               adjust = 4,
+               col = "gray10") +
+  geom_boxplot(aes(y = -.001, fill = variable),
+               width = .0005, orientation = "y", staplewidth = 0,
+               show.legend = FALSE) +
+  scale_fill_brewer() +
+  scale_y_continuous(breaks = NULL, name = NULL) +
+  scale_x_continuous(name = "RRi (ms)", n.breaks = 4) +
+  theme_classic(base_size = 12) +
+  theme(axis.text = element_text(size = rel(0.8)),
+        plot.margin = margin(0.05,.15,0,.5, "in"),
+        strip.background = element_rect(colour = NA, fill = "gray90"))
+
+figure <- ggpubr::ggarrange(
+  figure_b, figure_a, ncol = 1, nrow = 2,
+  heights = c(2,3)
+)
+
 ggsave("figures/2d-density-kernel.pdf", figure, "pdf", 
-       width = 7, height = 4, units = "in")
+       width = 7, height = 5, units = "in")
 ggsave("figures/2d-density-kernel.png", figure, "png", 
[…]

### 2024-11-05 R/7-meta-analysis.R
add descriptives table

@@ -1,0 +1,82 @@
+N <- 272
+draws = 1
+
+library(data.table)
+
+posteriors <- data.table(
+  id = rep(1:N, each = draws),
+  alpha = rnorm(N * draws, 840, 50),
+  alphaSigma = rnorm(N * draws, 50, 5) |> abs(),
+  beta = rnorm(N * draws, -400, 50),
+  betaSigma = rnorm(N * draws, 50, 5) |> abs(),
+  c = rnorm(N * draws, 0.8, 0.05),
+  cSigma = rnorm(N * draws, 0.05, 0.05) |> abs(),
+  lambda = rnorm(N * draws, -2, 0.1),
+  lambdaSigma = rnorm(N * draws, 0.1, 0.1) |> abs(),
+  phi = rnorm(N * draws, -1.5, 0.1),
+  phiSigma = rnorm(N * draws, 0.05, 0.1) |> abs(),
+  tau = rnorm(N * draws, 5, 0.1),
+  tauSigma = rnorm(N * draws, 0.1, 0.1) |> abs(),
+  delta = rnorm(N * draws, 5, 0.1),
+  deltaSigma = rnorm(N * draws, 0.1, 0.1) |> abs()
+)
+
+library(brms)
+
+m_mod_alpha <- bf(alpha | se(alphaSigma, sigma = TRUE) ~ 1 + (1 | id))
+m_mod_beta <- bf(beta | se(betaSigma, sigma = TRUE) ~ 1 + (1 | id))
+m_mod_c <- bf(c | se(cSigma, sigma = TRUE) ~ 1 + (1 | id))
+m_mod_lambda <- bf(lambda | se(lambdaSigma, sigma = TRUE) ~ 1 + (1 | id))
+m_mod_phi <- bf(phi | se(phiSigma, sigma = TRUE) ~ 1 + (1 | id))
+m_mod_tau <- bf(tau | se(tauSigma, sigma = TRUE) ~ 1 + (1 | id))
+m_mod_delta <- bf(delta | se(deltaSigma, sigma = TRUE) ~ 1 + (1 | id))
+
+priors <- c(
+  prior(normal(-400, 100), class = "Intercept", resp = "beta", ub = 0),
+  prior(normal(1, 0.5), class = "Intercept", resp = "c", lb = 0),
+  prior(normal(-2.3, 0.5), class = "Intercept", resp = "lambda", ub = 0),
[…]

### 2024-11-05 data-raw/process-data.R

@@ -1,98 +1,0 @@
-
-# Prepare workspace -------------------------------------------------------
-
-## Load packages
-library(data.table)
-library(brms)
-library(ggplot2)
-
-## Selection of txt files
-files <- list.files(path = "data-raw/hrv_data", full.names = TRUE, pattern = "\\.txt$")
-
-## Data import
-rri_data <- lapply(files, \(i) {
-  rr_interval <- as.double(x = readLines(i))
-  not_na <- !is.na(rr_interval)
-  rr_interval <- rr_interval[not_na & between(rr_interval, 400, 1200)]
-  time <- cumsum(rr_interval)/60000
-  data.table(rr_interval = rr_interval[time <= 20],
-             time = time[time <= 20],
-             file = gsub("data-raw/", "", i, fixed = TRUE))
-}) |> 
-  rbindlist(idcol = "id")
-
-## Data filtering
-id_to_keep <- rri_data[, max(time) > 10, id][V1 == TRUE, id]
-id_to_move <- rri_data[, max(time) > 10, id][V1 == FALSE, id]
-files_to_move <- rri_data[id %in% id_to_keep, files[!files %in% paste0("data-raw/", file)]]
-
-
-if (length(id_to_move) > 0) {
-  for (i in id_to_move) {
-    file <- rri_data[id == i, unique(file)]
-    from <- paste0("data-raw/", file)
-    to <- paste0("data-raw/hrv_data/unused/", gsub("hrv_data/", "", file))
-    file.copy(from = from, to = to, overwrite = TRUE)
-    file.remove(from)
-  }
-}
-
-if (length(files_to_move) > 0) {
-  for (i in files_to_move) {
-    to <- paste0("data-raw/hrv_data/unused/", gsub("data-raw/hrv_data/", "", i))
-    file.copy(from = i, to = to, overwrite = TRUE)
-    file.remove(i)
-  }
-}
-
-rm(files, files_to_move, id_to_move)
-
-rri_data <- rri_data[id %in% id_to_keep]
-
-
-# Processing signal data --------------------------------------------------
-
-## Denoising the signal data
-denoise <- function(x, w = 0.1, n = 3) {
-  signal::filtfilt(filt = signal::butter(n, w), x)
-}
-
-## Applying Butterworth 10 Hz low-pass filter
-rri_data[, j = rr_denoised := denoise(rr_interval), id]
-
-trim <- function(x, abs) {
[…]

### 2024-11-05 data-raw/process-descriptives.R

@@ -1,0 +1,59 @@
+library(readxl)
+library(data.table)
+
+
+descriptives <- (read_excel(path = "data-raw/descriptives.xlsx") |> 
+  as.data.table())[1:272,-10L]
+
+descriptives |> names() |> dput()
+
+names(descriptives) <- 
+  c("Sex", "Age", "SBP (mm hg)", "DBP (mm hg)", 
+    "MAP (mm hg)", "PP (mm hg)", "BMI", "Weight (kg)", 
+    "Height (cm)")
+
+descriptives$id <- 1:nrow(descriptives)
+descriptives[, Sex := factor(Sex, levels = 1:2, labels = c("Female", "Male"))]
+
+desc_long <- descriptives |> 
+  melt(id.vars = "id")
+
+mean_sd <- function(x) {
+  x <- as.numeric(x)
+  if (all(is.na(x))) {
+    return("—")
+  }
+  paste0(mean(x) |> round(2), " ± ",
+         sd(x) |> round(2))
+}
+
+tbl_overall <- desc_long[
+  j = list(Overall = mean_sd(value)), 
+  by = variable]
+
+desc_long_sex <- descriptives |> 
+  melt(id.vars = c("id", "Sex"))
+
+tbl_sex <- desc_long_sex[
+  j = tapply(value, Sex, mean_sd, simplify = F), 
+  by = variable]
+
+tbl_full <- merge.data.table(tbl_overall, tbl_sex, by = "variable", all = T)
+
+n_perc <- function(x, N) {
+  n <- length(x)
+  perc <- paste0("(",round(n * 100 / N, 1), "%)")
+  paste(n, perc)
+}
+
+sex_var <- descriptives$Sex
+
+tbl_full[
+  i = variable == "Sex", 
+  j = `:=`(
+    Female = n_perc(x = sex_var[sex_var == "Female"], N = 272),
+    Male = n_perc(x = sex_var[sex_var == "Male"], N = 272)
+  )
+]
+
+tbl_full

### 2024-11-05 data-raw/process-hrv-data.R

@@ -1,0 +1,98 @@
+
+# Prepare workspace -------------------------------------------------------
+
+## Load packages
+library(data.table)
+library(brms)
+library(ggplot2)
+
+## Selection of txt files
+files <- list.files(path = "data-raw/hrv_data", full.names = TRUE, pattern = "\\.txt$")
+
+## Data import
+rri_data <- lapply(files, \(i) {
+  rr_interval <- as.double(x = readLines(i))
+  not_na <- !is.na(rr_interval)
+  rr_interval <- rr_interval[not_na & between(rr_interval, 400, 1200)]
+  time <- cumsum(rr_interval)/60000
+  data.table(rr_interval = rr_interval[time <= 20],
+             time = time[time <= 20],
+             file = gsub("data-raw/", "", i, fixed = TRUE))
+}) |> 
+  rbindlist(idcol = "id")
+
+## Data filtering
+id_to_keep <- rri_data[, max(time) > 10, id][V1 == TRUE, id]
+id_to_move <- rri_data[, max(time) > 10, id][V1 == FALSE, id]
+files_to_move <- rri_data[id %in% id_to_keep, files[!files %in% paste0("data-raw/", file)]]
+
+
+if (length(id_to_move) > 0) {
+  for (i in id_to_move) {
+    file <- rri_data[id == i, unique(file)]
+    from <- paste0("data-raw/", file)
+    to <- paste0("data-raw/hrv_data/unused/", gsub("hrv_data/", "", file))
+    file.copy(from = from, to = to, overwrite = TRUE)
+    file.remove(from)
+  }
+}
+
+if (length(files_to_move) > 0) {
+  for (i in files_to_move) {
+    to <- paste0("data-raw/hrv_data/unused/", gsub("data-raw/hrv_data/", "", i))
+    file.copy(from = i, to = to, overwrite = TRUE)
+    file.remove(i)
+  }
+}
+
+rm(files, files_to_move, id_to_move)
+
+rri_data <- rri_data[id %in% id_to_keep]
+
+
+# Processing signal data --------------------------------------------------
+
+## Denoising the signal data
+denoise <- function(x, w = 0.1, n = 3) {
+  signal::filtfilt(filt = signal::butter(n, w), x)
+}
+
+## Applying Butterworth 10 Hz low-pass filter
+rri_data[, j = rr_denoised := denoise(rr_interval), id]
+
+trim <- function(x, abs) {
[…]

### 2024-11-06 paper.qmd
stage 1 ready

@@
 
-<!--- TABLE 2 HERE --->
-[**Table 2**]{#tbl-sample-characteristics}. Sample characteristics from which, continuous RRi monitoring data was collected during a rest-exercise-rest protocol.
+|Characteristic |Overall        |Female         |Male           |
+|:--------------|:--------------|:--------------|:--------------|
+|Sex            |—              |217 (79.8%)    |55 (20.2%)     |
+|Age            |71.14 ± 6.03   |70.73 ± 6.27   |72.73 ± 4.7    |
+|SBP (mm hg)    |130.23 ± 17.07 |129.58 ± 17.37 |132.8 ± 15.69  |
+|DBP (mm hg)    |77.1 ± 9.58    |76.68 ± 9.83   |78.75 ± 8.4    |
+|MAP (mm hg)    |94.81 ± 10.69  |94.31 ± 10.95  |96.76 ± 9.45   |
+|PP (mm hg)     |53.14 ± 14.07  |52.9 ± 14.26   |54.05 ± 13.38  |
+|BMI            |30.66 ± 5.43   |30.7 ± 5.64    |30.53 ± 4.53   |
+|Weight (kg)    |75.06 ± 14.23  |73.88 ± 14.09  |79.69 ± 13.95  |
+|Height (cm)    |156.56 ± 9.18  |155.29 ± 8.46  |161.55 ± 10.24 |
+
+[**Table 2**]{#tbl-sample-characteristics}. Sample characteristics from which, continuous RRi monitoring data was collected during a rest-exercise-rest protocol. Data is presented as Mean ± SD.
 
@@
 [**Figure 5**]{#fig-2d-kernel-density}.
-Upper panel shows the mean and standard deviation (SD) from each of the subject's RRi recordings, used for the standardization process.
-Lower panel indicate the 2D kernel density of standardized RRi dynamics over time from a sample of individuals subjected to a rest-exercise-rest protocol.
+Left panel shows the mean and standard deviation (SD) from each of the subject's RRi recordings, used for the standardization process.
+Right panel indicate the 2D kernel density of standardized RRi dynamics over time from a sample of individuals subjected to a rest-exercise-rest protocol.
 Darker colors indicate greater probability density.
 The contrary can be said about lighter colors.
+
+### Individual-level estimates
 

### 2024-11-06 R/5-2d-density-kernel.R

@@ -17,7 +17,7 @@ figure_a <- ggplot(rri_data, aes(time, rri_std)) +
   scale_y_continuous(expand = c(0,0), name = "Mean Standardized RRi",
                      breaks = -3:2) +
   scale_fill_brewer() +
-  theme_classic(base_size = 12) +
+  theme_classic(base_size = 14) +
   theme(axis.text = element_text(size = rel(0.8)))
 
 figure_b <- 
@@ -29,28 +29,29 @@ figure_b <-
     levels(variable) <- c("Mean", "SD")
   }) |> 
   ggplot(aes(value)) +
-  facet_grid(~ variable, scales = "free_x") +
-  geom_histogram(aes(y = after_stat(density), 
+  facet_wrap(~ variable, nrow = 2, scales = "free_x") +
+  geom_histogram(aes(y = after_stat(ndensity), 
                      fill = variable), 
                  bins = 15, show.legend = FALSE,
                  col = "gray10") +
   stat_density(geom = "line",
-               adjust = 4,
+               aes(y = after_stat(ndensity)),
+               adjust = 4, 
                col = "gray10") +
-  geom_boxplot(aes(y = -.001, fill = variable),
-               width = .0005, orientation = "y", staplewidth = 0,
-               show.legend = FALSE) +
+  geom_boxplot(aes(y = -.05, fill = variable),
+               width = .05, orientation = "y", staplewidth = 0,
+               show.legend = FALSE, outlier.color = NA) +
   scale_fill_brewer() +
-  scale_y_continuous(breaks = NULL, name = NULL) +
+  scale_y_continuous(breaks = NULL, name = NULL, expand = c(0.025,0,0.1,0)) +
   scale_x_continuous(name = "RRi (ms)", n.breaks = 4) +
-  theme_classic(base_size = 12) +
+  theme_classic(base_size = 14) +
   theme(axis.text = element_text(size = rel(0.8)),
-        plot.margin = margin(0.05,.15,0,.5, "in"),
+        plot.margin = margin(0.0825,.25,0.0825,.25, "in"),
         strip.background = element_rect(colour = NA, fill = "gray90"))
 
 figure <- ggpubr::ggarrange(
-  figure_b, figure_a, ncol = 1, nrow = 2,
-  heights = c(2,3)
+  figure_b, figure_a, ncol = 2, nrow = 1,
+  widths = c(1,2)
 )
 
 ggsave("figures/2d-density-kernel.pdf", figure, "pdf", 

### 2024-11-06 R/6-fit-model-to-data.R

@@ -7,34 +7,77 @@ library(data.table)
 
 ## Load the data
 data("rri_data")
-
 m_data <- rri_data[!is.na(rri_std)]
 
-# Load model --------------------------------------------------------------
+if (!file.exists("models/stage_1_model.RData")) {
 
-m_model <- readRDS(file = "models/m_prior_only.RDS")
-
-# Loop all ID -------------------------------------------------------------
-
-unique_ids <- m_data$id |> unique()
-list_models <- vector(mode = "list", length = length(unique_ids))
-
-for (i in unique_ids) {
-  print(paste0("Now on id ", i))
-  list_models[[i]] <- update(
-    object = m_model,
-    newdata = m_data[id == i],
-    seed = 1234,
-    iter = 10000, warmup = 5000,
-    chains = 5, cores = 5,
-    control = list(adapt_delta = 0.99,
-                   max_treedepth = 50),
-    sample_prior = "no",
-    file = paste0("models/id_level/m_id_",i,".RDS")
-  )
+  # Load model --------------------------------------------------------------
+  m_model <- readRDS(file = "models/m_prior_only.RDS")
+  
+  # Identify unique ID to loop for ------------------------------------------
+  unique_ids <- m_data$id |> unique()
+  list_models <- vector(mode = "list", length = length(unique_ids))
+  
+  # Loop all ID -------------------------------------------------------------
+  for (i in unique_ids) {
+    print(paste0("Now on id ", i))
+    file <- paste0("models/id_level/m_id_",i,".RDS")
+    
+    ## If model already exists, load it, then jump to next ID
+    if (file.exists(file)) {
+      list_models[[i]] <- readRDS(file)
+      next
+    }
+    
+    ## If not, sample the posterior for that ID
+    list_models[[i]] <- update(
+      object = m_model,
+      newdata = m_data[id == i],
+      seed = 1234,
+      iter = 10000, warmup = 5000,
+      chains = 5, cores = 5,
+      control = list(adapt_delta = 0.99,
+                     max_treedepth = 50),
+      sample_prior = "no",
+      file = file
+    )
+  }
+  
+  ## Merge all posterior samples into a single data.frame
+  df_models <- lapply(list_models, as_draws_df) |> 
+    rbindlist(idcol = "id")
+  
[…]

### 2024-11-06 R/7-meta-analysis.R

@@ -1,70 +1,66 @@
-N <- 272
-draws = 1
 
+# Prepare workspace -------------------------------------------------------
+
+## Load libraries
 library(data.table)
-
-posteriors <- data.table(
-  id = rep(1:N, each = draws),
-  alpha = rnorm(N * draws, 840, 50),
-  alphaSigma = rnorm(N * draws, 50, 5) |> abs(),
-  beta = rnorm(N * draws, -400, 50),
-  betaSigma = rnorm(N * draws, 50, 5) |> abs(),
-  c = rnorm(N * draws, 0.8, 0.05),
-  cSigma = rnorm(N * draws, 0.05, 0.05) |> abs(),
-  lambda = rnorm(N * draws, -2, 0.1),
-  lambdaSigma = rnorm(N * draws, 0.1, 0.1) |> abs(),
-  phi = rnorm(N * draws, -1.5, 0.1),
-  phiSigma = rnorm(N * draws, 0.05, 0.1) |> abs(),
-  tau = rnorm(N * draws, 5, 0.1),
-  tauSigma = rnorm(N * draws, 0.1, 0.1) |> abs(),
-  delta = rnorm(N * draws, 5, 0.1),
-  deltaSigma = rnorm(N * draws, 0.1, 0.1) |> abs()
-)
-
 library(brms)
+library(ggplot2)
 
-m_mod_alpha <- bf(alpha | se(alphaSigma, sigma = TRUE) ~ 1 + (1 | id))
-m_mod_beta <- bf(beta | se(betaSigma, sigma = TRUE) ~ 1 + (1 | id))
-m_mod_c <- bf(c | se(cSigma, sigma = TRUE) ~ 1 + (1 | id))
-m_mod_lambda <- bf(lambda | se(lambdaSigma, sigma = TRUE) ~ 1 + (1 | id))
-m_mod_phi <- bf(phi | se(phiSigma, sigma = TRUE) ~ 1 + (1 | id))
-m_mod_tau <- bf(tau | se(tauSigma, sigma = TRUE) ~ 1 + (1 | id))
-m_mod_delta <- bf(delta | se(deltaSigma, sigma = TRUE) ~ 1 + (1 | id))
+## Load data
+data("rri_data")
+m_data <- rri_data[!is.na(rri_std)]
+
+## Load adjusted model parameters
+id_data <- fread(file = "models/id_level_posterior.csv")
+
+
+# Fit model ---------------------------------------------------------------
+
+
[…]

### 2024-11-06 data-raw/process-descriptives.R

@@ -56,4 +56,16 @@ tbl_full[
   )
 ]
 
-tbl_full
+tbl_full |> 
+  knitr::kable(col.names = c("Characteristic", "Overall", "Female", "Male"))
+#> |Characteristic |Overall        |Female         |Male           |
+#> |:--------------|:--------------|:--------------|:--------------|
+#> |Sex            |—              |217 (79.8%)    |55 (20.2%)     |
+#> |Age            |71.14 ± 6.03   |70.73 ± 6.27   |72.73 ± 4.7    |
+#> |SBP (mm hg)    |130.23 ± 17.07 |129.58 ± 17.37 |132.8 ± 15.69  |
+#> |DBP (mm hg)    |77.1 ± 9.58    |76.68 ± 9.83   |78.75 ± 8.4    |
+#> |MAP (mm hg)    |94.81 ± 10.69  |94.31 ± 10.95  |96.76 ± 9.45   |
+#> |PP (mm hg)     |53.14 ± 14.07  |52.9 ± 14.26   |54.05 ± 13.38  |
+#> |BMI            |30.66 ± 5.43   |30.7 ± 5.64    |30.53 ± 4.53   |
+#> |Weight (kg)    |75.06 ± 14.23  |73.88 ± 14.09  |79.69 ± 13.95  |
+#> |Height (cm)    |156.56 ± 9.18  |155.29 ± 8.46  |161.55 ± 10.24 |

### 2024-11-06 R/6-fit-model-to-data.R
time to fit stage 2 model

@@ -9,7 +9,7 @@ library(data.table)
 data("rri_data")
 m_data <- rri_data[!is.na(rri_std)]
 
-if (!file.exists("models/stage_1_model.RData")) {
+if (!file.exists("models/stage_1_model.RDS")) {
 
   # Load model --------------------------------------------------------------
   m_model <- readRDS(file = "models/m_prior_only.RDS")
@@ -48,11 +48,11 @@ if (!file.exists("models/stage_1_model.RData")) {
     rbindlist(idcol = "id")
   
   ## Save the resulting id-wise posterior in a file
-  saveRDS(df_models, file = "models/stage_1_model.RData", compress = "xz")
+  saveRDS(df_models, file = "models/stage_1_model.RDS", compress = "xz")
   
 } else {
   ## Load the id-wise posterior file
-  df_models <- readRDS(file = "models/stage_1_model.RData")
+  df_models <- readRDS(file = "models/stage_1_model.RDS")
 }
 
 ## Transform alpha and beta parameters

### 2024-11-06 R/7-meta-analysis.R

@@ -60,19 +60,19 @@ m_fit <- brm(formula = m_model,
                    max_treedepth = 50)
     )
 
-loo(m_fit)
+saveRDS(m_fit, file = "stage_2_model.RDS")
 
-
-as_draws_df(m_fit_alpha) |> 
-  # tidybayes::spread_draws(
-  #   r_id[subject,], Intercept
-  # ) |> 
-  # transform(id_intercept = Intercept + r_id) |> 
-  ggplot(aes(x = b_Intercept)) +
-  ggdist::stat_halfeye(
-    fill = "lightblue",
-    density = "unbounded", trim = FALSE) +
-  scale_y_continuous(expand = c(0.02,0), breaks = NULL, name = NULL) +
-  scale_x_continuous(expand = c(0.2,0), n.breaks = 6) +
-  theme_classic() +
-  labs(x = expression(symbol(E)~"["*alpha*"]"))
+# 
+# as_draws_df(m_fit_alpha) |> 
+#   # tidybayes::spread_draws(
+#   #   r_id[subject,], Intercept
+#   # ) |> 
+#   # transform(id_intercept = Intercept + r_id) |> 
+#   ggplot(aes(x = b_Intercept)) +
+#   ggdist::stat_halfeye(
+#     fill = "lightblue",
+#     density = "unbounded", trim = FALSE) +
+#   scale_y_continuous(expand = c(0.02,0), breaks = NULL, name = NULL) +
+#   scale_x_continuous(expand = c(0.2,0), n.breaks = 6) +
+#   theme_classic() +
+#   labs(x = expression(symbol(E)~"["*alpha*"]"))

### 2024-11-07 paper.qmd
analysis script ready

@@
 |Age            |71.14 ± 6.03   |70.73 ± 6.27   |72.73 ± 4.7    |
-|SBP (mm hg)    |130.23 ± 17.07 |129.58 ± 17.37 |132.8 ± 15.69  |
-|DBP (mm hg)    |77.1 ± 9.58    |76.68 ± 9.83   |78.75 ± 8.4    |
-|MAP (mm hg)    |94.81 ± 10.69  |94.31 ± 10.95  |96.76 ± 9.45   |
-|PP (mm hg)     |53.14 ± 14.07  |52.9 ± 14.26   |54.05 ± 13.38  |
+|SBP (mm Hg)    |130.23 ± 17.07 |129.58 ± 17.37 |132.8 ± 15.69  |
+|DBP (mm Hg)    |77.1 ± 9.58    |76.68 ± 9.83   |78.75 ± 8.4    |
+|MAP (mm Hg)    |94.81 ± 10.69  |94.31 ± 10.95  |96.76 ± 9.45   |
+|PP (mm Hg)     |53.14 ± 14.07  |52.9 ± 14.26   |54.05 ± 13.38  |
 |BMI            |30.66 ± 5.43   |30.7 ± 5.64    |30.53 ± 4.53   |
@@
 
-In summary, this study introduces a non-linear model of RRi dynamics that captures complex, real-time fluctuations reflecting the autonomic nervous system’s response to exercise and recovery.
+In summary, this study introduces a non-linear model of RRi dynamics that captures complex, real-time fluctuations reflecting the autonomic nervous system's response to exercise and recovery.
 By focusing on RRi, the model offers enhanced sensitivity to physiological changes, making it well-suited for personalized applications in clinical and athletic settings.

### 2024-11-07 R/7-meta-analysis.R

@@ -13,66 +13,82 @@ m_data <- rri_data[!is.na(rri_std)]
 ## Load adjusted model parameters
 id_data <- fread(file = "models/id_level_posterior.csv")
 
+## Custom functions
+fit_model <- function(model, priors) {
+  file = paste0("models/group_level/",
+                deparse(substitute(model)),
+                ".RDS")
+  
+  if (file.exists(file)) {
+    m_fit <- readRDS(file)
+  } else {
+    m_fit <- brms::brm(formula = model,
+                       data = id_data,
+                       prior = priors,
+                       family = gaussian(),
+                       chains = 5, cores = 5, seed = 1234,
+                       iter = 10000, warmup = 5000,
+                       control = list(adapt_delta = .99,
+                                      max_treedepth = 50))
+    
+    saveRDS(object = m_fit, file = file, compress = "xz")
+  }
+  
+  return(m_fit)
+}
+
 
 # Fit model ---------------------------------------------------------------
 
+## Models
+m_alpha_model <- bf(alphaMu | se(alphaSE, sigma = TRUE) ~ 1 + (1 | id))
+m_beta_model <- bf(betaMu | se(betaSE, sigma = TRUE) ~ 1 + (1 | id))
+m_c_model <- bf(cMu | se(cSE, sigma = TRUE) ~ 1 + (1 | id))
+m_lambda_model <- bf(lambdaMu | se(lambdaSE, sigma = TRUE) ~ 1 + (1 | id))
+m_phi_model <- bf(phiMu | se(phiSE, sigma = TRUE) ~ 1 + (1 | id))
+m_tau_model <- bf(tauMu | se(tauSE, sigma = TRUE) ~ 1 + (1 | id))
+m_delta_model <- bf(deltaMu | se(deltaSE, sigma = TRUE) ~ 1 + (1 | id))
+m_sigma_model <- bf(sigmaMu | se(sigmaSE, sigma = TRUE) ~ 1 + (1 | id))
 
-m_model <- bf(alphaMu | se(alphaSE, sigma = TRUE) ~ 1 + (1 | id)) + 
-  bf(betaMu | se(betaSE, sigma = TRUE) ~ 1 + (1 | id)) + 
[…]

### 2024-11-11 paper.qmd
performance metrics and group level estimates

--- 2024-11-06 paper.qmd
@@
 
-**Title**: Modeling Nonlinear R-R Interval Dynamics during Exercise: A Novel Logistic Approach to Cardiovascular Recovery.
+**Title**: Enhancing Cardiovascular Monitoring: A Non-Linear Approach to RR Interval Dynamics in Exercise and Recovery.
 
-**Authors**: Matías Castillo-Aguilar^1,2^, [...], Cristian Núñez-Espinoza^1,3^\*.
+**Authors**: : Matías Castillo-Aguilar^1,2^, Diego Mabe-Castro^1,2^, David Medina^1^, Cristian Núñez-Espinosa^1,3^\*.
 
@@
 
-**Objective**: [...]. **Methods**: [...]. **Results**: [...]. **Conclusion**: [...].
+**Objective**: This study aims to develop and validate a novel non-linear model for characterizing RR interval (RRi) dynamics during exercise and recovery, enhancing the understanding of the cardiac autonomic function and its implications for both clinical practice and athletic training.
+**Methods**: A cohort of 272 participants underwent a validated exercise protocol, during which continuous heart rate data, including RR intervals, were collected.
+A non-linear logistic approach was employed to model the fluctuations in RRi.
+Sobol sensitivity analysis was conducted to assess the influence of model parameters on RR dynamics.
+**Results**: The proposed model successfully captured the complex fluctuations in RRi, demonstrating significant sensitivity to baseline RRi and recovery proportions, which were identified as key drivers of model output variance.
+Validation against physiological benchmarks confirmed the model's robustness and accuracy in real-time assessments, indicating its effectiveness in reflecting autonomic recovery post-exercise.
+**Conclusion**: This study presents a novel non-linear model of RR interval dynamics that provides valuable insights into cardiac autonomic responses during exercise and recovery.
+By enhancing the precision of cardiovascular assessments, the model holds significant promise for supporting personalized health interventions and optimizing performance in both clinical and athletic settings.
 
-**Keywords**: [...].
+**Keywords**: Heart Rate Variability, Exercise Physiology, Autonomic Nervous System, Cardiovascular System,  Models, Theoretical, Logistic Models.
 
@@
 Current research has extensively examined the mechanisms underlying cardiac autonomic dynamics in response to exercise and their links to health-related quality of life and cardiovascular disease risk.
-Understanding these autonomic processes offers valuable insights into optimizing exercise-induced adaptations, with implications for both younger and older individuals.
+Understanding these autonomic processes offers valuable insights into optimizing exercise-induced adaptations, with implications for younger and older individuals.
 
@@
 
-Given the unique characteristics of RRi, their direct relationship with cardiac electrical activity, and responsiveness to autonomic changes, there is a compelling need for a model that accurately represents RRi's non-linear fluctuations during exercise and recovery.
+Given RRi's unique characteristics, direct relationship with cardiac electrical activity, and responsiveness to autonomic changes, a model that accurately represents RRi’s non-linear fluctuations during exercise and recovery is compelling.
 Such a model would offer a more physiologically relevant representation of the heart's behavior compared to the broader HRV indices commonly used in research [@bacopoulou2021entropy].
@@
 This model is designed to capture the complex, real-time changes in RRi, providing meaningful parameters that can enhance our understanding of the physiological processes underlying cardiovascular adaptation to exercise.
-By focusing exclusively on RRi, the model can deliver insights directly applicable to athletic training regimens, recovery protocols, and clinical practices aimed at improving cardiovascular health.
+By focusing exclusively on RRi, the model can deliver insights directly applicable to athletic training regimens, recovery protocols, and clinical practices to improve cardiovascular health.
 The proposed model seeks to bridge the gap between existing modeling frameworks and the physiological reality of RRi dynamics during exercise.
@@
 # Methods
-
-```{r config}
-library(data.table)
-library(ggplot2)
-library(brms)
-```
 
@@
 The first logistic term models the decrease in RRi during exercise, where the parameter $\beta$ denotes the magnitude of this decline.
-The rate of decline is governed by $\lambda$, while $\tau$ represents the onset of the RRi decrease or the time at which the physiological shift begins.
+The rate of decrease is governed by $\lambda$, while $\tau$ represents the onset of the RRi decrease or the time the physiological shift begins.
 
@@
 Sobol analysis was selected for its robustness in handling non-linear and non-monotonic relationships, which are intrinsic to RRi dynamics in response to exercise.
-To compute Sobol indices, a total of 1000 Monte Carlo simulations were conducted, with each simulation involving 1000 randomly sampled parameter sets (1,000,000 model runs in total).
+To compute Sobol indices, 1000 Monte Carlo simulations were conducted, each involving 1000 randomly sampled parameter sets (1,000,000 model runs in total).
 For each set of parameters, RRi were calculated at each time point $t$ across a range from 0 to 20 minutes at intervals of 0.1 minutes (i.e., 6 seconds).
@@
 
+Sobol sensitivity analysis was selected due to its robustness in decomposing the variance attributed to each parameter in complex, non-linear, and non-monotonic models, such as the present logistic model of RRi.
+While methods like the Morris screening technique offer computational efficiency, Sobol's approach provides a more comprehensive assessment of parameter interactions, making it particularly suitable for physiological models where interaction effects may be critical to interpret.
+
 ## Usage of real-world RRi data
@@
 The dataset consisted of 272 participants who underwent a validated exercise protocol encompassing rest, exercise, and recovery phases within a single, continuous measurement session [@castillo2023validity].
-Continuous heart rate data, including RRi, were collected using the Polar Team2 system (Polar®) application, which is capable of capturing dynamic fluctuations associated with varying exercise intensities and recovery.
+Continuous heart rate data, including RRi, were collected using the Polar Team2 system (Polar®) application, capable of capturing dynamic fluctuations associated with varying exercise intensities and recovery.
 
 Preprocessing steps were conducted to remove artifacts and ectopic heartbeats, with less than 3% of data excluded following established guidelines [@malik1996heart].
-The preprocessed RRi data were then aggregated into time intervals to facilitate analysis, allowing the examination of acute exercise responses as well as post-exercise recovery patterns.
+The preprocessed RRi data were then aggregated into time intervals to facilitate analysis, allowing the examination of acute exercise responses and post-exercise recovery patterns.
 
-This real-world dataset provided a critical context for validating the model’s predictive capability against observed physiological responses, offering a robust foundation for understanding RRi dynamics under conditions of physical activity.
+This real-world dataset provided a critical context for validating the model’s predictive capability against observed physiological responses, offering a robust foundation for understanding RRi dynamics under physical activity conditions.
 
@@
 Parameter estimation was performed using Hamiltonian Monte Carlo (HMC) with the No-U-Turn Sampler (NUTS) to explore the parameter space.
-This method is suitable for high-dimensional spaces and utilizes gradient information for efficient sampling.
+This method suits high-dimensional spaces and utilizes gradient information for efficient sampling.
 The parameters $\alpha$, $\beta$, $c$, $\lambda$, $\phi$, $\tau$, and $\delta$ were estimated by sampling from the posterior distribution, which was constructed from observed RRi data and model predictions.
@@
 The gradient of the log-likelihood function for each parameter was computed during estimation using the brms R package, which employs the Stan probabilistic programming language.
-Convergence of the HMC chains was evaluated using the Gelman-Rubin diagnostic and trace plots to ensure reliable parameter estimates.
-This quantification of uncertainty is essential for assessing the stability of parameter estimates, particularly in the presence of noisy RRi data.
+Convergence of the HMC chains was assessed using standard diagnostics, including R-hat values, which were kept below 1.01 for all parameters, and effective sample sizes, which were targeted at a minimum of 1,000 for each parameter.
+Trace plots were inspected to confirm stable mixing, and potential issues with autocorrelation were addressed by increasing the warm-up period to ensure chain independence.
+These diagnostics collectively confirmed reliable posterior estimates for each parameter.
 
-The fitting process utilized five Markov Chain Monte Carlo (MCMC) chains, each consisting of 10,000 iterations with a burn-in period of 5,000 iterations, resulting in a total of 25,000 post-warmup samples.
+The fitting process utilized five Markov Chain Monte Carlo (MCMC) chains, each consisting of 10,000 iterations with a burn-in period of 5,000 iterations, resulting in 25,000 post-warmup samples.
 
-To enhance the exploration of parameter space we performed a two-stage analysis: individual-level and group-level estimates.
+To enhance the exploration of parameter space, we performed a two-stage analysis: individual-level and group-level estimates.
 
@@
 
-Firstly, each subject's RRi data $\text{RRi}_{i,t}$ was standardize against his own mean $\bar{\text{RRi}_i}$ and standard deviation $\sigma_{\text{RRi}_i}$ to improve convergence and exploration of the posterior distribution.
+Firstly, each subject's RRi data $\text{RRi}_{i,t}$ was standardized against his mean $\bar{\text{RRi}_i}$ and standard deviation $\sigma_{\text{RRi}_i}$ to improve convergence and exploration of the posterior distribution.
 The standardized RRi data $y_{i,t}$ for each time point $t$ was computed as:
@@
 [**Figure 1**]{#fig-prior-sim}.
-Prior predicted RRi response to exercise, illustrating uncertainty in model parameters from the prior distribution.
-Shaded areas represent 95%, 80%, and 60% quantile CI.
+Simulated standardized RRi dynamics based on prior parameter distributions, depicting predicted RRi responses to exercise.
+Shaded areas represent 95%, 80%, and 60% quantile credible intervals, offering insight into expected physiological variability across parameters.
 
@@
 
-After obtaining the posterior distribution for each subject's parameters, the mean and standard deviation for each parameter were calculated.
-These estimates were then used as input data to create a multivariate hierarchical model, capturing variability at both the subject and group levels.
+After obtaining the posterior distribution for each subject's parameters, each parameter's mean ($\theta^{\text{ obs}}$) and standard error ($\epsilon$) were calculated.
+These estimates were then used as input data to create a univariate hierarchical model, capturing variability at both the subject and group levels.
 The model is described as follows:
 
-For each subject $i$, the parameter vector $\theta_i = (\alpha_i, \beta_i, c_i, \lambda_i, \phi_i, \tau_i, \delta_i)^\top$ is modeled as:
+For each subject $i$, we estimated an interdependent stochastic process in which the true parameter $\theta_{k,i}$, with $k \in \{\alpha, \beta, c, \lambda, \phi, \tau, \delta\}$ with their corresponding standard error $\epsilon_{k,i}$ was used to model the observed parameter $\hat{\theta}_{k,i}$ as:
 
 $$
-\theta_{i,j} \sim \mathcal{N}(\mu_{j} + b_{i,j}, \sigma_j^2)
+\theta^{\text{ obs}}_{k,i} \sim \mathcal{N}(\theta_{k,i}, \epsilon_{k,i})
 $$
 
-where $\mu_{j}$ is the group-level mean for parameter $j$, $b_{i,j}$ represents the subject-level random effect for subject $i$ on parameter $j$, and $\sigma_j^2$ is the residual variance for parameter $j$.
-
-The subject-level effects $b_{i,j}$ are assumed to follow a multivariate normal distribution:
+Then, the true parameter $\theta_{k,i}$ was further modeled as:
 
 $$
-\mathbf{b}_i = (b_{i, \alpha}, b_{i, \beta}, b_{i, c}, b_{i, \lambda}, b_{i, \phi}, b_{i, \tau}, b_{i, \delta})^\top \sim \mathcal{N}(\mathbf{0}, \Sigma_b)
+\theta_{k,i} \sim \mathcal{N}(\mu_{k} + b_{k,i}, \sigma_k^2)
 $$
 
-where $\Sigma_b$ is the covariance matrix that captures correlations between the random effects for different parameters across subjects.
+where $\mu_{k}$ is the group-level mean for parameter $k$, $b_{k,i}$ represents the subject-level random effect for subject $i$ on parameter $k$ and $\sigma_k^2$ is the residual variance for parameter $k$.
+The subject-level effects $b_{k,i}$ were assumed to be distributed as $b_{k,i} \sim \mathcal{N}(0,\sigma^2)$, with $\sigma$ being the standard error of the subject-level effect.
 
-This hierarchical structure enables us to capture individual variability through subject-level random effects while also estimating group-level effects across all parameters, thus providing estimates into both subject-level and population-level model parameters.
+This hierarchical structure enables us to capture individual variability through subject-level random effects while estimating group-level effects across all parameters, thus providing estimates into subject and population-level model parameters.
 
@@
 
-The model's predictive performance was assessed using approximate leave-one-out (LOO) cross-validation for Bayesian models, utilizing Pareto-smoothed importance sampling.
-The primary evaluation metrics included root mean square error (RMSE) and mean absolute percentage error (MAPE).
-These metrics were selected for their capacity to quantify both absolute and relative errors, thereby providing a comprehensive assessment across the varying scales of RRi dynamics, including resting baseline, peak decline, and recovery.
+The primary performance metrics included R^2^, root mean square error (RMSE) and mean absolute percentage error (MAPE), estimated for each subject.
+Bootstrap resampling across each metric was performed to estimate the mean performance of the model.
+These metrics were selected for their capacity to assess model capacity to predict the observed RRi and to quantify absolute and relative errors, thereby providing a comprehensive assessment across the varying scales of RRi dynamics, including resting baseline, peak decline, and recovery.
 
-Also, residual analysis was conducted to evaluate the model’s accuracy in capturing RRi dynamics.
+Also, residual analysis was conducted to evaluate the model's accuracy in capturing RRi dynamics.
 Residuals were defined as the difference between observed and predicted RRi values.
-These residuals were analyzed for temporal structure and autocorrelation to ensure that no systematic patterns remained in the errors.
+These residuals were analyzed for temporal structure and partial autocorrelation to ensure that no systematic patterns remained in the errors.
 This indicates that the model has sufficiently captured the underlying dynamics of the RRi response to exercise.
@@
 
-According to the proposed model in @eq-main-model, the dynamics of RRi in response to physical exertion can be represented as a linear combination of a baseline RRi $\alpha$ and two logistic functions, denoted as $f_1(t)$ and $f_2(t)$.
+According to the proposed model in @eq-main-model, the dynamics of RRi in response to physical exertion can be represented as a linear combination of a baseline RRi $\alpha$ and two logistic functions denoted as $f_1(t)$ and $f_2(t)$.
 The function $f_1(t)$ models the initial decay in RRi following the initiation of exercise while $f_2(t)$ characterizes the recovery phase after exercise cessation.
@@
 Beyond this point, the denominator approaches 1, allowing the logistic function to attain the asymptotic level determined by the numerator.
-The behavior of the model constituents is illustrated in [Figure 2](#fig-linear-constituents).
+[Figure 2](#fig-linear-constituents) illustrates the behavior of the model constituents.
 
@@
 [**Figure 3**]{#fig-sensitivity}.
-Sensitivity analysis results illustrating the impact of parameter variability on model predictions and the percentage of explained variance accounted for each model parameter.
+Sensitivity analysis results illustrate the impact of parameter variability on model predictions and the percentage of explained variance accounted for each model parameter.
 Shaded areas represent 95%, 80%, and 60% CI estimated from Monte Carlo samples.
-Time in x-axis is truncated at 12 minutes, given that around this time, the proportion of the total variance explained by the respective model parameter stabilizes respective of time.
+Time in the x-axis is truncated at 12 minutes, given that around this time, the proportion of the total variance explained by the respective model parameter stabilizes respective of time.
 
@@
 Therefore, they can significantly affect RRi dynamics over time.
-The influence of these rate parameters on RRi fluctuations over time is depicted in [Figure 4](#fig-data-sim).
+[Figure 4](#fig-data-sim) depicts the influence of these rate parameters on RRi fluctuations over time.
 
@@
 
-Initial exploration of RRi dynamics using two-dimensional density kernel estimation (see [Figure 5](#fig-2d-kernel-density)) indicates a clear drop in RRi around the 5-7 minutes mark, associated with the exercise-induced cardiovascular stress.
+Initial exploration of RRi dynamics using two-dimensional density kernel estimation (see [Figure 5](#fig-2d-kernel-density)) indicates a clear drop in RRi around the 5-7 minutes mark, associated the exercise-induced cardiovascular stress.
 However, greater variability across individuals in post-exercise recovery can be observed.
@@
 Left panel shows the mean and standard deviation (SD) from each of the subject's RRi recordings, used for the standardization process.
-Right panel indicate the 2D kernel density of standardized RRi dynamics over time from a sample of individuals subjected to a rest-exercise-rest protocol.
+Right panel indicates the 2D kernel density of standardized RRi dynamics over time from a sample of individuals subjected to the rest-exercise-rest protocol.
 Darker colors indicate greater probability density.
@@
 
-### Individual-level estimates
+### Model performance
+
+Relative performance metrics, estimated through bootstrapped resampling, suggest that the model tends to deviate a 3.4% (CI~95%~[3.06, 3.81]) from the observed RRi data.
+This is equivalent to a 32.6 ms in the RRi scale (CI~95%~[30.01, 35.77]).
+Residuals analysis showed that the estimated partial correlation function (ACF) from the model residuals indicates a correlation among non-explained errors greater than 0.1 up to the 5th lag.
+However, the partial ACF is significant (CI-wise) and strictly positive or negative up to the second lag.
+Correlations among model residuals against other time indices remained insignificant (see [Figure 6](#fig-error-rates)).
+
+![](figures/error-estimates.png)
+[**Figure 6**]{#fig-error-rates}.
+Individual-level performance metrics.
+Left upper panels indicates bootstrapped MAPE and RMSE, as metrics of relative and absolute model deviance from observed RRi.
+Upper right panel shows the individual-level estimates of model performance and the relationship between them.
+Lower panel indicates the partial autocorrelation function (ACF) of model residuals with corresponding quantile-based CI.
+
+### Group-level estimates
+
+Once subject-level RRi data was fitted using the proposed non-lineaer equation, a population-parameter value was estimated with a generalized hierarchical model with measurement error as part of an interdependent stochastic process using the mean and standard error of the estimated posterior distribution of the subject-level parameter, derived from HMC-NUTS.
+
+The $alpha$ parameter, associated with the baseline RRi before exercise, was estimated at 861.8 ms CI~95%~[850.6, 872.9].
+The $beta$ parameter, which indicates exercise-induced drop in RRi, was estimated at -345.5 ms CI~95%~[-359.8, -331.0].
+The proportion of recovery $c$, which can be expressed a the magnitude relative to $beta$ after exercise cessation, was estimated at 0.84 CI~95%~[0.823, 0.856].
+
+The rate parameters $\lambda$ and $\phi$, associated with the velocity of RRi drop and recovery, were estimated at -3.05 CI~95%~[-3.16, -2.94] and -2.60 CI~95%~[-2.71, -2.48], and after transforming them into a percentage of decay per minute using $1 - exp(\theta)$, we estimated the rate of decay and recovery are 95.2% ms per minute CI~95%~[0.947, 0.957] and 92.5% ms per minute CI~95%~[0.917, 0.933].
+
+The parameters associates with the time at which the initial RRi decay occurs $\tau$ were estimated at 6.71 minutes CI~95%~[6.61, 6.81], meanwhile the time duration associated with the exercise-induced stress in RRi $\delta$ was estimated at 3.24 minutes CI~95%~[3.05, 3.44].
+
+Finally, the standard deviation around the estimated RRi data $\sigma$, associated with the non-modeled noised, was estimated at 27.6 ms CI~95%~[26.5, 28.7].
+In [Figure 7](#fig-group-level-estimates), the model parameter's posterior distribution can be observed.
+
+![](figures/group-level-estimates.png)
+[**Figure 7**]{#fig-group-level-estimates}.
+Posterior probability distributions from group-level estimates of the expectation of each model parameter ($E[\theta]$) with quantile-based 95% CI (upper panel), and the transformed rate parameters into a percentage scale (lower panel).
 
@@
 
-In this study, we present a non-linear model designed to RRi fluctuations during exercise and recovery, providing an in-depth view of cardiac autonomic dynamics.
-By focusing on RRi rather than aggregate HRV metrics, our approach enables real-time tracking of cardiovascular responses to physical stress, which has implications for both clinical and athletic applications.
+This study presents a non-linear model designed for RRi fluctuations during exercise and recovery, providing an in-depth view of cardiac autonomic dynamics.
+By focusing on RRi rather than aggregate HRV metrics, our approach enables real-time tracking of cardiovascular responses to physical stress, which has implications for clinical and athletic applications.
 This model marks a step forward in computational physiology, bridging the gap between theoretical frameworks and practical monitoring applications.
@@
 <!--- Sensitivity Analysis and Model Robustness --->
-The Sobol sensitivity analysis revealed baseline RRi ($\alpha$) and recovery proportion ($c$) as key drivers of model output variance, consistent with physiological expectations.
-Baseline RRi reflects intrinsic cardiac properties, while recovery proportion captures autonomic re-engagement post-exercise, both central to cardiovascular adaptation.
-The relatively low sensitivity of decay and recovery rate parameters ($\lambda$ and $\phi$) suggests the model is robust to moderate fluctuations in these rates, which enhances its reliability for both individualized and group-based applications.
+The Sobol sensitivity analysis revealed baseline RRi ($\alpha$) and recovery proportion ($c$) as key drivers of model output variance that are consistent with physiological expectations.
+Baseline RRi reflects intrinsic cardiac properties, while recovery proportion captures autonomic re-engagement post-exercise, central to cardiovascular adaptation [@hautala2003short].
+The relatively low sensitivity of decay and recovery rate parameters ($\lambda$ and $\phi$) suggests the model is robust to moderate fluctuations in these rates, enhancing its reliability for both individualized and group-based applications.
 
-Sobol analysis was selected for its capacity to handle non-linear, time-varying relationships, but it assumes parameter independence and demands substantial computational resources.
+Compared to previous research, our findings align with the importance of capturing nonlinear dynamics in heart rate variability to understand cardiac responses during exercise [@gronwald2019effects].
+Similarly, previous studies have shown that dynamic fluctuations in RRi can serve as critical indicators of cardiorespiratory fitness, supporting the need for models to address the complexity of cardiovascular responses during physical stress [@mongin2022hrv].
+However, while many existing models focus primarily on linear metrics or aggregate HRV measures, our study provides a high-resolution analysis of RRi dynamics that enhances interpretability and application across diverse fitness levels and exercise intensities.
+
+Sobol analysis was selected for its capacity to handle non-linear, time-varying relationships, but it assumes parameter independence and demands substantial computational resources [@harenberg2017uncertainty; @cheng2017mixed; @herman2013method].
 This assumption may overlook interdependencies typical in biological systems, where parameters such as decay rates and recovery kinetics often interact.
-For future work, alternative sensitivity methods, such as Bayesian sensitivity analysis, could address interdependencies and potentially improve interpretive accuracy in complex biological contexts.
-Additionally, real-time applications in resource-limited environments may benefit from computational optimizations, such as surrogate modeling or sparse-grid approximations, to reduce processing time while maintaining analytical depth.
+For future work, alternative sensitivity methods, such as Bayesian sensitivity analysis, could address interdependencies and improve interpretive accuracy in complex biological contexts.
+Additionally, real-time applications in resource-limited environments may benefit from computational optimizations, such as surrogate modeling or sparse-grid approximations, to reduce processing time while maintaining analytical depth [@bornn2010efficient; @xue2021bayesian].
 
@@
 The practical application of this model extends to clinical monitoring and athletic training, but further validation is essential to confirm its robustness in real-world conditions.
-To support clinical decision-making, future studies should validate the model against established physiological benchmarks, such as VO₂ max and lactate thresholds, which would facilitate its integration into routine health assessments.
+To support clinical decision-making, future studies should validate the model against established physiological benchmarks, such as VO~2~ max and lactate thresholds, facilitating its integration into routine health assessments.
 For instance, this model could be valuable in cardiovascular rehabilitation programs, where tracking real-time autonomic responses can help personalize exercise regimens, ensuring adequate recovery without overstressing the cardiovascular system.
-Furthermore, for at-risk populations, the model could assist in detecting autonomic irregularities indicative of early cardiovascular dysfunction.
+Furthermore, for at-risk populations, the model could assist in detecting autonomic irregularities indicative of early cardiovascular dysfunction [@thayer2010relationship].
 
-For athletic applications, the model has potential in guiding interval training, where it could help identify optimal recovery points between intense exercise bouts.
-In this setting, real-time tracking of RRi dynamics may help prevent overtraining and optimize performance by aligning training load with individual recovery capacity.
+For athletic applications, the model has the potential to guide interval training, where it could help identify optimal recovery points between intense exercise bouts.
+This is particularly relevant given the findings of other research, which suggest that precise monitoring of heart rate dynamics can prevent overtraining and optimize performance [@eser2022acute].
 However, extending this model beyond controlled, single-bout settings requires further validation, particularly in dynamic environments where external factors like temperature, altitude, and psychological stress can influence RRi.
-Integrating the model with wearable devices would be essential for field applications, but these devices must capture RRi with sufficient accuracy to maintain model precision.
-Future studies should focus on adapting the model for wearable technology and testing it in diverse settings to evaluate its reliability across environmental and situational variations.
+Integrating the model with wearable devices capable of accurately capturing RRi is essential for field applications, enabling real-time monitoring and feedback [@lundstrom2023practices].
 
@@
 Traditional models of RRi often rely on linear assumptions or simple exponential functions, which fail to capture the transient, non-stationary nature of RRi during exercise and recovery.
-By using logistic functions to model both the decay and recovery phases, our non-linear approach accommodates gradual autonomic shifts more effectively than exponential functions, offering a physiologically relevant representation of RRi dynamics.
-This model also enhances interpretability by isolating time-dependent contributions of each parameter, as demonstrated in the Sobol analysis, thereby enabling applications across varied fitness levels and exercise intensities.
-
-While recent studies have developed non-linear models, many focus on HRV aggregates that summarize rather than directly model RRi changes.
-By focusing on RRi, our model provides a high-resolution view of cardiac adaptation, which is particularly valuable for both clinical and athletic settings.
-Furthermore, the ability to adapt the model parameters to real-time changes in RRi has the potential to enhance sensitivity in detecting subtle shifts in autonomic state, which aggregated HRV indices may obscure.
+Using logistic functions to model the decay and recovery phases, our non-linear approach accommodates gradual autonomic shifts more effectively than exponential functions, offering a physiologically relevant representation of RRi dynamics.
+While recent studies have developed non-linear models, many focus on HRV aggregates that summarize rather than directly model RRi changes [@molkkari2020online].
+By concentrating on RRi, our approach offers a high-resolution view of cardiac adaptation, making it particularly valuable for clinical and athletic contexts.
+Adapting model parameters to real-time changes in RRi enhances sensitivity in detecting subtle shifts in autonomic state, which aggregated HRV indices may obscure [@mabe2024associations].
 
@@
 First, the uniform parameter sampling used in sensitivity analysis, though practical, may not fully capture individual variability in populations with extreme autonomic profiles or chronic health conditions.
-Moving forward, incorporating empirical distributions or Bayesian priors based on population data could refine parameter estimates, enhancing individualization for diverse clinical populations, such as older adults or individuals with cardiovascular disease.
+Incorporating empirical distributions or Bayesian priors based on population data could refine parameter estimates, enhancing individualization for diverse clinical populations, such as older adults or individuals with cardiovascular disease [@kristal1995heart].
 
-Another area for future research is multi-session modeling to account for cumulative adaptations over repeated exercise sessions.
-Real-world training and rehabilitation often involve multiple bouts of exercise that can introduce cumulative autonomic fatigue or adaptation.
-Extending the model to track such longitudinal effects could help identify early signs of overtraining or autonomic exhaustion, thereby providing a preventive tool for both clinical and athletic populations.
-Additionally, considering environmental and psychological factors—such as temperature, stress, or sleep quality—would add robustness, ensuring that model predictions remain accurate across varied real-world contexts.
+Moreover, an area for future research includes multi-session modeling to account for cumulative adaptations over repeated exercise sessions.
+Studies have shown that cumulative effects from multiple exercise bouts can introduce autonomic fatigue or adaptation [@castillo2021cardiac].
+Extending the model to track such longitudinal effects could help identify early signs of overtraining or autonomic exhaustion, providing a preventive tool for clinical and athletic populations.
+Additionally, considering environmental and psychological factors (i.e., such as temperature, stress, or sleep quality) would add robustness, ensuring that model predictions remain accurate across varied real-world contexts.
 
@@
 
-In summary, this study introduces a non-linear model of RRi dynamics that captures complex, real-time fluctuations reflecting the autonomic nervous system’s response to exercise and recovery.
-By focusing on RRi, the model offers enhanced sensitivity to physiological changes, making it well-suited for personalized applications in clinical and athletic settings.
-The use of Sobol sensitivity analysis enabled a time-dependent interpretation of parameter effects, underscoring the model’s adaptability to diverse physiological states and exercise intensities.
-Nevertheless, further validation, particularly with multi-session and wearable-based applications, is essential to establish the model’s robustness and applicability in dynamic, real-world environments.
-By advancing precision in cardiovascular monitoring, this model holds promise for supporting individualized health and performance interventions, contributing meaningfully to the fields of computational physiology and applied health sciences.
+This study presents a novel non-linear model for characterizing RRi dynamics during exercise and recovery, providing a more accurate and nuanced understanding of cardiac autonomic function compared to traditional HRV metrics.
+Our findings emphasize the important roles of baseline RRi and recovery proportion in reflecting autonomic re-engagement, enhancing the sensitivity of real-time assessments of cardiovascular responses to physical stress.
+This model advances computational physiology and holds significant promise for improving cardiovascular health monitoring and performance optimization.
 
@@
 
-This research did not receive any funding.
+This work was funded by ANID Proyecto Fondecyt Iniciación Nº11220116.
 

### 2024-11-11 R/0-test-sim-data.R

@@ -1,162 +1,0 @@
-
-# Prepare workspace -------------------------------------------------------
-
-## Load libraries
-library(data.table)
-library(ggplot2)
-library(brms)
-
-## Main equation
-RRi_model <- function(t, alpha, beta, c, lambda, phi, tau, delta) {
-  alpha +
-    (beta / (1 + exp(lambda * (t - tau)))) +
-    ((-beta * c) / (1 + exp(phi * (t - tau - delta))))
-}
-
-simulate_data <- function(n_id, n_time, noise = 10, seed = 1234) {
-  set.seed(seed)
-  within(
-    data = data.table::data.table(), 
-    expr = {
-      alpha = rep(runif(n_id, 800, 1000), each = n_time)
-      beta = rep(runif(n_id, -500, -300), each = n_time)
-      c = rep(runif(n_id, .7, 1.1), each = n_time)
-      lambda = rep(runif(n_id, log(1-.9), log(1-.7)), each = n_time)
-      phi = rep(runif(n_id, log(1-.7), log(1-.5)), each = n_time)
-      tau = rep(runif(n_id, 3, 7), each = n_time)
-      delta = rep(runif(n_id, 1, 3), each = n_time)
-      time = rep(x = seq(0, 20, length.out = n_time), times = n_id)
-      rri_true = RRi_model(time, alpha, beta, c, lambda, phi, tau, delta)
-      rri_hat = rri_true + brms::rstudent_t(n_id * n_time, df = 5, sigma = noise)
-      id = rep(1:n_id, each = n_time)
-    }
-  )
-}
-
-sample_data <- function(data, n, id_var, na.rm = FALSE) {
-  na_fun <- function(x) return(x)
-  if (isTRUE(na.rm)) na_fun <- na.omit
-  na_fun(data)[, .SD[sort(sample.int(.N, n))], by = id_var]
-}
-
-## Load the data
-data("rri_data")
-
-## Standardize as with the simulated dataset
-rri_data[, rri_mean := mean(rr_denoised, na.rm = TRUE), by = id]
-rri_data[, rri_sd := sd(rr_denoised, na.rm = TRUE), by = id]
-rri_data[, rri_std := (rr_denoised - rri_mean) / rri_sd, by = id]
-
-m_data <- sample_data(rri_data, 200, "id", na.rm = TRUE)
-
-## Check the 2d-density kernel of time-rri_std
[…]

### 2024-11-11 R/1-prior-rri-model.R

@@ -19,11 +19,10 @@ RRi_model <- function(t, alpha, beta, c, lambda, phi, tau, delta) {
 ## Plot posterior density of a chosen parameter
 plot_post_dens <- function(posterior, x, xlab) {
   ggplot(posterior, aes(x = {{x}})) +
-    geom_histogram(aes(y = after_stat(density)), 
-                   col = "darkblue", fill = "lightblue", bins = 20) +
-    stat_density(outline.type = "both", fill = NA, 
-                 col = "gray20", adjust = 1.5,) +
-    scale_y_continuous(expand = c(0,0,0.05,0), breaks = NULL, name = NULL) +
+    ggdist::stat_slabinterval(
+      fill = "lightblue",
+      trim = FALSE, density = "unbounded") +
+    scale_y_continuous(expand = c(0.1,0,0.05,0), breaks = NULL, name = NULL) +
     scale_x_continuous(expand = c(0,0)) +
     labs(x = xlab) +
     theme_classic(base_size = 12) +

### 2024-11-11 R/5-2d-density-kernel.R

@@ -30,19 +30,15 @@ figure_b <-
   }) |> 
   ggplot(aes(value)) +
   facet_wrap(~ variable, nrow = 2, scales = "free_x") +
-  geom_histogram(aes(y = after_stat(ndensity), 
-                     fill = variable), 
-                 bins = 15, show.legend = FALSE,
-                 col = "gray10") +
-  stat_density(geom = "line",
-               aes(y = after_stat(ndensity)),
-               adjust = 4, 
-               col = "gray10") +
-  geom_boxplot(aes(y = -.05, fill = variable),
-               width = .05, orientation = "y", staplewidth = 0,
-               show.legend = FALSE, outlier.color = NA) +
-  scale_fill_brewer() +
-  scale_y_continuous(breaks = NULL, name = NULL, expand = c(0.025,0,0.1,0)) +
+  ggdist::stat_slabinterval(
+    aes(fill = variable),
+    normalize = "panels",
+    adjust = 20, trim = FALSE,
+    density = "unbounded",
+    show.legend = FALSE
+  ) +
+  ggsci::scale_fill_observable() +
+  scale_y_continuous(breaks = NULL, name = NULL, expand = c(0.1,0,0.1,0)) +
   scale_x_continuous(name = "RRi (ms)", n.breaks = 4) +
   theme_classic(base_size = 14) +
   theme(axis.text = element_text(size = rel(0.8)),
@@ -55,6 +51,6 @@ figure <- ggpubr::ggarrange(
 )
 
 ggsave("figures/2d-density-kernel.pdf", figure, "pdf", 
-       width = 7, height = 5, units = "in")
+       width = 7, height = 4, units = "in")
 ggsave("figures/2d-density-kernel.png", figure, "png", 
-       width = 7, height = 5, units = "in", dpi = 500)
+       width = 7, height = 4, units = "in", dpi = 500)

### 2024-11-11 R/6-fit-model-to-data.R

@@ -9,6 +9,13 @@ library(data.table)
 data("rri_data")
 m_data <- rri_data[!is.na(rri_std)]
 
+## Add custom function
+RRi_sim <- function(t, alpha, beta, c, lambda, phi, tau, delta) {
+  alpha +
+    (beta / (1 + exp(lambda * (t - tau)))) +
+    ((-beta * c) / (1 + exp(phi * (t - tau - delta))))
+}
+
 if (!file.exists("models/stage_1_model.RDS")) {
 
   # Load model --------------------------------------------------------------
@@ -64,6 +71,7 @@ df_models <- merge.data.table(
 
 df_models[, b_alpha_Intercept := b_alpha_Intercept * rri_sd + rri_mean, id]
 df_models[, b_beta_Intercept := b_beta_Intercept * rri_sd, id]
+df_models[, sigma := sigma * rri_sd, id]
 
 ## Compute mean and SD for each parameter, for each ID
 mu_data <- df_models[, lapply(.SD, mean), id, .SDcols = b_alpha_Intercept:sigma]
@@ -81,3 +89,28 @@ id_data <- merge.data.table(mu_data, se_data, by = "id")
 
 ## Save resulting data.frame in CSV file
 fwrite(id_data, file = "models/id_level_posterior.csv")
+
+## Generate data with estimated parameters and observed data
+long_data <- merge.data.table(
+  m_data[, .(id, time, rr_denoised)],
+  id_data[, .SD, .SDcols = id:deltaMu],
+  by = "id",
+)
+
+## Add estimated RRi data for each id
+long_data[, pred_rr := RRi_sim(
+  t = time, alpha = alphaMu,
+  beta = betaMu, c = cMu,
+  lambda = lambdaMu, phi = phiMu,
+  tau = tauMu, delta = deltaMu
+), id]
+
+## Compute error metrics
+long_data[, error := rr_denoised - pred_rr, id]
+long_data[, rr_bar := mean(rr_denoised, na.rm = TRUE), id]
+long_data[, mape := mean(abs(error/rr_denoised), na.rm = T), id]
+long_data[, rmse := sqrt(mean((error^2), na.rm = TRUE)), id]
+long_data[, rsquared := 1-sum(error^2)/sum((rr_denoised-rr_bar)^2), id]
+
+## Save the computed metrics
+fwrite(long_data, file = "models/error_performance.csv")

### 2024-11-11 R/7-meta-analysis.R

@@ -15,28 +15,19 @@ id_data <- fread(file = "models/id_level_posterior.csv")
 
 ## Custom functions
 fit_model <- function(model, priors) {
-  file = paste0("models/group_level/",
-                deparse(substitute(model)),
-                ".RDS")
-  
-  if (file.exists(file)) {
-    m_fit <- readRDS(file)
-  } else {
-    m_fit <- brms::brm(formula = model,
-                       data = id_data,
-                       prior = priors,
-                       family = gaussian(),
-                       chains = 5, cores = 5, seed = 1234,
-                       iter = 10000, warmup = 5000,
-                       control = list(adapt_delta = .99,
-                                      max_treedepth = 50))
-    
-    saveRDS(object = m_fit, file = file, compress = "xz")
-  }
-  
-  return(m_fit)
+  brms::brm(formula = model,
+            data = id_data,
+            prior = priors,
+            family = gaussian(),
+            chains = 5, cores = 5, seed = 1234,
+            iter = 10000, warmup = 5000,
+            control = list(adapt_delta = .99,
+                           max_treedepth = 50),
+            file = paste0("models/group_level/",
+                          deparse(substitute(model)),
+                          ".RDS")
+  )
 }
-
 
 # Fit model ---------------------------------------------------------------
 
@@ -80,9 +71,9 @@ m_delta_prior <-
   prior(normal(1.5, 0.1), class = "sigma", lb = 0) +
   prior(normal(0.5, 0.5), class = "sd", lb = 0)
 m_sigma_prior <-
-  prior(normal(0.3, 0.05), class = "Intercept", lb = 0) +
-  prior(normal(0.05, 0.01), class = "sigma", lb = 0) +
-  prior(normal(0.05, 0.01), class = "sd", lb = 0)
+  prior(normal(30, 5), class = "Intercept", lb = 0) +
+  prior(normal(10, 1), class = "sigma", lb = 0) +
+  prior(normal(5, 1), class = "sd", lb = 0)
 
 m_alpha_fit <- fit_model(m_alpha_model, m_alpha_prior)
 m_beta_fit <- fit_model(m_beta_model, m_beta_prior)
@@ -92,3 +83,20 @@ m_phi_fit <- fit_model(m_phi_model, m_phi_prior)
[…]

### 2024-11-11 R/8-error-estimates.R

@@ -1,0 +1,115 @@
+
+# Prepare workspace -------------------------------------------------------
+
+## Load libraries ----
+library(data.table)
+library(ggplot2)
+library(ggdist)
+
+## Load data ----
+error_data_long <- fread(file = "models/error_performance.csv")
+
+## Prepare data ----
+
+### Unique MAPE and RMSE for each id
+error_data <- error_data_long[, lapply(.SD, unique), id, .SDcols = c(14:16)]
+
+### PACF for each id
+acf_data <- error_data_long[
+  j = list(acf = as.numeric(pacf(error, lag.max = 30, plot = F)$acf),
+           lag = 1:30), 
+  by = id]
+
+### Bootstrap MAPE and RMSE for better estimates
+set.seed(1234)
+boot_error <- error_data[, lapply(.SD, function(x) {
+  boot = 50000
+  x_len = length(x)
+  vapply(seq_len(boot), function(i) {
+    mean(x[sample.int(x_len, x_len, replace = TRUE)])
+  }, FUN.VALUE = 1)
+}), .SDcols = mape:rsquared]
+
+boot_error[, mc_sample := seq_len(.N)]
+
+# -------------------------------------------------------------------------
+
+f_1 <- ggplot(error_data, aes(mape, rmse)) +
+  geom_point(col = "gray20", alpha = .5) +
+  scale_y_continuous(name = "RMSE (log-scale)", expand = c(0,0.05),
+                     breaks = c(1,2,4,9,20)*10,
+                     transform = "log10") +
+  scale_x_continuous(labels = scales::label_percent(),
+                     breaks = c(1,2,4,9,20)/100,
+                     name = "MAPE (log-scale)", expand = c(0,0.05),
+                     transform = "log10") +
+  theme_classic(base_size = 12) +
+  theme(axis.text = element_text(size = rel(.8)))
+
+error_long <- melt.data.table(boot_error, id.vars = "mc_sample")
+error_long[, variable := `levels<-`(variable, c("MAPE", "RMSE", "R^2"))]
+
+f_2a <- ggplot(error_long[variable == "MAPE"], aes(variable, value)) +
+  facet_wrap(~ variable, scales = "free") +
+  ggdist::stat_halfeye(fill = "#4269D0FF") +
+  scale_y_continuous(name = NULL, expand = c(0.1,0),
[…]

### 2024-11-11 R/9-group-level-estimates.R

@@ -1,0 +1,127 @@
+
+# Prepare workspace -------------------------------------------------------
+
+## Load libraries
+library(data.table)
+library(ggplot2)
+library(ggdist)
+
+## Load data
+m_fits_df <- fread(file = "models/group_level_posterior.csv")
+m_fits_df[, Parameter := factor(Parameter, 
+                                levels = c("alpha", "beta", "c", "lambda", 
+                                           "phi", "delta", "tau", "sigma"))]
+
+
+
+# Plot dens only ----------------------------------------------------------
+
+f_1 <- ggplot(m_fits_df, aes(Parameter, b_Intercept)) +
+  facet_wrap(~ Parameter, nrow = 1, ncol = 8, scales = "free",
+             labeller = label_parsed) +
+  ggdist::stat_halfeye(aes(fill = Parameter),
+                       show.legend = FALSE,
+                       normalize = "panels",
+                       trim = FALSE, density = "unbounded") +
+  scale_x_discrete(expand = c(0.2,0,0.2,0), 
+                   name = NULL, breaks = NULL) +
+  scale_y_continuous(name = expression(symbol(E)~"["*theta*"]")) +
+  ggsci::scale_fill_observable() +
+  theme_classic(base_size = 12) +
+  theme(strip.background = element_rect(fill = "grey90", colour = NA),
+        axis.text = element_text(size = rel(.8)))
+
+f_2 <- 
+  m_fits_df[Parameter %in% c("lambda", "phi"), 1 - exp(b_Intercept), Parameter] |> 
+  ggplot(aes(V1, Parameter)) +
+  facet_wrap(~ Parameter, nrow = 1, ncol = 2, scales = "free",
+             labeller = label_parsed) +
+  ggdist::stat_halfeye(aes(fill = Parameter),
+                       show.legend = FALSE,
+                       normalize = "panels",
+                       trim = FALSE, density = "unbounded") +
+  scale_y_discrete(expand = c(0.15,0,0.2,0), 
+                   name = NULL, breaks = NULL) +
+  scale_x_continuous(name = "Rate of change per minute (%)",
+                     labels = scales::label_percent(),
+                     n.breaks = 4,
+                     expand = c(0.2,0)) +
+  ggsci::scale_fill_observable() +
+  theme_classic(base_size = 12) +
[…]

### 2024-11-11 R/_test-sim-data.R

@@ -1,0 +1,162 @@
+
+# Prepare workspace -------------------------------------------------------
+
+## Load libraries
+library(data.table)
+library(ggplot2)
+library(brms)
+
+## Main equation
+RRi_model <- function(t, alpha, beta, c, lambda, phi, tau, delta) {
+  alpha +
+    (beta / (1 + exp(lambda * (t - tau)))) +
+    ((-beta * c) / (1 + exp(phi * (t - tau - delta))))
+}
+
+simulate_data <- function(n_id, n_time, noise = 10, seed = 1234) {
+  set.seed(seed)
+  within(
+    data = data.table::data.table(), 
+    expr = {
+      alpha = rep(runif(n_id, 800, 1000), each = n_time)
+      beta = rep(runif(n_id, -500, -300), each = n_time)
+      c = rep(runif(n_id, .7, 1.1), each = n_time)
+      lambda = rep(runif(n_id, log(1-.9), log(1-.7)), each = n_time)
+      phi = rep(runif(n_id, log(1-.7), log(1-.5)), each = n_time)
+      tau = rep(runif(n_id, 3, 7), each = n_time)
+      delta = rep(runif(n_id, 1, 3), each = n_time)
+      time = rep(x = seq(0, 20, length.out = n_time), times = n_id)
+      rri_true = RRi_model(time, alpha, beta, c, lambda, phi, tau, delta)
+      rri_hat = rri_true + brms::rstudent_t(n_id * n_time, df = 5, sigma = noise)
+      id = rep(1:n_id, each = n_time)
+    }
+  )
+}
+
+sample_data <- function(data, n, id_var, na.rm = FALSE) {
+  na_fun <- function(x) return(x)
+  if (isTRUE(na.rm)) na_fun <- na.omit
+  na_fun(data)[, .SD[sort(sample.int(.N, n))], by = id_var]
+}
+
+## Load the data
+data("rri_data")
+
+## Standardize as with the simulated dataset
+rri_data[, rri_mean := mean(rr_denoised, na.rm = TRUE), by = id]
+rri_data[, rri_sd := sd(rr_denoised, na.rm = TRUE), by = id]
+rri_data[, rri_std := (rr_denoised - rri_mean) / rri_sd, by = id]
+
+m_data <- sample_data(rri_data, 200, "id", na.rm = TRUE)
+
+## Check the 2d-density kernel of time-rri_std
[…]

### 2024-12-09 paper.qmd
preliminary version of paper is ready

@@
 bibliography: "misc/ref.bib"
-csl: "misc/frontiers-in-physiology.csl"
+csl: "misc/scientific-reports.csl"
 editor_options: 
@@
 
-**Authors**: : Matías Castillo-Aguilar^1,2^, Diego Mabe-Castro^1,2^, David Medina^1^, Cristian Núñez-Espinosa^1,3^\*.
+**Authors**: : Matías Castillo-Aguilar^1,2^, Diego Mabe-Castro^1,2^, David Medina^4,5^, Cristian Núñez-Espinosa^1,3^\*.
 
@@
 ^3^ School of Medicine, University of Magallanes, Punta Arenas, Chile.
+
+^4^ Departamento de Ingeniería en Computación, Universidad de Magallanes, Punta Arenas, Chile.
+
+^5^ Centre for Biotechnology and Bioengineering, CeBiB, Universidad de Chile, Santiago, Chile.
+
 :::
@@
 
-**Objective**: This study aims to develop and validate a novel non-linear model for characterizing RR interval (RRi) dynamics during exercise and recovery, enhancing the understanding of the cardiac autonomic function and its implications for both clinical practice and athletic training.
-**Methods**: A cohort of 272 participants underwent a validated exercise protocol, during which continuous heart rate data, including RR intervals, were collected.
-A non-linear logistic approach was employed to model the fluctuations in RRi.
-Sobol sensitivity analysis was conducted to assess the influence of model parameters on RR dynamics.
-**Results**: The proposed model successfully captured the complex fluctuations in RRi, demonstrating significant sensitivity to baseline RRi and recovery proportions, which were identified as key drivers of model output variance.
-Validation against physiological benchmarks confirmed the model's robustness and accuracy in real-time assessments, indicating its effectiveness in reflecting autonomic recovery post-exercise.
-**Conclusion**: This study presents a novel non-linear model of RR interval dynamics that provides valuable insights into cardiac autonomic responses during exercise and recovery.
-By enhancing the precision of cardiovascular assessments, the model holds significant promise for supporting personalized health interventions and optimizing performance in both clinical and athletic settings.
+The aim of this work was to develop and validate a novel non-linear model to characterize RR interval (RRi) dynamics throughout a rest-exercise-recovery protocol, offering a more precise and physiologically relevant representation of cardiac autonomic responses than traditional HRV metrics or linear approaches.
+Using data from a cohort of 272 elderly participants, the model employs logistic functions to capture the non-stationary and transient nature of RRi dynamics, with parameter estimation achieved via Hamiltonian Monte Carlo.
+Sobol sensitivity analysis identified baseline RRi ($\alpha$) and recovery proportion ($c$) as the primary drivers of variability, underscoring their critical roles in autonomic regulation and resilience.
+Validation against real-world RRi data demonstrated robust model performance, accurately reflecting autonomic recovery and exercise-induced fluctuations.
+By advancing real-time cardiovascular assessments, this framework holds significant potential for clinical applications in rehabilitation and cardiovascular monitoring, as well as athletic contexts for optimizing performance and recovery.
+Taken together, these findings highlight the model's ability to provide precise, physiologically relevant assessments of autonomic function, paving the way for its use in personalized health monitoring and performance optimization across diverse populations.
 
@@
 
-Current research has extensively examined the mechanisms underlying cardiac autonomic dynamics in response to exercise and their links to health-related quality of life and cardiovascular disease risk.
-Understanding these autonomic processes offers valuable insights into optimizing exercise-induced adaptations, with implications for younger and older individuals.
+Current research has extensively examined the mechanisms underlying cardiac autonomic dynamics in response to exercise and their links to health-related quality of life and cardiovascular disease risk [@eser2022exercise;@castillo2023validity;@mabe2024associations].
 
-In this context, the study of R-R intervals (RRi) in response to exercise has emerged as an important research area, given its relevance to cardiovascular health, athletic performance, and physiological adaptation [@kristal1995heart; @thayer2010relationship; @dong2016role; @lundstrom2023practices].
+In this context, the study of R-R intervals (RRi), defined as the time intervals between heartbeats, and its link with exercise has emerged as an important research area, given its relevance to cardiovascular health, athletic performance, and physiological adaptation [@kristal1995heart; @thayer2010relationship; @dong2016role; @lundstrom2023practices].
+Hence, being able to analyze the temporal dynamics of RRi in response to exercise can provide valuable insights into the mechanisms by which the cardiovascular system adapts to physical stressors, such as exercise-induced fatigue and competition-related strain [@castillo2021cardiac; @eser2022exercise].
+
 Unlike heart rate variability (HRV), which aggregates autonomic responses over time, RRi analysis provides a more granular, direct view of cardiac electrical activity during or immediately following exercise, particularly in older adults [@mongin2022hrv; @castillo2023validity; @mabe2024associations].
-Analyzing the temporal dynamics of RRi (i.e., the time between successive heartbeats) provides invaluable insights into how the cardiovascular system behaves to and recovers from physical stressors such as exercise-induced fatigue and competition-related strain [@castillo2021cardiac; @eser2022exercise].
 
-Understanding these fluctuations is particularly relevant during dynamic exercise periods, where the autonomic nervous system (ANS) shifts between parasympathetic withdrawal and sympathetic activation [@boettger2010heart].
-Modeling RRi dynamics, rather than relying on broader HRV metrics, allows for a direct assessment of physiological markers of autonomic adaptation to stress [@hautala2003short].
+Thus, understanding RRi fluctuations in response to exercise is particularly relevant during dynamic exercise periods, where the autonomic nervous system (ANS) shifts between parasympathetic withdrawal and sympathetic activation [@boettger2010heart].
+Therefore, directly modeling RRi dynamics, rather than relying on broader HRV metrics, allows for a direct assessment of physiological markers of autonomic adaptation to stress [@hautala2003short].
 This approach is valuable for identifying recovery patterns and understanding cardiovascular reactivity across individuals with various fitness levels [@mongin2022hrv].
 
-Despite its importance, modeling RRi behavior during and after exercise in a continuous measurement poses significant challenges.
-Traditional approaches, such as linear regression and time-series analysis, often fail to capture the intricate transitions in RRi, especially under intense exertion and recovery phases.
-This difficulty arises due to the inherently non-linear and non-stationary nature of HRV [@gronwald2019effects].
-While linear models oversimplify these dynamics, advanced non-linear approaches have been developed to address the limitations of linear analysis.
-However, many focus on HRV summaries rather than direct RRi modeling [@gronwald2019non].
+Modeling the RRi behavior has been traditionally approached by leveraging linear regression and time-series analysis [@lian2011simple].
+However, this oppose significant challenges, like oversimplifying overall exercise dynamics without capturing the intricacies of exercise-induced intricate transitions in RRi, especially under intense exertion and recovery phases [@gronwald2019effects].
+More recently, advanced non-linear approaches have been developed to address the limitations of linear methods like decision tree-based ensemble algorithms and convolutional neural networks [@wang2020automated;@lee2023real;@berrahou2024arrhythmia].
+However, many of the more advance alternatives fails to generate physiologically meaningful model parameters, without a direct link to biological processes [@gronwald2019non].
 
-Recent studies have begun exploring non-linear models for RRi dynamics, recognizing their potential to capture the complexity of cardiovascular response to exercise.
-Exponential decay models, for example, have been proposed to describe RRi recovery, while logistic functions have been used to model the gradual return to baseline after high-intensity exercise [@molkkari2020online; @gronwald2019non].
+Alongside this line of inquiry, recent studies have begun exploring non-linear models for RRi dynamics, recognizing their potential to capture the complexity of cardiovascular response to exercise [@da2024post].
+Exponential decay models, for example, have been proposed to describe RRi recovery [@sahoo2024unanticipated], while logistic functions have been used to model the gradual return to baseline after high-intensity exercise [@molkkari2020online; @gronwald2019non].
 These models offer advantages over traditional HRV metrics by providing a more detailed understanding of the cardiovascular system's response to exercise [@wu2003nonlinear].
-However, despite these advancements, few models are specifically designed to capture real-time RRi fluctuations, and even fewer consider individual variability factors such as fitness level, autonomic balance, and exercise intensity [@kanniainen2023estimation].
+However, despite these advancements, few models are specifically designed to capture real-time RRi fluctuations, and even fewer consider physiologically meaningful parameters that allow modeling individual variability across a myriad of exercise dimensions [@kanniainen2023estimation].
 
-Given RRi's unique characteristics, direct relationship with cardiac electrical activity, and responsiveness to autonomic changes, a model that accurately represents RRi’s non-linear fluctuations during exercise and recovery is compelling.
-Such a model would offer a more physiologically relevant representation of the heart's behavior compared to the broader HRV indices commonly used in research [@bacopoulou2021entropy].
+Considering the current state of the art, there is a clear need to develop a non-linear model that accurately represents RRi's non-linear transitions during exercise and recovery, where the estimation of clinically significant parameters is compelling.
+Such a model would need to offer a more physiologically relevant representation of the heart's behavior compared to the broader HRV indices commonly used in research [@bacopoulou2021entropy].
+This model would need to be complex enough to capture the non-linear, exercise-induced, cardiovascular dynamics.
+But, at the same time, simple enough to be able to provide practical and significant model parameters related to observed physiological processes.
 
-Hence, the primary objective of this paper is to present a novel non-linear model that characterizes RRi fluctuations during exercise and recovery.
-This model is designed to capture the complex, real-time changes in RRi, providing meaningful parameters that can enhance our understanding of the physiological processes underlying cardiovascular adaptation to exercise.
-By focusing exclusively on RRi, the model can deliver insights directly applicable to athletic training regimens, recovery protocols, and clinical practices to improve cardiovascular health.
-The proposed model seeks to bridge the gap between existing modeling frameworks and the physiological reality of RRi dynamics during exercise.
+Hence, the primary objective of this paper is to present a novel non-linear model that characterizes continuous RRi transitions from rest to exercise and recovery.
+This model is designed to capture the non-linear changes in RRi, providing meaningful parameters that can enhance our understanding of the physiological processes underlying cardiovascular adaptation to exercise.
 
@@
 
-## Model Specification
+## Data collection and preprocessing
 
-The mathematical model proposed to characterize the RRi response to exercise and recovery is defined by @eq-main-model.
-
-$$
-\begin{aligned}
-\text{RRi}(t) = \alpha + \frac{\beta}{1 + e^{\lambda(t - \tau)}} + \frac{-\beta \cdot c}{1 + e^{\phi(t - \tau - \delta)}}
-\end{aligned}
-$${#eq-main-model}
-
-This model includes two logistic functions representing the RRi dynamics across exercise and recovery phases.
-The first logistic term models the decrease in RRi during exercise, where the parameter $\beta$ denotes the magnitude of this decline.
-The rate of decrease is governed by $\lambda$, while $\tau$ represents the onset of the RRi decrease or the time the physiological shift begins.
-
-The second logistic term accounts for RRi recovery post-exercise.
-Here, $c$ scales the magnitude of recovery relative to the initial decline represented by $\beta$, effectively capturing the proportion of the decline regained during recovery.
-The rate at which RRi returns to baseline is controlled by $\phi$, and $\delta$ indicates the lag following the cessation of exercise, marking the beginning of recovery.
-
-This logistic structure is well-suited to modeling the RRi dynamics, providing a smooth, continuous transition for the decline and recovery phases.
-Logistic growth functions are particularly effective in physiological modeling contexts, where transitions (e.g., rest to exercise or exercise to recovery) occur gradually and non-linearly.
-Compared to conditional models, which may introduce abrupt transitions, this model is designed to minimize discontinuities, thus offering a realistic representation of RRi responses without artifacts.
-
-## Sensitivity to model parameters
-
-To assess the sensitivity of model parameters in influencing RRi over time, we implemented a Sobol sensitivity analysis using Monte Carlo simulations.
-Sobol analysis was selected for its robustness in handling non-linear and non-monotonic relationships, which are intrinsic to RRi dynamics in response to exercise.
-To compute Sobol indices, 1000 Monte Carlo simulations were conducted, each involving 1000 randomly sampled parameter sets (1,000,000 model runs in total).
-For each set of parameters, RRi were calculated at each time point $t$ across a range from 0 to 20 minutes at intervals of 0.1 minutes (i.e., 6 seconds).
-The resulting Sobol indices provided a measure of the contribution of each parameter to the variance in RRi at each time point.
-
-The first-order Sobol index for each parameter was computed by isolating the variance attributable to each parameter while averaging over the others.
-To achieve this, we perturbed each parameter individually while holding all other parameters at their average values across the samples.
-The proportion of variance explained by each parameter at each time point was calculated by dividing the variance of the estimated RRi of the perturbed parameter by the total variance, which accounts for the variation in all model parameters simultaneously, yielding time-dependent Sobol indices.
-The selected parameter ranges, provided in [Table 1](#tbl-sens-params), reflect a 50% variation in both directions from a reference value for each parameter.
-
-|  | $\alpha$| $\beta$| $c$| $\lambda$| $\phi$| $\tau$| $\delta$|
-|----|--------:|-------:|----:|-------:|-------:|------:|--------:|
-|     x 0.5 |  400| -200| 0.5| -1.151| -0.602| 2.5| 1|
-| Reference |  800| -400| 1.0| -2.303| -1.203|   5| 2|
-|     x 1.5 | 1200| -600| 1.5| -3.454| -1.806| 7.5| 3|
-
-[**Table 1**]{#tbl-sens-params}.
-Parameter ranges for sensitivity analysis to be used as the limits to sample from uniform distribution and designed to evaluate model response to variability.
-Each parameter's range reflects a ±50% variation from its reference value.
-It is worth noting that reference values for $\lambda$ and $\phi$ parameters were set considering a reference rate of 90% and 70% change per minute in the original scale of these parameters.
-
-Sobol sensitivity analysis was selected due to its robustness in decomposing the variance attributed to each parameter in complex, non-linear, and non-monotonic models, such as the present logistic model of RRi.
-While methods like the Morris screening technique offer computational efficiency, Sobol's approach provides a more comprehensive assessment of parameter interactions, making it particularly suitable for physiological models where interaction effects may be critical to interpret.
-
-## Usage of real-world RRi data
-
-To further assess the performance and applicability of the proposed model, real-world RRi data were analyzed in addition to the synthetic data generated through simulation.
+To further assess the performance of the proposed model, real-world RRi data were analyzed in addition to the synthetic data generated through simulation.
 This dataset was derived from a cohort participating in the FONDECYT Project No.
@@
 
-This real-world dataset provided a critical context for validating the model’s predictive capability against observed physiological responses, offering a robust foundation for understanding RRi dynamics under physical activity conditions.
+This real-world dataset provided a critical context for validating the model's predictive capability against observed physiological responses, offering a robust foundation for understanding RRi dynamics under physical activity conditions.
 
@@
 
-Parameter estimation was performed using Hamiltonian Monte Carlo (HMC) with the No-U-Turn Sampler (NUTS) to explore the parameter space.
-This method suits high-dimensional spaces and utilizes gradient information for efficient sampling.
+Parameter estimation was performed using Hamiltonian Monte Carlo (HMC) with the No-U-Turn Sampler (NUTS) to explore the parameter space [@hoffman2014no].
 The parameters $\alpha$, $\beta$, $c$, $\lambda$, $\phi$, $\tau$, and $\delta$ were estimated by sampling from the posterior distribution, which was constructed from observed RRi data and model predictions.
-The Bayesian framework allows the incorporation of prior distributions for parameters, enhancing the reliability of the estimates.
 
-The gradient of the log-likelihood function for each parameter was computed during estimation using the brms R package, which employs the Stan probabilistic programming language.
-Convergence of the HMC chains was assessed using standard diagnostics, including R-hat values, which were kept below 1.01 for all parameters, and effective sample sizes, which were targeted at a minimum of 1,000 for each parameter.
-Trace plots were inspected to confirm stable mixing, and potential issues with autocorrelation were addressed by increasing the warm-up period to ensure chain independence.
+The gradient of the log-likelihood function for each parameter was computed during estimation using the brms R package (v2.21.0), which employs the Stan probabilistic programming language.
+Convergence of the HMC chains was assessed using standard diagnostics, including R-hat values, which were kept below 1.01 for all parameters [@vehtari2019rank], and effective sample sizes, which were targeted at a minimum of 1,000 for each parameter [@brms].
+Trace plots were inspected to confirm stable mixing.
 These diagnostics collectively confirmed reliable posterior estimates for each parameter.
@@
 
-To enhance the exploration of parameter space, we performed a two-stage analysis: individual-level and group-level estimates.
+To enhance the exploration of parameter space, we performed a two-stage analysis: An individual-level estimation of parameter values that were then used for the estimation population-level parameters.
 
@@
 
-Firstly, each subject's RRi data $\text{RRi}_{i,t}$ was standardized against his mean $\bar{\text{RRi}_i}$ and standard deviation $\sigma_{\text{RRi}_i}$ to improve convergence and exploration of the posterior distribution.
+Firstly, each subject's RRi data $\text{RRi}_{i,t}$ was standardized against his mean $\bar{\text{RRi}_i}$ and standard deviation $S_{\text{RRi}_i}$ to improve convergence and exploration of the posterior distribution.
 The standardized RRi data $y_{i,t}$ for each time point $t$ was computed as:
@@
 $$
-y_{i,t} = \frac{\text{RRi}_{i,t} - \bar{\text{RRi}_i}}{\sigma_{\text{RRi}_i}}
-$$
+y_{i,t} = \frac{\text{RRi}_{i,t} - \bar{\text{RRi}_i}}{S_{\text{RRi}_i}}
+$${#eq-ind-std}
 This standardization allowed the model to focus on relative changes in RRi dynamics, independent of individual baseline differences.
@@
 y_{i,t} = \alpha_i + \frac{\beta_i}{1 + e^{\lambda_i \cdot (t - \tau_i)}} + \frac{-\beta_i \cdot c}{1 + e^{\phi_i \cdot (t - \tau_i - \delta_i)}} + \epsilon_{i,t}
-$$
+$${#eq-indiv-level}
 where $\alpha_i$, $\beta_i$, $c_i$, $\lambda_i$, $\phi_i$, $\tau_i$, $\delta_i$ are the individual-specific model parameters and $\epsilon_{i,t} \sim \mathcal{N}(0, \sigma^2)$ is the residual error term at each time point $t$.
@@
 \begin{aligned}
-\alpha_i^{\text{RRi}} &= \alpha_i \cdot \sigma_{\text{RRi}_i} + \bar{\text{RRi}}_i \\
-\beta_i^{\text{RRi}} &= \beta_i \cdot \sigma_{\text{RRi}_i}
+\alpha_i^{\text{RRi}} &= \alpha_i \cdot S_{\text{RRi}_i} + \bar{\text{RRi}}_i \\
+\beta_i^{\text{RRi}} &= \beta_i \cdot S_{\text{RRi}_i}
 \end{aligned}
-$$
-
-Priors were chosen based on physiological constraints and graphical visualization of standardized RRi data, ensuring identifiability of model parameters by constraining the parameter space to plausible values, which improves model convergence.
-The prior distributions are defined as follows:
-
-$$
-\begin{aligned} 
-\alpha &\sim \mathcal{N}(1, 0.5) \\
-\beta &\sim \mathcal{N}(-2.5, 0.5) \text{ with } \beta \leq 0 \\
-c &\sim \mathcal{N}(0.8, 0.2) \text{ with } c \geq 0 \\
-\lambda &\sim \mathcal{N}(-2, 0.5) \text{ with } \lambda \leq 0 \\
-\phi &\sim \mathcal{N}(-2, 0.5) \text{ with } \phi \leq 0 \\
-\tau &\sim \mathcal{N}(5, 0.5) \text{ with } \tau \geq 0 \\
-\delta &\sim \mathcal{N}(5, 0.5) \text{ with } \delta \geq 0
-\end{aligned}
-$$
-
-Simulated standardized RRi dynamics based on prior parameter distributions are shown in [Figure 1](#fig-prior-sim).
-
-![](figures/prior-model.png)
-
-[**Figure 1**]{#fig-prior-sim}.
-Simulated standardized RRi dynamics based on prior parameter distributions, depicting predicted RRi responses to exercise.
-Shaded areas represent 95%, 80%, and 60% quantile credible intervals, offering insight into expected physiological variability across parameters.
+$${#eq-ind-unstd}
 
@@
 These estimates were then used as input data to create a univariate hierarchical model, capturing variability at both the subject and group levels.
-The model is described as follows:
+The modeling process is described as follows:
 
-For each subject $i$, we estimated an interdependent stochastic process in which the true parameter $\theta_{k,i}$, with $k \in \{\alpha, \beta, c, \lambda, \phi, \tau, \delta\}$ with their corresponding standard error $\epsilon_{k,i}$ was used to model the observed parameter $\hat{\theta}_{k,i}$ as:
+For each subject $i$, we estimated an interdependent stochastic process in which the true parameter $\theta_{k,i}$, with $k \in \{\alpha, \beta, c, \lambda, \phi, \tau, \delta\}$ with their corresponding standard error $\epsilon_{k,i}$ was used to model the observed parameter $\theta^{\text{ obs}}_{k,i}$ as:
 
@@
 \theta^{\text{ obs}}_{k,i} \sim \mathcal{N}(\theta_{k,i}, \epsilon_{k,i})
-$$
+$${#eq-gp-observed}
 
@@
 \theta_{k,i} \sim \mathcal{N}(\mu_{k} + b_{k,i}, \sigma_k^2)
-$$
+$${#eq-gp-true}
 
@@
 
-## Model Evaluation
+## Model Performance
 
-The primary performance metrics included R^2^, root mean square error (RMSE) and mean absolute percentage error (MAPE), estimated for each subject.
-Bootstrap resampling across each metric was performed to estimate the mean performance of the model.
-These metrics were selected for their capacity to assess model capacity to predict the observed RRi and to quantify absolute and relative errors, thereby providing a comprehensive assessment across the varying scales of RRi dynamics, including resting baseline, peak decline, and recovery.
+The primary performance metrics included R^2^, root mean square error (RMSE), and mean absolute percentage error (MAPE), estimated for each subject.
+Bootstrap resampling across each metric was performed to estimate the mean performance of the model and corresponding quantile-based 95% CI.
 
-Also, residual analysis was conducted to evaluate the model's accuracy in capturing RRi dynamics.
+Also, residual analysis were conducted to evaluate the model's accuracy in capturing RRi dynamics.
 Residuals were defined as the difference between observed and predicted RRi values.
@@
 
+## Model parameters sensitivity
+
+Once a model that described RRi behavior in response to exercise was obtained, an assessment of the proportion of the variance explained by each model parameter was then computed.
+
+To assess the sensitivity of model parameters influencing RRi over time, we implemented a Sobol sensitivity analysis using Monte Carlo simulations.
+Sobol index ($S_{\text{ind}}$) provide a measure of the proportion of the contribution of each parameter to the variance in RRi at each time point, and it was selected for its robustness in handling non-linear and non-monotonic relationships, which are intrinsic to RRi dynamics in response to exercise [@zhang2015sobol].
+
+To compute $S_{\text{ind}}$, 1000 Monte Carlo simulations were conducted, each involving 1000 randomly sampled parameter sets (1,000,000 model runs in total).
+For each set of parameters, RRi were calculated at each time point $t$ across a range from 0 to 20 minutes at intervals of 0.1 minutes.
+The 95% CI parameter values estimated from HMC-NUTS were then used as input ranges for $S_{\text{ind}}$ computation.
+Finally, the mean values of $S_{\text{ind}}$ over the 20-min time span for each model parameter were estimated and reported, next with their corresponding 95% CI using normal approximation based on estimated standard errors (SE).
+
 # Results
 
-## Non-linear model and deterministic behaviour
+## Problem characterization
 
-### RRi as a linear combination of logistic functions
+RRi dynamics in response to exercise tends to follow a U-shaped form.
+The initial decrease in RRi is associated with exercise onset, related to an increase in heart rate.
+After exercise cessation, an opposite increase in RRi is observed, associated with the cardiovascular recovery phase.
+In both cases, the drop and recovery phases occurs at different rates, some individuals experience a quick recovery in RRi after exercise, however, in some others this slope is less steep.
+Additionally, the new baseline reached following exercise cessation is often below the RRi baseline before exercise.
 
-According to the proposed model in @eq-main-model, the dynamics of RRi in response to physical exertion can be represented as a linear combination of a baseline RRi $\alpha$ and two logistic functions denoted as $f_1(t)$ and $f_2(t)$.
+These hallmarks of RRi dynamics in response to exercise, highlights the complex and non-linear behavior of the cardiovascular response in the context of both rest and exercise conditions.
+An example RRi record data is shown in [figure 1](#fig-rri-example).
+
+<!---- FIGURE 1 ---->
+
+## Model construction
+
+The process of deriving the final equation for modeling RRi fluctuations was guided by an iterative exploration of mathematical functions capable of capturing the observed dynamics.
+Initially, exponential and logarithmic functions were considered due to their simplicity and wide applicability in describing temporal changes.
+Exponential functions were hypothesized to capture the rapid initial adaptations of RRi post-exercise onset, while logarithmic functions were explored for their capacity to describe asymptotic behaviors observed in some physiological variables.
+
+However, neither approach successfully reproduced the non-linear and bidirectional nature of the RRi fluctuations.
+Exponential functions, while effective at modeling monotonic decay or growth, could not account for the observed sigmoidal transitions.
+Similarly, logarithmic functions, with their inherent monotonicity, failed to represent the plateauing behavior seen in real-world data.
+
+To address these limitations, we shifted to logistic functions, which inherently model sigmoidal transitions.
+Logistic functions introduce parameters for growth rate and inflection point, allowing for precise control over the shape and timing of the transition between dynamic states.
+By using two coupled logistic functions, one to represent the initial decrease in RRi and a second, inverted logistic function to describe the recovery phase, we achieved a model structure that could flexibly reproduce the observed non-linear variations.
+
+This approach provided a biologically plausible representation, with parameters that directly correspond to identifiable physiological features, such as the rate of adaptation and recovery, the time to peak response, and the extent of deviation from baseline.
+The logistic function framework emerged as the optimal solution after systematic testing and evaluation against empirical data, ensuring that the model accurately captured both the qualitative and quantitative aspects of RRi dynamics.
+
+The mathematical model proposed to characterize the RRi response to exercise and recovery is defined by @eq-main-model.
+
+$$
+\begin{aligned}
+\text{RRi}(t) = \alpha + \frac{\beta}{1 + e^{\lambda(t - \tau)}} + \frac{-c \cdot \beta}{1 + e^{\phi(t - \tau - \delta)}}
+\end{aligned}
+$${#eq-main-model}
+
+This model includes two logistic functions representing the RRi dynamics across exercise and recovery phases.
+The first logistic term models the decrease in RRi during exercise, where the parameter $\beta$ denotes the magnitude of this decline.
+The rate of decrease is governed by $\lambda$, while $\tau$ represents the onset of the RRi decrease or the time the physiological shift begins.
+
+The second logistic term accounts for RRi recovery post-exercise.
+Here, $c$ scales the magnitude of recovery relative to the initial decline represented by $\beta$, capturing the proportion of the decline regained during recovery.
+The rate at which RRi returns to baseline is controlled by $\phi$, and $\delta$ indicates the lag following the cessation of exercise, marking the beginning of recovery, respectively.
+
+Additionally, the dynamics of RRi in response to physical exertion can be represented as a linear combination of a baseline RRi $\alpha$ and two logistic functions denoted as $f_1(t)$ and $f_2(t)$.
 The function $f_1(t)$ models the initial decay in RRi following the initiation of exercise while $f_2(t)$ characterizes the recovery phase after exercise cessation.
 
-The fundamental structure of both logistic functions can be expressed as:
+Essentially, the fundamental structure of both logistic functions can be expressed as:
 
@@
 For $f_1(t)$, this parameter is specified as $\beta$, indicating the absolute change in RRi at the onset of exercise.
-In contrast, for $f_2(t)$, $a_1$ is reparameterized as $-\beta \cdot c$, where $c$ denotes the proportion of change relative to the initial drop indicated by $\beta$.
-This reparameterization ensures that, after the initial decline, the second logistic function facilitates the return of RRi toward the baseline value $\alpha$.
+In contrast, for $f_2(t)$, $a_1$ is parametrized as $-c \cdot \beta$, where $c$ denotes the proportion of change relative to the initial drop indicated by $\beta$.
+This parametrization ensures that, after the initial decline, the second logistic function facilitates the return of RRi toward the baseline value $\alpha$.
 
@@
 This rate parameter is expressed on a logarithmic scale; to convert it to a percentage change per unit of time, it can be scaled as $1 - \exp(a_2)$.
-For instance, a 90% decrease per unit time corresponds to $a_2 = \log(1 - 0.9)$, resulting in an approximate value of -2.302585.
+<!-- For instance, a 90% decrease per unit time corresponds to $a_2 = \log(1 - 0.9)$, resulting in an approximate value of -2.302585.
+-->
 
@@
 
-![](figures/model-constituents.png)
+<!---- FIGURE 2 ---->
 
-[**Figure 2**]{#fig-linear-constituents}.
-The RRi dynamics in response to exercise are expressed as a linear combination of model constituents based on the baseline RRi $\alpha$ and two logistic functions, denoted $f_1(t)$ and $f_2(t)$, respectively.
+## Sample characteristics
 
-### Sensitivity to parameter variability
+The sample used to assess RRi dynamics consists of a group of 272 subjects selected from a local community of elderly individuals.
+The sample characteristics can be seen in [Table 1](#tbl-sample-characteristics)
+
+<!---- TABLE 1 ---->
+
+Initial graphical exploration of RRi dynamics (see [Figure 3](#fig-2d-kernel-density)) indicates a clear drop in RRi around the 5-7 minutes mark, associated the exercise-induced cardiovascular stress.
+However, greater variability across individuals in post-exercise recovery can be observed.
+
+<!---- FIGURE 3 ---->
+
+## Parameter estimation
+
+### Priors
+
+Given the observed parameters that reproduced the observed RRi patterns in exercise and rest conditions, priors were chosen based on physiological constraints and graphical visualization of standardized RRi data.
+Hence, ensuring identifiability of model parameters by constraining the parameter space to plausible values, in order to improve model convergence and parameter exploration.
+The prior distributions were defined as follows:
+
+$$
+\begin{aligned} 
+\alpha &\sim \mathcal{N}(1, 0.5) \\
+\beta &\sim \mathcal{N}(-2.5, 0.5) \text{ with } \beta \leq 0 \\
+c &\sim \mathcal{N}(0.8, 0.2) \text{ with } c \geq 0 \\
+\lambda &\sim \mathcal{N}(-2, 0.5) \text{ with } \lambda \leq 0 \\
+\phi &\sim \mathcal{N}(-2, 0.5) \text{ with } \phi \leq 0 \\
+\tau &\sim \mathcal{N}(5, 0.5) \text{ with } \tau \geq 0 \\
+\delta &\sim \mathcal{N}(5, 0.5) \text{ with } \delta \geq 0
+\end{aligned}
+$${#eq-pe-priors}
+
+Simulated standardized RRi dynamics based on prior parameter distributions are shown in [Figure 4](#fig-prior-sim).
+
+<!---- FIGURE 4 ---->
+
+### Parameter estimates
+
+Once subject-level RRi data was fitted using the proposed in @eq-indiv-level, a population-parameter value was estimated based on the proposed group-level methodology.
+The estimated parameter values can be seen in [Table 2](#tbl-parameters)
+
+<!---- TABLE 2 ---->
+
+In [Figure 5](#fig-group-level-estimates), the model parameter's posterior distribution can be observed.
+
+<!---- FIGURE 5 ---->
+
+## Model evaluation
+
+### Model performance
+
+Relative performance metrics, estimated through bootstrapped resampling, suggest that the model tends to deviate a 3.4% (CI~95%~[3.06, 3.81]) from the observed RRi data.
+This is equivalent to a 32.6 ms in the RRi scale (CI~95%~[30.01, 35.77]).
+Additionally, the bootstrapped R^2^ indicates that the model explains 0.868 (CI~95%~[0.834, 0.895]) of the total variance observed in RRi.
+
+Residuals analysis showed that the estimated partial correlation function (ACF) from the model residuals indicates a correlation among non-explained errors greater than 0.1 up to the 5th lag.
+However, the partial ACF is significant (CI-wise) and strictly positive or negative up to the second lag.
+Correlations among model residuals against other time indices remained insignificant (see [Figure 6](#fig-error-rates)).
+
+<!---- FIGURE 6 ---->
+
+### Model parameters sensitivity
 
@@
 Conversely, the rate parameters for the initial decay during exercise, $\lambda$, and the recovery post-exercise, $\phi$, show lower sensitivity, suggesting that they are not primary sources of variation in predicted RRi trajectories when assessed in isolation.
-The results of the sensitivity analysis are illustrated in [Figure 3](#fig-sensitivity).
+The results of the sensitivity analysis can be seen in [Table 3](#tbl-sens-params).
 
-![](figures/sobol-sensitivity.png)
+<!---- TABLE 3 ---->
 
-[**Figure 3**]{#fig-sensitivity}.
-Sensitivity analysis results illustrate the impact of parameter variability on model predictions and the percentage of explained variance accounted for each model parameter.
-Shaded areas represent 95%, 80%, and 60% CI estimated from Monte Carlo samples.
-Time in the x-axis is truncated at 12 minutes, given that around this time, the proportion of the total variance explained by the respective model parameter stabilizes respective of time.
+# Discussion
 
-Despite the limited sensitivity of $\lambda$ and $\phi$ to mean RRi, these parameters play a pivotal role in determining the rate of change in RRi in response to physical exertion.
-Therefore, they can significantly affect RRi dynamics over time.
-[Figure 4](#fig-data-sim) depicts the influence of these rate parameters on RRi fluctuations over time.
+To the best of our knowledge, this study represents the first attempt to develop a non-linear model specifically designed to continuously capture RR interval dynamics across a complete rest-exercise-rest protocol.
+Previous studies have either focused on aggregate HRV indices or utilized simplified linear or exponential models, which are insufficient to describe the complex, non-stationary transitions observed during and after exercise [@molkkari2020online].
+By employing a combination of logistic functions, our model uniquely accounts for the gradual shifts in autonomic regulation denoted by RRi dynamics, offering a detailed and physiologically relevant representation of cardiac dynamics.
+This continuous modeling framework allows for the integration of both exercise-induced RRi decline and post-exercise recovery within a single unified structure, bridging a critical gap in the current literature.
+Such an approach not only advances our understanding of cardiovascular responses but also opens new avenues for real-time monitoring and intervention in both clinical and athletic settings.
 
-![](figures/rate-sensitivity.png)
+The proposed model demonstrates a precise capacity to reproduce RRi dynamics, with its combination of logistic functions, capturing the key transitions of cardiac response, the initial decline during exercise and the subsequent recovery.
+This design accommodates the inherent non-linearity and non-stationarity of RRi dynamics, overcoming the limitations of linear models and exponential functions commonly used in prior studies [@silva2019exponential; @mongin2022decrease].
 
-[**Figure 4**]{#fig-data-sim}.
-Simulated RRi dynamics during exercise with varying $\lambda$ and $\phi$ rate parameters expressed as percent change per unit of time.
-In this simulation, the exercise-induced RRi drop occurs at 5 minutes ($\tau$), and cardiovascular recovery begins 2 minutes after exercise initiation ($\delta$).
-The model assumes a 90% recovery ($c$ = 0.90) of RRi values following a 400 ms drop ($\beta$) from a baseline of 800 ms ($\alpha$).
+Compared to previous research, our findings align with efforts made on capturing nonlinear dynamics in HRV to understand cardiac responses during exercise [@gronwald2019effects].
+Similarly, previous studies have shown that dynamic fluctuations in RRi can serve as critical indicators of cardiorespiratory fitness, supporting the need for models to address the complexity of cardiovascular responses during physical stress [@mongin2022hrv].
+However, while many existing models focus primarily on linear metrics or aggregate HRV measures, our study provides a high-resolution analysis of RRi dynamics that enhances interpretability and application across diverse fitness levels and exercise intensities.
 
-## Model behavior to real RRi data
+The flexibility of the logistic components allows for physiologically interpretable parameters, such as baseline RRi ($\alpha$) and recovery proportion ($c$), which directly correlate with intrinsic cardiac function and autonomic recovery capacity, respectively.
+These features position the model as a robust framework for investigating the cardiovascular system's dynamic adaptation to physical stressors.
+For example, prior studies have highlighted the inadequacy of linear HRV metrics in capturing transient autonomic shifts [@gregoire2023autonomic]; our results align with this critique, demonstrating the advantages of modeling RRi directly.
 
-### Sample characteristics
+Unlike prior research that aggregates HRV measures or applies simple decay models, our approach directly models RRi changes, offering richer physiological insight.
+For instance, commonly utilized exponential decay models for post-exercise recovery are used, however they fail to incorporate the transition dynamics observed during exercise itself [@sahoo2024unanticipated].
+By integrating both exercise and recovery phases, our model provides a more comprehensive view of autonomic regulation.
 
-The sample used to assess RRi dynamics consists of a group of 272 subjects selected from a local community of elderly individuals.
-The sample characteristics can be seen in [Table 2](#tbl-sample-characteristics)
+Moreover, the sensitivity of parameters such as $\lambda$ (decay rate) and $(\phi$ (recovery rate) was found to be relatively low, suggesting that the model is robust to variability in these rates while remaining sensitive to key physiological parameters ($\alpha$ and $c$).
+This robustness makes it suitable for both individualized monitoring and population-level analyses, offering versatility in its application across different use cases.
+
+The Sobol sensitivity analysis revealed that baseline RRi ($\alpha$) and recovery proportion ($c$) are the primary drivers of model output variance, emphasizing their physiological importance.
+These findings are consistent with prior research, which identified baseline cardiac function as a determinant of cardiovascular health and recovery proportion as a marker of autonomic resilience [@hautala2003short].
+
+However, the Sobol method assumes parameter independence, which may overlook interactions that are common in biological systems [@harenberg2017uncertainty; @cheng2017mixed; @herman2013method].
+For example, the interplay between $\lambda$ and $c$, which dictates the rate and magnitude of recovery, is likely critical but remains unexplored in the current framework.
+Future studies could explore Bayesian sensitivity analysis or variance decomposition methods that account for parameter interdependence [@bornn2010efficient; @xue2021bayesian].
+
+This model demonstrates significant potential for practical applications in both clinical and athletic settings.
+In clinical contexts, it could aid in tailoring cardiovascular rehabilitation protocols by monitoring autonomic recovery in real-time, ensuring safe and effective exercise regimens for at-risk populations [@carrasco2024hrv].
+This aligns with previous research, which highlight the importance of individualizing rehabilitation programs to optimize recovery [@takahashi2020signs; @rizvi2023enhancing; @carrasco2024hrv].
+
+In athletic settings, the model could guide training strategies, particularly for interval training, where determining optimal recovery periods is crucial.
+Similar findings suggest that precise monitoring of RRi dynamics can prevent overtraining and enhance performance [@hebisz2022heart; @nuuttila2024monitoring].
+The model's ability to integrate real-time data from wearable devices further enhances its applicability in dynamic, uncontrolled environments, enabling field-based monitoring and feedback [@zimatore2020recurrence].
+
+While the model presents substantial advances, it has limitations that warrant consideration.
+First, the assumption of uniform parameter sampling in sensitivity analysis, while practical, may not fully capture the variability observed in populations with extreme autonomic profiles [@kristal1995heart].
+Empirical distributions or Bayesian priors, could improve parameter estimation and enhance the model's applicability to diverse populations [@xue2021bayesian].
+
+Another limitation lies in the demographic composition of the sample, which consisted exclusively of elderly individuals.
+While this population provides valuable insights into age-specific cardiovascular dynamics, the findings may not fully generalize to younger populations, whose autonomic responses to exercise and recovery differ significantly due to higher baseline vagal tone, greater cardiac plasticity, and distinct metabolic profiles [@castillo2021cardiac; @garavaglia2021effect].
+Previous studies have demonstrated that younger individuals exhibit faster autonomic recovery and greater adaptability during physical exertion compared to older populations, suggesting that the parameter estimates derived from this model may vary across age groups [@choi2020declining; @garavaglia2021effect].
+Future research should validate the model in more diverse cohorts, including younger adults and athletes, to ensure broader applicability and to explore potential age-dependent modifications of the model's parameters.
+This would enhance its utility in clinical and athletic contexts, where age diversity is a critical factor [@choi2020declining; @garavaglia2021effect].
+
+Additionally, environmental and psychological factors, such as temperature, stress, or sleep quality, were not explicitly considered in this study.
+Future work could integrate these variables into the model, enhancing its robustness and applicability across varied real-world scenarios.
+This aligns with calls for more integrative modeling approaches in cardiovascular research [@takahashi2020signs; @hebisz2022heart; @nuuttila2024monitoring].
+
+# Conclusion
+
+In summary, this study presents a novel non-linear model for RRi dynamics, capturing the complex and transient autonomic responses during rest-exercise-recovery protocols, overcoming the limitations of traditional autonomic metrics.
+By identifying baseline RRi and recovery proportion as the dominant contributors to variability, the model emphasizes their critical roles in reflecting autonomic regulation and resilience.
+Validated across a cohort of elderly participants, the model demonstrates robust performance in real-time cardiovascular assessments, offering significant potential for clinical applications such as rehabilitation and monitoring in at-risk populations, as well as athletic contexts like fatigue management and performance optimization.
+While the model's applicability is currently constrained by its focus on elderly individuals, future validation in younger cohorts and under diverse environmental conditions will enhance its generalizability and utility.
+This work establishes a foundational framework for advancing personalized cardiovascular health monitoring and intervention.
+
+# References
+
+<div id="refs"></div>
+
+# Author Contributions
+
+Conceptualization, MC-A; Data curation, MC-A; Investigation, MC-A; Methodology, MC-A, NMD; Supervision, CN-E; Formal analysis, MC-A; Visualization, MC-A; Writing–original draft, MC-A, CN-E, DM-C; Writing–review & editing, MC-A, CN-E, DM.
+All authors have read and agreed to the published version of the manuscript.
+
+# Data Availability Statement
+
+The data supporting the conclusions of this article will be available from the authors without reservation.
+
+# Competing Interests Statement
+
+The authors declare that this research was conducted without any commercial or financial relationships that could be construed as potential conflicts of interest.
+
+# Funding
+
+This work was funded by ANID Proyecto Fondecyt Iniciación Nº11220116.
+
+{{< pagebreak >}}
+
+# Figures
+
+![](figures/example-data.png)
+[**Figure 1**]{#fig-rri-example}.
+Example data of RRi recordings of 6 different subjects over a 20 minute rest-exercise-recovery protocol in a sample of elderly individuals.
+The subject-level data shows the inter-individual variability of RRi dynamics in response exercised-induced cardiovascular stress, with similar behavior and trajectories of recovery over time.
+
+{{< pagebreak >}}
+
+![](figures/model-constituents.png)
+[**Figure 2**]{#fig-linear-constituents}.
+The RRi dynamics in response to exercise are expressed as a linear combination of model constituents based on the baseline RRi $\alpha$ and two logistic functions, denoted $f_1(t)$ and $f_2(t)$, respectively.
+The vertical dashed lines represent the time at which the exercise and recovery onset given by $\tau = 5$ and $\delta = 2$.
+
+{{< pagebreak >}}
+
+![](figures/2d-density-kernel.png)
+[**Figure 3**]{#fig-2d-kernel-density}.
+(A) Mean and SD from each of the subject's RRi recordings, used for the standardization process.
+(B) 2D kernel density of standardized RRi dynamics over time from a sample of individuals subjected to the rest-exercise-rest protocol.
+Darker colors indicate greater probability density.
+The contrary can be said about lighter colors.
+
+{{< pagebreak >}}
+
+![](figures/prior-model.png)
+[**Figure 4**]{#fig-prior-sim}.
+(A) Simulated standardized RRi dynamics based on prior parameter distributions, illustrating predicted RRi responses to exercise.
+Shaded areas represent 95%, 80%, and 60% quantile CI, offering insight into expected physiological variability across parameters.
+(B) Prior distributions and 95% CI, used to generate prior predictions, based on physiological constraints and graphical visualization of standardized RRi data.
+
+{{< pagebreak >}}
+
+![](figures/group-level-estimates.png)
+[**Figure 5**]{#fig-group-level-estimates}.
+(A) Posterior probability distributions of the expectation for each population-parameter estimates ($E[\theta]$) with quantile-based 95% CI.
+(B) Transformed rate parameters into a percentage scale using the $1 - \exp(\theta)$ transformation.
+
+{{< pagebreak >}}
+
+![](figures/error-estimates.png)
+[**Figure 6**]{#fig-error-rates}.
+Individual-level performance metrics.
+(A) Bootstrapped MAPE and RMSE, as metrics of relative and absolute model deviance from observed RRi.
+(B) Individual-level estimates of model performance and the relationship between them.
+(C) Partial autocorrelation function (ACF) of model residuals with corresponding quantile-based CI.
+
+{{< pagebreak >}}
+
+# Tables
 
 |Characteristic |Overall        |Female         |Male           |
-|:--------------|:--------------|:--------------|:--------------|
+|:--------------|--------------:|--------------:|--------------:|
 |Sex            |—              |217 (79.8%)    |55 (20.2%)     |
@@
 
-[**Table 2**]{#tbl-sample-characteristics}. Sample characteristics from which, continuous RRi monitoring data was collected during a rest-exercise-rest protocol. Data is presented as Mean ± SD.
+[**Table 1**]{#tbl-sample-characteristics}.
+Sample characteristics from which, continuous RRi monitoring data was collected during a rest-exercise-rest protocol.
+Data is presented as Mean ± standard deviation (SD).
+SBP, systolic blood pressure; DBP, diastolic blood pressure; MAP, mean arterial pressure; PP, pulse pressure; BMI, body mass index.
 
-```{r}
-data("rri_data")
-```
+{{< pagebreak >}}
 
-Initial exploration of RRi dynamics using two-dimensional density kernel estimation (see [Figure 5](#fig-2d-kernel-density)) indicates a clear drop in RRi around the 5-7 minutes mark, associated the exercise-induced cardiovascular stress.
-However, greater variability across individuals in post-exercise recovery can be observed.
+| Parameter | Estimate^1^ | SE^1^ | Lower^2^ | Upper^2^ |
+|-----|---:|---:|---:|---:|
+| $\alpha$  |  861.78 | 5.73 |  850.57 |  872.85 |
+| $\beta$   | -345.49 | 7.41 | -359.81 | -330.97 |
+| $c$       |    0.84 | 0.01 |    0.82 |    0.86 |
+| $\lambda$ |   -3.05 | 0.06 |   -3.16 |   -2.94 |
+| $\phi$    |   -2.60 | 0.06 |   -2.71 |   -2.48 |
+| $\tau$    |    6.71 | 0.05 |    6.61 |    6.81 |
+| $\delta$  |    3.24 | 0.10 |    3.05 |    3.44 |
+| $\sigma$  |   27.57 | 0.57 |   26.45 |   28.70 |
 
-![](figures/2d-density-kernel.png)
+[**Table 2**]{#tbl-parameters}.
+Population-parameter values estimated from group-level analysis.
+^1^ Estimates and SE are computed as median and mean absolute deviation of the posterior distribution, respectively; ^2^ Lower and Upper bounds from the quantile-based CI~95%~ of the posterior distribution.
 
-[**Figure 5**]{#fig-2d-kernel-density}.
-Left panel shows the mean and standard deviation (SD) from each of the subject's RRi recordings, used for the standardization process.
-Right panel indicates the 2D kernel density of standardized RRi dynamics over time from a sample of individuals subjected to the rest-exercise-rest protocol.
-Darker colors indicate greater probability density.
-The contrary can be said about lighter colors.
+{{< pagebreak >}}
 
-### Model performance
+|Parameter | Estimate^1^|  SE^1^ | Lower^2^| Upper^2^|
+|:---------|-----------:|-------:|--------:|--------:|
+|$\alpha$  |     0.56808| 0.01813|  0.53255|  0.60361|
+|$\beta$   |     0.02378| 0.00111|  0.02160|  0.02596|
+|$c$       |     0.21406| 0.00914|  0.19615|  0.23197|
+|$\lambda$ |     0.00045| 0.00002|  0.00041|  0.00049|
+|$\phi$    |     0.00012| 0.00001|  0.00010|  0.00014|
+|$\tau$    |     0.04031| 0.00160|  0.03717|  0.04345|
+|$\delta$  |     0.15387| 0.00291|  0.14817|  0.15957|
 
-Relative performance metrics, estimated through bootstrapped resampling, suggest that the model tends to deviate a 3.4% (CI~95%~[3.06, 3.81]) from the observed RRi data.
-This is equivalent to a 32.6 ms in the RRi scale (CI~95%~[30.01, 35.77]).
-Residuals analysis showed that the estimated partial correlation function (ACF) from the model residuals indicates a correlation among non-explained errors greater than 0.1 up to the 5th lag.
-However, the partial ACF is significant (CI-wise) and strictly positive or negative up to the second lag.
-Correlations among model residuals against other time indices remained insignificant (see [Figure 6](#fig-error-rates)).
-
-![](figures/error-estimates.png)
-[**Figure 6**]{#fig-error-rates}.
-Individual-level performance metrics.
-Left upper panels indicates bootstrapped MAPE and RMSE, as metrics of relative and absolute model deviance from observed RRi.
-Upper right panel shows the individual-level estimates of model performance and the relationship between them.
-Lower panel indicates the partial autocorrelation function (ACF) of model residuals with corresponding quantile-based CI.
-
-### Group-level estimates
-
-Once subject-level RRi data was fitted using the proposed non-lineaer equation, a population-parameter value was estimated with a generalized hierarchical model with measurement error as part of an interdependent stochastic process using the mean and standard error of the estimated posterior distribution of the subject-level parameter, derived from HMC-NUTS.
-
-The $alpha$ parameter, associated with the baseline RRi before exercise, was estimated at 861.8 ms CI~95%~[850.6, 872.9].
-The $beta$ parameter, which indicates exercise-induced drop in RRi, was estimated at -345.5 ms CI~95%~[-359.8, -331.0].
-The proportion of recovery $c$, which can be expressed a the magnitude relative to $beta$ after exercise cessation, was estimated at 0.84 CI~95%~[0.823, 0.856].
-
-The rate parameters $\lambda$ and $\phi$, associated with the velocity of RRi drop and recovery, were estimated at -3.05 CI~95%~[-3.16, -2.94] and -2.60 CI~95%~[-2.71, -2.48], and after transforming them into a percentage of decay per minute using $1 - exp(\theta)$, we estimated the rate of decay and recovery are 95.2% ms per minute CI~95%~[0.947, 0.957] and 92.5% ms per minute CI~95%~[0.917, 0.933].
-
-The parameters associates with the time at which the initial RRi decay occurs $\tau$ were estimated at 6.71 minutes CI~95%~[6.61, 6.81], meanwhile the time duration associated with the exercise-induced stress in RRi $\delta$ was estimated at 3.24 minutes CI~95%~[3.05, 3.44].
-
-Finally, the standard deviation around the estimated RRi data $\sigma$, associated with the non-modeled noised, was estimated at 27.6 ms CI~95%~[26.5, 28.7].
-In [Figure 7](#fig-group-level-estimates), the model parameter's posterior distribution can be observed.
-
-![](figures/group-level-estimates.png)
-[**Figure 7**]{#fig-group-level-estimates}.
-Posterior probability distributions from group-level estimates of the expectation of each model parameter ($E[\theta]$) with quantile-based 95% CI (upper panel), and the transformed rate parameters into a percentage scale (lower panel).
-
-# Discussion
-
-This study presents a non-linear model designed for RRi fluctuations during exercise and recovery, providing an in-depth view of cardiac autonomic dynamics.
-By focusing on RRi rather than aggregate HRV metrics, our approach enables real-time tracking of cardiovascular responses to physical stress, which has implications for clinical and athletic applications.
-This model marks a step forward in computational physiology, bridging the gap between theoretical frameworks and practical monitoring applications.
-
-<!--- Sensitivity Analysis and Model Robustness --->
-The Sobol sensitivity analysis revealed baseline RRi ($\alpha$) and recovery proportion ($c$) as key drivers of model output variance that are consistent with physiological expectations.
-Baseline RRi reflects intrinsic cardiac properties, while recovery proportion captures autonomic re-engagement post-exercise, central to cardiovascular adaptation [@hautala2003short].
-The relatively low sensitivity of decay and recovery rate parameters ($\lambda$ and $\phi$) suggests the model is robust to moderate fluctuations in these rates, enhancing its reliability for both individualized and group-based applications.
-
-Compared to previous research, our findings align with the importance of capturing nonlinear dynamics in heart rate variability to understand cardiac responses during exercise [@gronwald2019effects].
-Similarly, previous studies have shown that dynamic fluctuations in RRi can serve as critical indicators of cardiorespiratory fitness, supporting the need for models to address the complexity of cardiovascular responses during physical stress [@mongin2022hrv].
-However, while many existing models focus primarily on linear metrics or aggregate HRV measures, our study provides a high-resolution analysis of RRi dynamics that enhances interpretability and application across diverse fitness levels and exercise intensities.
-
-Sobol analysis was selected for its capacity to handle non-linear, time-varying relationships, but it assumes parameter independence and demands substantial computational resources [@harenberg2017uncertainty; @cheng2017mixed; @herman2013method].
-This assumption may overlook interdependencies typical in biological systems, where parameters such as decay rates and recovery kinetics often interact.
-For future work, alternative sensitivity methods, such as Bayesian sensitivity analysis, could address interdependencies and improve interpretive accuracy in complex biological contexts.
-Additionally, real-time applications in resource-limited environments may benefit from computational optimizations, such as surrogate modeling or sparse-grid approximations, to reduce processing time while maintaining analytical depth [@bornn2010efficient; @xue2021bayesian].
-
-<!--- Validation and Practical Applications --->
-The practical application of this model extends to clinical monitoring and athletic training, but further validation is essential to confirm its robustness in real-world conditions.
-To support clinical decision-making, future studies should validate the model against established physiological benchmarks, such as VO~2~ max and lactate thresholds, facilitating its integration into routine health assessments.
-For instance, this model could be valuable in cardiovascular rehabilitation programs, where tracking real-time autonomic responses can help personalize exercise regimens, ensuring adequate recovery without overstressing the cardiovascular system.
-Furthermore, for at-risk populations, the model could assist in detecting autonomic irregularities indicative of early cardiovascular dysfunction [@thayer2010relationship].
-
-For athletic applications, the model has the potential to guide interval training, where it could help identify optimal recovery points between intense exercise bouts.
-This is particularly relevant given the findings of other research, which suggest that precise monitoring of heart rate dynamics can prevent overtraining and optimize performance [@eser2022acute].
-However, extending this model beyond controlled, single-bout settings requires further validation, particularly in dynamic environments where external factors like temperature, altitude, and psychological stress can influence RRi.
-Integrating the model with wearable devices capable of accurately capturing RRi is essential for field applications, enabling real-time monitoring and feedback [@lundstrom2023practices].
-
-<!--- Advances Over Existing Models --->
-Traditional models of RRi often rely on linear assumptions or simple exponential functions, which fail to capture the transient, non-stationary nature of RRi during exercise and recovery.
-Using logistic functions to model the decay and recovery phases, our non-linear approach accommodates gradual autonomic shifts more effectively than exponential functions, offering a physiologically relevant representation of RRi dynamics.
-While recent studies have developed non-linear models, many focus on HRV aggregates that summarize rather than directly model RRi changes [@molkkari2020online].
-By concentrating on RRi, our approach offers a high-resolution view of cardiac adaptation, making it particularly valuable for clinical and athletic contexts.
-Adapting model parameters to real-time changes in RRi enhances sensitivity in detecting subtle shifts in autonomic state, which aggregated HRV indices may obscure [@mabe2024associations].
-
-<!--- Limitations and Future Directions --->
-This model has limitations that warrant careful consideration.
-First, the uniform parameter sampling used in sensitivity analysis, though practical, may not fully capture individual variability in populations with extreme autonomic profiles or chronic health conditions.
-Incorporating empirical distributions or Bayesian priors based on population data could refine parameter estimates, enhancing individualization for diverse clinical populations, such as older adults or individuals with cardiovascular disease [@kristal1995heart].
-
-Moreover, an area for future research includes multi-session modeling to account for cumulative adaptations over repeated exercise sessions.
-Studies have shown that cumulative effects from multiple exercise bouts can introduce autonomic fatigue or adaptation [@castillo2021cardiac].
-Extending the model to track such longitudinal effects could help identify early signs of overtraining or autonomic exhaustion, providing a preventive tool for clinical and athletic populations.
-Additionally, considering environmental and psychological factors (i.e., such as temperature, stress, or sleep quality) would add robustness, ensuring that model predictions remain accurate across varied real-world contexts.
-
-# Conclusion
-
-This study presents a novel non-linear model for characterizing RRi dynamics during exercise and recovery, providing a more accurate and nuanced understanding of cardiac autonomic function compared to traditional HRV metrics.
-Our findings emphasize the important roles of baseline RRi and recovery proportion in reflecting autonomic re-engagement, enhancing the sensitivity of real-time assessments of cardiovascular responses to physical stress.
-This model advances computational physiology and holds significant promise for improving cardiovascular health monitoring and performance optimization.
-
-# Authors' Contributions
-
-Conceptualization, MC-A; Data curation, MC-A; Investigation, MC-A; Methodology, MC-A, NMD; Supervision, CN-E; Formal analysis, MC-A; Visualization, MC-A; Writing–original draft, MC-A, CN-E, [...]; Writing–review & editing, MC-A, CN-E, [...].
-All authors have read and agreed to the published version of the manuscript.
-
-# Funding
-
-This work was funded by ANID Proyecto Fondecyt Iniciación Nº11220116.
-
-# Data Availability Statement
-
-The data supporting the conclusions of this article will be available from the authors without reservation.
-
-# Conflict of Interest
-
-The authors declare that this research was conducted without any commercial or financial relationships that could be construed as potential conflicts of interest.
-
-# References
-
-<div id="refs"></div>
+[**Table 3**]{#tbl-sens-params}. Estimated $S_{\text{ind}}$ of model parameters. Each parameter's $S_{\text{ind}}$ reflects a uniform variation from the 95% CIs of the estimated parameter values.

### 2024-12-09 R/1-prior-rri-model.R

@@ -20,12 +20,12 @@ RRi_model <- function(t, alpha, beta, c, lambda, phi, tau, delta) {
 plot_post_dens <- function(posterior, x, xlab) {
   ggplot(posterior, aes(x = {{x}})) +
     ggdist::stat_slabinterval(
-      fill = "lightblue",
+      fill = "gray",
       trim = FALSE, density = "unbounded") +
     scale_y_continuous(expand = c(0.1,0,0.05,0), breaks = NULL, name = NULL) +
     scale_x_continuous(expand = c(0,0)) +
     labs(x = xlab) +
-    theme_classic(base_size = 12) +
+    theme_classic(base_line_size = 1/4, base_size = 12) +
     theme(axis.text = element_text(size = rel(.8)))
 }
 
@@ -93,12 +93,12 @@ pred <- predict(
 )
 
 fig_a <- ggplot(pred, aes(time_points, Estimate)) +
-  geom_ribbon(aes(ymin = Q2.5, ymax = Q97.5), fill = "#E3F2FDFF") +
-  geom_ribbon(aes(ymin = Q10, ymax = Q90), fill = "#BADEFAFF") +
-  geom_ribbon(aes(ymin = Q20, ymax = Q80), fill = "#90CAF8FF") +
-  geom_line(show.legend = FALSE, linewidth = 1, col = "blue4") +
+  geom_ribbon(aes(ymin = Q2.5, ymax = Q97.5), fill = "gray90") +
+  geom_ribbon(aes(ymin = Q10, ymax = Q90), fill = "gray75") +
+  geom_ribbon(aes(ymin = Q20, ymax = Q80), fill = "gray60") +
+  geom_line(show.legend = FALSE, linewidth = 1, col = "gray10") +
   labs(x = "Time (minutes)", y = "Predicted RRi") +
-  theme_classic(base_size = 12) +
+  theme_classic(base_line_size = 1/4, base_size = 12) +
   theme(axis.text = element_text(size = rel(.8)))
 
 posterior <- as_draws_df(m_prior_only) |> as.data.table()
@@ -112,13 +112,14 @@ fig_b6 <- plot_post_dens(posterior, b_tau_Intercept, expression(tau))
 fig_b7 <- plot_post_dens(posterior, b_delta_Intercept, expression(delta))
 
 figure <- ggpubr::ggarrange(
-  ggpubr::ggarrange(fig_b1, fig_b2, fig_b3, nrow = 1, align = "hv"),
[…]

### 2024-12-09 R/2-model-constituents.R

@@ -13,7 +13,7 @@ sim_data[, `:=`(alpha = 800,
                 `italic(f)[1](t)` = round(-400 / (1 + exp(log(1 - 0.7)*(t - 5)))),
                 `italic(f)[2](t)` = round((400 * 0.8) / (1 + exp(log(1 - 0.7)*(t - 7)))))]
 
-sim_data[, `RRi == alpha + italic(f)[1](t) + italic(f)[2](t)` := 
+sim_data[, `alpha + italic(f)[1](t) + italic(f)[2](t)` := 
            alpha + `italic(f)[1](t)` + `italic(f)[2](t)`]
 
 sim_data <- melt(sim_data, id.vars = "t", value.name = "RRi", variable.name = "Parameter")
@@ -22,17 +22,17 @@ sim_data <- melt(sim_data, id.vars = "t", value.name = "RRi", variable.name = "P
 # Visualize model constituents --------------------------------------------
 
 figure <- ggplot(sim_data, aes(t, RRi)) +
-  facet_wrap(~ Parameter, scales = "free_y", labeller = label_parsed) +
-  geom_line(aes(col = Parameter), show.legend = FALSE, linewidth = 1) +
-  geom_vline(xintercept = c(5,7), color = "gray50", linewidth = .25) +
+  facet_wrap(~ Parameter, scales = "free_y", labeller = label_parsed, nrow = 1) +
+  geom_line(col = "black", show.legend = FALSE, linewidth = 1/4) +
+  geom_vline(xintercept = c(5,7), color = "gray50", linewidth = 1/4, linetype = 2) +
   scale_y_continuous(expand = c(.05, 5), n.breaks = 4) +
-  ggsci::scale_color_observable() +
-  theme_classic(base_size = 12) +
+  scale_x_continuous(limits = c(0, 15)) +
+  theme_classic(base_line_size = 1/4, base_size = 12) +
   theme(strip.background = element_rect(fill = "grey90", colour = NA),
         axis.text = element_text(size = rel(.8))) +
-  labs(x = "Time (minutes)", y = "R-R intervals (ms)")
+  labs(x = "Time (minutes)", y = "RRi (ms)")
 
 ggsave("figures/model-constituents.pdf", figure, "pdf", 
-       width = 7, height = 5, units = "in")
+       width = 7, height = 4, units = "in")
 ggsave("figures/model-constituents.png", figure, "png", 
-       width = 7, height = 5, units = "in", dpi = 500)
+       width = 7, height = 4, units = "in", dpi = 500)

### 2024-12-09 R/3-sobol-sensitivity.R

@@ -29,12 +29,12 @@ rr_interval_model <- function(params, time) {
 # Generate random uniform samples for each parameter
 sobol_sample <- function(n_samples) {
   out <- data.frame(
-    alpha = runif(n_samples, min = 400, max = 1200), 
-    beta = runif(n_samples, min = -600, max = -200), 
-    c = runif(n_samples, min = 0.5, max = 1.5), 
-    lambda = runif(n_samples, min = -3.454, max = -1.151), 
-    phi = runif(n_samples, min = -1.806, max = -0.602),
-    tau = runif(n_samples, min = 2.5, max = 7.5), 
+    alpha = runif(n_samples, min = 850.57, max = 872.85), 
+    beta = runif(n_samples, min = -359.81, max = -330.97), 
+    c = runif(n_samples, min = 0.82, max = 0.86), 
+    lambda = runif(n_samples, min = -3.16, max = -2.94), 
+    phi = runif(n_samples, min = -2.71, max = -2.48),
+    tau = runif(n_samples, min = 3.05, max = 3.44), 
     delta = runif(n_samples, min = 1, max = 3)
   )
   
@@ -105,54 +105,56 @@ if (!file.exists("R/sobol-sensitivity.RDS")) { ## Omit computation if already ex
   sobol_results <- readRDS("R/sobol-sensitivity.RDS")
 }
 
+sobol_results[
+  j = list(estimate = median(mean), se = median(se)) |> lapply(round, 5), 
+  by = variable
+  ][, `:=`(
+    lower = round(estimate - se * qnorm(.975), 5),
+    upper = round(estimate + se * qnorm(.975), 5))
+  ] |> knitr::kable()
+
 
 # Plot results ------------------------------------------------------------
 
-## Function to plot function
-plot_sobol_sens <- function(vars, yexp, ylim, ind_cols) {
-  cols <- ggsci::pal_bmj()(7)
-  ggplot(sobol_results[time <= 12 & variable %in% vars], aes(time, mean)) +
-    facet_grid(cols = vars(variable),
-               labeller = label_parsed,
-               scales = "free_y") +
-    geom_line(aes(col = variable), show.legend = FALSE) +
-    geom_ribbon(aes(fill = variable, 
-                    ymin = mean - se * qnorm(.975),
-                    ymax = mean + se * qnorm(.975)), 
-                show.legend = FALSE, alpha = .3) +
-    geom_ribbon(aes(fill = variable, 
[…]

### 2024-12-09 R/4-rate-sensitivity.R

@@ -91,12 +91,12 @@ sim_data[, phi := factor(phi,
 
 figure <- ggplot(sim_data, aes(time, hrv)) +
   facet_wrap(~ phi, labeller = labeller(phi = label_parsed)) +
-  geom_line(aes(col = as.factor(x = lambda), group = id), linewidth = 0.7) +
-  geom_vline(xintercept = c(5,7), color = "gray50", linewidth = .25) +
+  geom_line(aes(col = as.factor(x = lambda), group = id), linewidth = 1/4) +
+  geom_vline(xintercept = c(5,7), color = "gray50", linewidth = 1/4, linetype = 2) +
   scale_y_continuous(breaks = c(500, 600, 700, 800)) +
   scale_alpha_manual(values = c(1, .85, .7, .55)) +
-  ggsci::scale_color_observable() +
-  theme_classic(base_size = 12) +
+  scale_color_grey(start = 0, end = 0.8) +
+  theme_classic(base_line_size = 1/4, base_size = 12) +
   theme(strip.background = element_rect(fill = "grey90", colour = NA),
         legend.position = "right",
         axis.text = element_text(size = rel(.8))) +

### 2024-12-09 R/5-2d-density-kernel.R

@@ -16,8 +16,8 @@ figure_a <- ggplot(rri_data, aes(time, rri_std)) +
   scale_x_continuous(expand = c(0,0), name = "Time (minutes)") +
   scale_y_continuous(expand = c(0,0), name = "Mean Standardized RRi",
                      breaks = -3:2) +
-  scale_fill_brewer() +
-  theme_classic(base_size = 14) +
+  scale_fill_grey(start = 1, end = 0) +
+  theme_classic(base_line_size = 1/4, base_size = 12) +
   theme(axis.text = element_text(size = rel(0.8)))
 
 figure_b <- 
@@ -31,23 +31,23 @@ figure_b <-
   ggplot(aes(value)) +
   facet_wrap(~ variable, nrow = 2, scales = "free_x") +
   ggdist::stat_slabinterval(
-    aes(fill = variable),
+    fill = "gray",
     normalize = "panels",
     adjust = 20, trim = FALSE,
     density = "unbounded",
     show.legend = FALSE
   ) +
-  ggsci::scale_fill_observable() +
   scale_y_continuous(breaks = NULL, name = NULL, expand = c(0.1,0,0.1,0)) +
   scale_x_continuous(name = "RRi (ms)", n.breaks = 4) +
-  theme_classic(base_size = 14) +
+  theme_classic(base_line_size = 1/4, base_size = 12) +
   theme(axis.text = element_text(size = rel(0.8)),
         plot.margin = margin(0.0825,.25,0.0825,.25, "in"),
         strip.background = element_rect(colour = NA, fill = "gray90"))
 
 figure <- ggpubr::ggarrange(
   figure_b, figure_a, ncol = 2, nrow = 1,
-  widths = c(1,2)
+  widths = c(1,2), labels = c("A.", "B."), font.label = list(size = 12),
+  hjust = -0.2, vjust = 2
 )
 
 ggsave("figures/2d-density-kernel.pdf", figure, "pdf", 

### 2024-12-09 R/7-meta-analysis.R

@@ -97,6 +97,6 @@ m_fits_df <- lapply(models, as_draws_df) |>
 m_fits_df[, Parameter := factor(
   x = Parameter, 
   levels = c("alpha", "beta", "c", "lambda", "phi", "tau", "delta", "sigma")
-)]
+)] 
 
 fwrite(m_fits_df, file = "models/group_level_posterior.csv")

### 2024-12-09 R/8-error-estimates.R

@@ -43,7 +43,7 @@ f_1 <- ggplot(error_data, aes(mape, rmse)) +
                      breaks = c(1,2,4,9,20)/100,
                      name = "MAPE (log-scale)", expand = c(0,0.05),
                      transform = "log10") +
-  theme_classic(base_size = 12) +
+  theme_classic(base_line_size = 1/4, base_size = 12) +
   theme(axis.text = element_text(size = rel(.8)))
 
 error_long <- melt.data.table(boot_error, id.vars = "mc_sample")
@@ -51,38 +51,37 @@ error_long[, variable := `levels<-`(variable, c("MAPE", "RMSE", "R^2"))]
 
 f_2a <- ggplot(error_long[variable == "MAPE"], aes(variable, value)) +
   facet_wrap(~ variable, scales = "free") +
-  ggdist::stat_halfeye(fill = "#4269D0FF") +
+  ggdist::stat_halfeye(fill = "gray") +
   scale_y_continuous(name = NULL, expand = c(0.1,0),
                      labels = scales::label_percent(),
                      n.breaks = 7) +
   scale_x_discrete(name = NULL, breaks = NULL, 
                    expand = c(0.1,0,0,0)) +
-  theme_classic(base_size = 12) +
+  theme_classic(base_line_size = 1/4, base_size = 12) +
   theme(strip.background = element_rect(fill = "grey90", colour = NA),
         axis.text = element_text(size = rel(.8)),
         plot.margin = margin(0.1,0.1,0.1,0.1,"in"))
 
 f_2b <- ggplot(error_long[variable == "RMSE"], aes(variable, value)) +
   facet_wrap(~ variable, scales = "free") +
-  ggdist::stat_halfeye(fill = "#EFB118FF") +
+  ggdist::stat_halfeye(fill = "gray") +
   scale_y_continuous(name = NULL, expand = c(0.1,0),
                      n.breaks = 6) +
   scale_x_discrete(name = NULL, breaks = NULL, 
                    expand = c(0.1,0,0,0)) +
-  theme_classic(base_size = 12) +
+  theme_classic(base_line_size = 1/4, base_size = 12) +
   theme(strip.background = element_rect(fill = "grey90", colour = NA),
         axis.text = element_text(size = rel(.8)),
         plot.margin = margin(0.1,0.1,0.1,0.1,"in"))
 
[…]

### 2024-12-09 R/9-group-level-estimates.R

@@ -19,15 +19,14 @@ m_fits_df[, Parameter := factor(Parameter,
 f_1 <- ggplot(m_fits_df, aes(Parameter, b_Intercept)) +
   facet_wrap(~ Parameter, nrow = 1, ncol = 8, scales = "free",
              labeller = label_parsed) +
-  ggdist::stat_halfeye(aes(fill = Parameter),
+  ggdist::stat_halfeye(fill = "gray",
                        show.legend = FALSE,
                        normalize = "panels",
                        trim = FALSE, density = "unbounded") +
   scale_x_discrete(expand = c(0.2,0,0.2,0), 
                    name = NULL, breaks = NULL) +
   scale_y_continuous(name = expression(symbol(E)~"["*theta*"]")) +
-  ggsci::scale_fill_observable() +
-  theme_classic(base_size = 12) +
+  theme_classic(base_line_size = 1/4, base_size = 12) +
   theme(strip.background = element_rect(fill = "grey90", colour = NA),
         axis.text = element_text(size = rel(.8)))
 
@@ -36,7 +35,7 @@ f_2 <-
   ggplot(aes(V1, Parameter)) +
   facet_wrap(~ Parameter, nrow = 1, ncol = 2, scales = "free",
              labeller = label_parsed) +
-  ggdist::stat_halfeye(aes(fill = Parameter),
+  ggdist::stat_halfeye(fill = "gray",
                        show.legend = FALSE,
                        normalize = "panels",
                        trim = FALSE, density = "unbounded") +
@@ -46,13 +45,14 @@ f_2 <-
                      labels = scales::label_percent(),
                      n.breaks = 4,
                      expand = c(0.2,0)) +
-  ggsci::scale_fill_observable() +
-  theme_classic(base_size = 12) +
+  theme_classic(base_line_size = 1/4, base_size = 12) +
   theme(strip.background = element_rect(fill = "grey90", colour = NA),
         axis.text = element_text(size = rel(.8)))
 
 figure <- ggpubr::ggarrange(
-  f_1, f_2, ncol = 1, nrow = 2, heights = c(7,3)
+  f_1, f_2, ncol = 1, nrow = 2, heights = c(7,3),
+  labels = c("A.", "B."), font.label = list(size = 12),
+  vjust = c(1.5, 2.5)
 )
 
 ggsave("figures/group-level-estimates.pdf", figure, "pdf", 

### 2024-12-09 R/example-data.R

@@ -1,0 +1,24 @@
+
+# Prepare workspace -------------------------------------------------------
+
+## Load libraries
+library(data.table)
+library(ggplot2)
+
+## Load data
+load("data/rri_data.RData")
+
+## Generate graphics
+figure <- ggplot(rri_data[id %in% c(1:3,7:9)], aes(time, rr_denoised)) +
+  facet_wrap(~ id, nrow = 2, ncol = 3) +
+  geom_line(lwd = 1/4, na.rm = TRUE) +
+  labs(x = "Time (min)", y = "RRi (ms)") +
+  theme_classic(base_line_size = 1/4, base_size = 12) +
+  theme(strip.background = element_blank(), strip.text = element_blank())
+  
+## Save the plot
+ggsave("figures/example-data.pdf", figure, "pdf", 
+       width = 7, height = 5, units = "in")
+ggsave("figures/example-data.png", figure, "png", 
+       width = 7, height = 5, units = "in", dpi = 500)
+  

### 2025-01-15 paper.qmd
state of the WD when submitted to SR

@@
 
-**Authors**: : Matías Castillo-Aguilar^1,2^, Diego Mabe-Castro^1,2^, David Medina^4,5^, Cristian Núñez-Espinosa^1,3^\*.
+**Authors**: Matías Castillo-Aguilar^1,2^, Diego Mabe-Castro^1,2^, David Medina^4,5^, Cristian Núñez-Espinosa^1,3^\*.
 
@@
 
-The aim of this work was to develop and validate a novel non-linear model to characterize RR interval (RRi) dynamics throughout a rest-exercise-recovery protocol, offering a more precise and physiologically relevant representation of cardiac autonomic responses than traditional HRV metrics or linear approaches.
+This work aimed to develop and validate a novel non-linear model to characterize RR interval (RRi) dynamics throughout a rest-exercise-recovery protocol, offering a more precise and physiologically relevant representation of cardiac autonomic responses than traditional HRV metrics or linear approaches.
 Using data from a cohort of 272 elderly participants, the model employs logistic functions to capture the non-stationary and transient nature of RRi dynamics, with parameter estimation achieved via Hamiltonian Monte Carlo.
@@
 Validation against real-world RRi data demonstrated robust model performance, accurately reflecting autonomic recovery and exercise-induced fluctuations.
-By advancing real-time cardiovascular assessments, this framework holds significant potential for clinical applications in rehabilitation and cardiovascular monitoring, as well as athletic contexts for optimizing performance and recovery.
-Taken together, these findings highlight the model's ability to provide precise, physiologically relevant assessments of autonomic function, paving the way for its use in personalized health monitoring and performance optimization across diverse populations.
+By advancing real-time cardiovascular assessments, this framework holds significant potential for clinical applications in rehabilitation and cardiovascular monitoring in athletic contexts to optimize performance and recovery.
+These findings highlight the model's ability to provide precise, physiologically relevant assessments of autonomic function, paving the way for its use in personalized health monitoring and performance optimization across diverse populations.
 
@@
 
-In this context, the study of R-R intervals (RRi), defined as the time intervals between heartbeats, and its link with exercise has emerged as an important research area, given its relevance to cardiovascular health, athletic performance, and physiological adaptation [@kristal1995heart; @thayer2010relationship; @dong2016role; @lundstrom2023practices].
-Hence, being able to analyze the temporal dynamics of RRi in response to exercise can provide valuable insights into the mechanisms by which the cardiovascular system adapts to physical stressors, such as exercise-induced fatigue and competition-related strain [@castillo2021cardiac; @eser2022exercise].
+In this context, the study of R-R intervals (RRi), defined as the time intervals between heartbeats and their link with exercise, has emerged as an important research area, given its relevance to cardiovascular health, athletic performance, and physiological adaptation [@kristal1995heart; @thayer2010relationship; @dong2016role; @lundstrom2023practices].
+Hence, analyzing the temporal dynamics of RRi in response to exercise can provide valuable insights into how the cardiovascular system adapts to physical stressors, such as exercise-induced fatigue and competition-related strain [@castillo2021cardiac; @eser2022exercise].
 
@@
 
-Modeling the RRi behavior has been traditionally approached by leveraging linear regression and time-series analysis [@lian2011simple].
+Modeling the RRi behavior has traditionally been approached by leveraging linear regression and time-series analysis [@lian2011simple].
 However, this oppose significant challenges, like oversimplifying overall exercise dynamics without capturing the intricacies of exercise-induced intricate transitions in RRi, especially under intense exertion and recovery phases [@gronwald2019effects].
 More recently, advanced non-linear approaches have been developed to address the limitations of linear methods like decision tree-based ensemble algorithms and convolutional neural networks [@wang2020automated;@lee2023real;@berrahou2024arrhythmia].
-However, many of the more advance alternatives fails to generate physiologically meaningful model parameters, without a direct link to biological processes [@gronwald2019non].
+However, many more advanced alternatives fail to generate physiologically meaningful model parameters without directly linking to biological processes [@gronwald2019non].
 
@@
 Such a model would need to offer a more physiologically relevant representation of the heart's behavior compared to the broader HRV indices commonly used in research [@bacopoulou2021entropy].
-This model would need to be complex enough to capture the non-linear, exercise-induced, cardiovascular dynamics.
-But, at the same time, simple enough to be able to provide practical and significant model parameters related to observed physiological processes.
+This model must be complex enough to capture the non-linear, exercise-induced cardiovascular dynamics.
+But, simultaneously, it has to be simple enough to provide practical and significant model parameters related to observed physiological processes.
 
@@
 The dataset consisted of 272 participants who underwent a validated exercise protocol encompassing rest, exercise, and recovery phases within a single, continuous measurement session [@castillo2023validity].
-Continuous heart rate data, including RRi, were collected using the Polar Team2 system (Polar®) application, capable of capturing dynamic fluctuations associated with varying exercise intensities and recovery.
+Continuous heart rate data, including RRi, were collected using the Polar Team2 system (Polar®) application, which is capable of capturing dynamic fluctuations associated with varying exercise intensities and recovery.
 
@@
 The gradient of the log-likelihood function for each parameter was computed during estimation using the brms R package (v2.21.0), which employs the Stan probabilistic programming language.
-Convergence of the HMC chains was assessed using standard diagnostics, including R-hat values, which were kept below 1.01 for all parameters [@vehtari2019rank], and effective sample sizes, which were targeted at a minimum of 1,000 for each parameter [@brms].
+Convergence of the HMC chains was assessed using standard diagnostics, including R-hat values, kept below 1.01 for all parameters [@vehtari2019rank], and effective sample sizes, targeted at a minimum of 1,000 for each parameter [@brms].
 Trace plots were inspected to confirm stable mixing.
@@
 
-To enhance the exploration of parameter space, we performed a two-stage analysis: An individual-level estimation of parameter values that were then used for the estimation population-level parameters.
+To enhance the exploration of parameter space, we performed a two-stage analysis: We assessed parameter values at the individual level, which we then used to estimate  population-level parameters.
 
@@
 $$
-y_{i,t} = \alpha_i + \frac{\beta_i}{1 + e^{\lambda_i \cdot (t - \tau_i)}} + \frac{-\beta_i \cdot c}{1 + e^{\phi_i \cdot (t - \tau_i - \delta_i)}} + \epsilon_{i,t}
+y_{i,t} = \alpha_i + \frac{\beta_i}{1 + e^{\lambda_i \cdot (t - \tau_i)}} + \frac{-c_i \cdot \beta_i}{1 + e^{\phi_i \cdot (t - \tau_i - \delta_i)}} + \epsilon_{i,t}
 $${#eq-indiv-level}
@@
 
-where $\mu_{k}$ is the group-level mean for parameter $k$, $b_{k,i}$ represents the subject-level random effect for subject $i$ on parameter $k$ and $\sigma_k^2$ is the residual variance for parameter $k$.
+where $\mu_{k}$ is the group-level mean for parameter $k$, $b_{k,i}$ represents the subject-level random effect for the subject $i$ on parameter $k$ and $\sigma_k^2$ is the residual variance for the parameter $k$.
 The subject-level effects $b_{k,i}$ were assumed to be distributed as $b_{k,i} \sim \mathcal{N}(0,\sigma^2)$, with $\sigma$ being the standard error of the subject-level effect.
 
-This hierarchical structure enables us to capture individual variability through subject-level random effects while estimating group-level effects across all parameters, thus providing estimates into subject and population-level model parameters.
+This hierarchical structure enables us to capture individual variability through subject-level random effects while estimating group-level effects across all parameters, thus providing estimates of subject- and population-level model parameters.
 
@@
 
-The primary performance metrics included R^2^, root mean square error (RMSE), and mean absolute percentage error (MAPE), estimated for each subject.
+The primary performance metrics, estimated for each subject, included R^2^, root mean square error (RMSE), and mean absolute percentage error (MAPE), estimated for each subject.
 Bootstrap resampling across each metric was performed to estimate the mean performance of the model and corresponding quantile-based 95% CI.
@@
 
-To assess the sensitivity of model parameters influencing RRi over time, we implemented a Sobol sensitivity analysis using Monte Carlo simulations.
+We implemented a Sobol sensitivity analysis using Monte Carlo simulations to assess the sensitivity of model parameters influencing RRi over time.
 Sobol index ($S_{\text{ind}}$) provide a measure of the proportion of the contribution of each parameter to the variance in RRi at each time point, and it was selected for its robustness in handling non-linear and non-monotonic relationships, which are intrinsic to RRi dynamics in response to exercise [@zhang2015sobol].
 
-To compute $S_{\text{ind}}$, 1000 Monte Carlo simulations were conducted, each involving 1000 randomly sampled parameter sets (1,000,000 model runs in total).
+To compute $S_{\text{ind}}$, 1000 Monte Carlo simulations were conducted, each involving 1000 randomly sampled parameter sets (1,000,000 model runs).
 For each set of parameters, RRi were calculated at each time point $t$ across a range from 0 to 20 minutes at intervals of 0.1 minutes.
 The 95% CI parameter values estimated from HMC-NUTS were then used as input ranges for $S_{\text{ind}}$ computation.
-Finally, the mean values of $S_{\text{ind}}$ over the 20-min time span for each model parameter were estimated and reported, next with their corresponding 95% CI using normal approximation based on estimated standard errors (SE).
+Finally, the mean values of $S_{\text{ind}}$ over the 20-minute time span for each model parameter were estimated and reported, with their corresponding 95% CI using normal approximation based on estimated standard errors (SE).
 
@@
 RRi dynamics in response to exercise tends to follow a U-shaped form.
-The initial decrease in RRi is associated with exercise onset, related to an increase in heart rate.
-After exercise cessation, an opposite increase in RRi is observed, associated with the cardiovascular recovery phase.
-In both cases, the drop and recovery phases occurs at different rates, some individuals experience a quick recovery in RRi after exercise, however, in some others this slope is less steep.
+The initial decrease in RRi is associated with exercise onset, related to an increased heart rate.
+After exercise cessation, an opposite increase in RRi is observed, which is associated with the cardiovascular recovery phase.
+In both cases, the drop and recovery phases occurs at different rates; some individuals experience a quick recovery in RRi after exercise; however, in some others, this slope is less steep.
 Additionally, the new baseline reached following exercise cessation is often below the RRi baseline before exercise.
 
-These hallmarks of RRi dynamics in response to exercise, highlights the complex and non-linear behavior of the cardiovascular response in the context of both rest and exercise conditions.
-An example RRi record data is shown in [figure 1](#fig-rri-example).
+These hallmarks of RRi dynamics in response to exercise, highlight the complex and non-linear behavior of the cardiovascular response in the context of both rest and exercise conditions.
+[Figure 1](#fig-rri-example) shows an example of RRi record data.
 
@@
 Initially, exponential and logarithmic functions were considered due to their simplicity and wide applicability in describing temporal changes.
-Exponential functions were hypothesized to capture the rapid initial adaptations of RRi post-exercise onset, while logarithmic functions were explored for their capacity to describe asymptotic behaviors observed in some physiological variables.
+Exponential functions were hypothesized to capture the rapid initial adaptations of RRi post-exercise onset.
+In contrast, logarithmic functions were explored for their capacity to describe asymptotic behaviors observed in some physiological variables.
 
 However, neither approach successfully reproduced the non-linear and bidirectional nature of the RRi fluctuations.
-Exponential functions, while effective at modeling monotonic decay or growth, could not account for the observed sigmoidal transitions.
+While effective at modeling monotonic decay or growth, exponential functions could not account for the observed sigmoidal transitions.
 Similarly, logarithmic functions, with their inherent monotonicity, failed to represent the plateauing behavior seen in real-world data.
 
-To address these limitations, we shifted to logistic functions, which inherently model sigmoidal transitions.
+We shifted to logistic functions to address these limitations, which inherently model sigmoidal transitions.
 Logistic functions introduce parameters for growth rate and inflection point, allowing for precise control over the shape and timing of the transition between dynamic states.
-By using two coupled logistic functions, one to represent the initial decrease in RRi and a second, inverted logistic function to describe the recovery phase, we achieved a model structure that could flexibly reproduce the observed non-linear variations.
+By using two coupled logistic functions, one to represent the initial decrease in RRi and a second inverted logistic function describing the recovery phase, we achieved a model structure that could flexibly reproduce the observed non-linear variations.
 
 This approach provided a biologically plausible representation, with parameters that directly correspond to identifiable physiological features, such as the rate of adaptation and recovery, the time to peak response, and the extent of deviation from baseline.
-The logistic function framework emerged as the optimal solution after systematic testing and evaluation against empirical data, ensuring that the model accurately captured both the qualitative and quantitative aspects of RRi dynamics.
+The logistic function framework emerged as the optimal solution after systematic testing and evaluation against empirical data, ensuring that the model accurately captured the qualitative and quantitative aspects of RRi dynamics.
 
@@
 
-In this equation, $a_1$ represents the asymptotic value approached by the logistic function, which can be either positive (indicating an increase) or negative (indicating a decrease).
+In this equation, $a_1$ represents the asymptotic value the logistic function approaches, which can be either positive (indicating an increase) or negative (indicating a decrease).
 For $f_1(t)$, this parameter is specified as $\beta$, indicating the absolute change in RRi at the onset of exercise.
@@
 
-The sample used to assess RRi dynamics consists of a group of 272 subjects selected from a local community of elderly individuals.
-The sample characteristics can be seen in [Table 1](#tbl-sample-characteristics)
+The sample used to assess RRi dynamics consists of 272 subjects selected from a local community of elderly individuals. The sample characteristics can be seen in [Table 1](#tbl-sample-characteristics)
 
@@
 
-Initial graphical exploration of RRi dynamics (see [Figure 3](#fig-2d-kernel-density)) indicates a clear drop in RRi around the 5-7 minutes mark, associated the exercise-induced cardiovascular stress.
+An initial graphical exploration of RRi dynamics (see [Figure 3](#fig-2d-kernel-density)) indicates a clear drop in RRi around the 5-7 minutes mark, associated with exercise-induced cardiovascular stress.
 However, greater variability across individuals in post-exercise recovery can be observed.
@@
 
-Given the observed parameters that reproduced the observed RRi patterns in exercise and rest conditions, priors were chosen based on physiological constraints and graphical visualization of standardized RRi data.
-Hence, ensuring identifiability of model parameters by constraining the parameter space to plausible values, in order to improve model convergence and parameter exploration.
+Given the parameters that reproduced the observed RRi patterns in exercise and rest conditions, priors were chosen based on physiological constraints and the graphical visualization of standardized RRi data.
+Hence, ensuring the identifiability of model parameters by constraining the parameter space to plausible values to improve model convergence and parameter exploration.
 The prior distributions were defined as follows:
@@
 
-Once subject-level RRi data was fitted using the proposed in @eq-indiv-level, a population-parameter value was estimated based on the proposed group-level methodology.
+Once subject-level RRi data was fitted using the proposed model in @eq-indiv-level, a population-parameter value was estimated based on the proposed group-level methodology.
 The estimated parameter values can be seen in [Table 2](#tbl-parameters)
@@
 
-Relative performance metrics, estimated through bootstrapped resampling, suggest that the model tends to deviate a 3.4% (CI~95%~[3.06, 3.81]) from the observed RRi data.
+Estimated through bootstrapped resampling, relative performance metrics suggest that the model tends to deviate a 3.4% (CI~95%~[3.06, 3.81]) from the observed RRi data.
 This is equivalent to a 32.6 ms in the RRi scale (CI~95%~[30.01, 35.77]).
@@
 Residuals analysis showed that the estimated partial correlation function (ACF) from the model residuals indicates a correlation among non-explained errors greater than 0.1 up to the 5th lag.
-However, the partial ACF is significant (CI-wise) and strictly positive or negative up to the second lag.
+However, the partial ACF is significant (CI-wise) and strictly positive or negative until the second lag.
 Correlations among model residuals against other time indices remained insignificant (see [Figure 6](#fig-error-rates)).
@@
 Conversely, the rate parameters for the initial decay during exercise, $\lambda$, and the recovery post-exercise, $\phi$, show lower sensitivity, suggesting that they are not primary sources of variation in predicted RRi trajectories when assessed in isolation.
-The results of the sensitivity analysis can be seen in [Table 3](#tbl-sens-params).
+The results of the sensitivity analysis are in [Table 3](#tbl-sens-params).
 
@@
 
-To the best of our knowledge, this study represents the first attempt to develop a non-linear model specifically designed to continuously capture RR interval dynamics across a complete rest-exercise-rest protocol.
+To the best of our knowledge, this study represents the first attempt to develop a non-linear model specifically designed to capture RRi dynamics across a complete rest-exercise-rest protocol continuously.
 Previous studies have either focused on aggregate HRV indices or utilized simplified linear or exponential models, which are insufficient to describe the complex, non-stationary transitions observed during and after exercise [@molkkari2020online].
 By employing a combination of logistic functions, our model uniquely accounts for the gradual shifts in autonomic regulation denoted by RRi dynamics, offering a detailed and physiologically relevant representation of cardiac dynamics.
-This continuous modeling framework allows for the integration of both exercise-induced RRi decline and post-exercise recovery within a single unified structure, bridging a critical gap in the current literature.
-Such an approach not only advances our understanding of cardiovascular responses but also opens new avenues for real-time monitoring and intervention in both clinical and athletic settings.
+This continuous modeling framework integrates exercise-induced RRi decline and post-exercise recovery within a single unified structure, bridging a critical gap in the current literature.
+Such an approach advances our understanding of cardiovascular responses and opens new avenues for real-time monitoring and intervention in both clinical and athletic settings.
 
-The proposed model demonstrates a precise capacity to reproduce RRi dynamics, with its combination of logistic functions, capturing the key transitions of cardiac response, the initial decline during exercise and the subsequent recovery.
+The proposed model demonstrates a precise capacity to reproduce RRi dynamics.
+With its combination of logistic functions, it captures the key transitions of cardiac response, the initial decline during exercise, and the subsequent recovery.
 This design accommodates the inherent non-linearity and non-stationarity of RRi dynamics, overcoming the limitations of linear models and exponential functions commonly used in prior studies [@silva2019exponential; @mongin2022decrease].
 
-Compared to previous research, our findings align with efforts made on capturing nonlinear dynamics in HRV to understand cardiac responses during exercise [@gronwald2019effects].
-Similarly, previous studies have shown that dynamic fluctuations in RRi can serve as critical indicators of cardiorespiratory fitness, supporting the need for models to address the complexity of cardiovascular responses during physical stress [@mongin2022hrv].
+Compared to previous research, our findings align with efforts to capture nonlinear dynamics in HRV to understand cardiac responses during exercise [@gronwald2019effects].
+Similarly, previous studies have shown that dynamic fluctuations in RRi can serve as critical indicators of cardiorespiratory fitness.
+This supports the need for models to address the complexity of cardiovascular responses during physical stress [@mongin2022hrv].
 However, while many existing models focus primarily on linear metrics or aggregate HRV measures, our study provides a high-resolution analysis of RRi dynamics that enhances interpretability and application across diverse fitness levels and exercise intensities.
@@
 Unlike prior research that aggregates HRV measures or applies simple decay models, our approach directly models RRi changes, offering richer physiological insight.
-For instance, commonly utilized exponential decay models for post-exercise recovery are used, however they fail to incorporate the transition dynamics observed during exercise itself [@sahoo2024unanticipated].
-By integrating both exercise and recovery phases, our model provides a more comprehensive view of autonomic regulation.
+For instance, commonly utilized exponential decay models for post-exercise recovery are used but fail to incorporate the transition dynamics observed during exercise itself [@sahoo2024unanticipated].
+By integrating exercise and recovery phases, our model provides a more comprehensive view of autonomic regulation.
 
 Moreover, the sensitivity of parameters such as $\lambda$ (decay rate) and $(\phi$ (recovery rate) was found to be relatively low, suggesting that the model is robust to variability in these rates while remaining sensitive to key physiological parameters ($\alpha$ and $c$).
-This robustness makes it suitable for both individualized monitoring and population-level analyses, offering versatility in its application across different use cases.
+This robustness makes it suitable for individualized monitoring and population-level analyses, offering versatility in its application across different use cases.
 
@@
 
-This model demonstrates significant potential for practical applications in both clinical and athletic settings.
+This model demonstrates significant potential for practical applications in clinical and athletic settings.
 In clinical contexts, it could aid in tailoring cardiovascular rehabilitation protocols by monitoring autonomic recovery in real-time, ensuring safe and effective exercise regimens for at-risk populations [@carrasco2024hrv].
-This aligns with previous research, which highlight the importance of individualizing rehabilitation programs to optimize recovery [@takahashi2020signs; @rizvi2023enhancing; @carrasco2024hrv].
+This aligns with previous research, which highlights the importance of individualizing rehabilitation programs to optimize recovery [@takahashi2020signs; @rizvi2023enhancing; @carrasco2024hrv].
 
-In athletic settings, the model could guide training strategies, particularly for interval training, where determining optimal recovery periods is crucial.
+The model could guide training strategies in athletic settings, particularly for interval training, where determining optimal recovery periods is crucial.
 Similar findings suggest that precise monitoring of RRi dynamics can prevent overtraining and enhance performance [@hebisz2022heart; @nuuttila2024monitoring].
@@
 
-Additionally, environmental and psychological factors, such as temperature, stress, or sleep quality, were not explicitly considered in this study.
+Additionally, this study did not explicitly consider environmental and psychological factors, such as temperature, stress, or sleep quality.
 Future work could integrate these variables into the model, enhancing its robustness and applicability across varied real-world scenarios.
@@
 In summary, this study presents a novel non-linear model for RRi dynamics, capturing the complex and transient autonomic responses during rest-exercise-recovery protocols, overcoming the limitations of traditional autonomic metrics.
-By identifying baseline RRi and recovery proportion as the dominant contributors to variability, the model emphasizes their critical roles in reflecting autonomic regulation and resilience.
-Validated across a cohort of elderly participants, the model demonstrates robust performance in real-time cardiovascular assessments, offering significant potential for clinical applications such as rehabilitation and monitoring in at-risk populations, as well as athletic contexts like fatigue management and performance optimization.
+The model emphasizes their critical roles in reflecting autonomic regulation and resilience by identifying baseline RRi and recovery proportion as the dominant contributors to variability.
+Validated across a cohort of elderly participants, the model demonstrates robust performance in real-time cardiovascular assessments, offering significant potential for clinical applications such as rehabilitation and monitoring in at-risk populations and athletic contexts like fatigue management and performance optimization.
 While the model's applicability is currently constrained by its focus on elderly individuals, future validation in younger cohorts and under diverse environmental conditions will enhance its generalizability and utility.
@@
 
-Conceptualization, MC-A; Data curation, MC-A; Investigation, MC-A; Methodology, MC-A, NMD; Supervision, CN-E; Formal analysis, MC-A; Visualization, MC-A; Writing–original draft, MC-A, CN-E, DM-C; Writing–review & editing, MC-A, CN-E, DM.
+Conceptualization, MC-A; Data curation, MC-A; Investigation, MC-A, DM-C; Methodology, MC-A, NMD; Supervision, CN-E; Formal analysis, MC-A; Visualization, MC-A; Writing–original draft, MC-A, CN-E, DM-C; Writing–review & editing, MC-A, CN-E, DM.
 All authors have read and agreed to the published version of the manuscript.
@@
 [**Figure 1**]{#fig-rri-example}.
-Example data of RRi recordings of 6 different subjects over a 20 minute rest-exercise-recovery protocol in a sample of elderly individuals.
-The subject-level data shows the inter-individual variability of RRi dynamics in response exercised-induced cardiovascular stress, with similar behavior and trajectories of recovery over time.
+Example data of RRi recordings of 6 subjects over a 20-minute rest-exercise-recovery protocol in a sample of elderly individuals.
+The subject-level data shows the inter-individual variability of RRi dynamics in response to exercised-induced cardiovascular stress, with similar behavior and recovery trajectories over time.
 
@@
 [**Figure 3**]{#fig-2d-kernel-density}.
-(A) Mean and SD from each of the subject's RRi recordings, used for the standardization process.
+(A) Mean and SD from each subject's RRi recordings were used for the standardization process.
 (B) 2D kernel density of standardized RRi dynamics over time from a sample of individuals subjected to the rest-exercise-rest protocol.
@@
 Shaded areas represent 95%, 80%, and 60% quantile CI, offering insight into expected physiological variability across parameters.
-(B) Prior distributions and 95% CI, used to generate prior predictions, based on physiological constraints and graphical visualization of standardized RRi data.
+(B) Prior distributions and 95% CI were used to generate prior predictions based on physiological constraints and graphical visualization of standardized RRi data.
 
@@
 [**Figure 5**]{#fig-group-level-estimates}.
-(A) Posterior probability distributions of the expectation for each population-parameter estimates ($E[\theta]$) with quantile-based 95% CI.
+(A) Posterior probability distributions of the expectation for each population-parameter estimate ($E[\theta]$) with quantile-based 95% CI.
 (B) Transformed rate parameters into a percentage scale using the $1 - \exp(\theta)$ transformation.
@@
 Individual-level performance metrics.
-(A) Bootstrapped MAPE and RMSE, as metrics of relative and absolute model deviance from observed RRi.
+(A) Bootstrapped MAPE and RMSE are metrics of relative and absolute model deviance from observed RRi.
 (B) Individual-level estimates of model performance and the relationship between them.
@@
 [**Table 1**]{#tbl-sample-characteristics}.
-Sample characteristics from which, continuous RRi monitoring data was collected during a rest-exercise-rest protocol.
+Sample characteristics from which continuous RRi monitoring data was collected during a rest-exercise-rest protocol.
 Data is presented as Mean ± standard deviation (SD).
@@
 |:---------|-----------:|-------:|--------:|--------:|
-|$\alpha$  |     0.56808| 0.01813|  0.53255|  0.60361|
-|$\beta$   |     0.02378| 0.00111|  0.02160|  0.02596|
-|$c$       |     0.21406| 0.00914|  0.19615|  0.23197|
-|$\lambda$ |     0.00045| 0.00002|  0.00041|  0.00049|
-|$\phi$    |     0.00012| 0.00001|  0.00010|  0.00014|
-|$\tau$    |     0.04031| 0.00160|  0.03717|  0.04345|
-|$\delta$  |     0.15387| 0.00291|  0.14817|  0.15957|
+|$\alpha$  |     0.61329| 0.01756|  0.57887|  0.64771|
+|$\beta$   |     0.06651| 0.00286|  0.06090|  0.07212|
+|$c$       |     0.18939| 0.00815|  0.17342|  0.20536|
+|$\lambda$ |     0.00147| 0.00007|  0.00133|  0.00161|
+|$\phi$    |     0.00160| 0.00008|  0.00144|  0.00176|
+|$\tau$    |     0.04982| 0.00172|  0.04645|  0.05319|
+|$\delta$  |     0.07896| 0.00239|  0.07428|  0.08364|
 

### 2025-01-15 supplementary.qmd
state of the WD when submitted to SR

---
title: "Supplementary Material"
format: 
  docx:
    reference-doc: "ref.docx"
bibliography: "misc/ref.bib"
csl: "misc/scientific-reports.csl"
editor_options: 
  chunk_output_type: console
execute: 
  echo: false
  warning: false
  error: false
---

**Title**: Enhancing Cardiovascular Monitoring: A Non-Linear Approach to RR Interval Dynamics in Exercise and Recovery.

**Authors**: Matías Castillo-Aguilar^1,2^, Diego Mabe-Castro^1,2^, David Medina^4,5^, Cristian Núñez-Espinosa^1,3^\*.

::: {custom-style="Afiliation"}
^1^ Centro Asistencial Docente e Investigación (CADI-UMAG), University of Magallanes, Punta Arenas, Chile.

^2^ Kinesiology Department, University of Magallanes, Punta Arenas, Chile.

^3^ School of Medicine, University of Magallanes, Punta Arenas, Chile.

^4^ Departamento de Ingeniería en Computación, Universidad de Magallanes, Punta Arenas, Chile.

^5^ Centre for Biotechnology and Bioengineering, CeBiB, Universidad de Chile, Santiago, Chile.

:::

### \*Correspondence:

Cristian Núñez-Espinosa, School of Medicine, University of Magallanes, Punta Arenas, Chile. Centro Asistencial de Docencia e Investigación CADI-UMAG, Chile. e-mail: [cristian.nunez\@umag.cl](mailto:cristian.nunez@umag.cl). Address: Avenida Bulnes 01855, Box 113-D. Phone: +56 61 2201411.

{{< pagebreak >}}

# Methods

## RRi Data Processing

This section describes the detailed methodology used to process the RR interval (RRi) data, highlighting the decisions made and the rationale behind each step. The processing pipeline was implemented in **R** using key libraries, including `data.table` and `signal`, and was designed to ensure the integrity and standardization of RRi data for subsequent modeling.

### Data Import and Initial Filtering

RRi data were collected from `.txt` files stored in the `data-raw/hrv_data` directory. Each file contained a series of RR intervals, measured in milliseconds. To ensure data quality, only RRi values within a physiologically plausible range (400–1200 ms) were retained. This range was selected based on established thresholds for human heart rates during moderate exercise and recovery phases. RR intervals outside this range were likely artifacts or measurement errors and were excluded to prevent skewing the analysis.

The cumulative sum of valid RRi values was calculated to determine the time elapsed (in minutes) for each recording. Data exceeding 20 minutes of cumulative time were excluded to standardize the evaluation period, focusing on the typical duration of rest-exercise-recovery protocols. This step ensured that subsequent analyses could directly compare individuals without temporal biases.

The filtered RRi data were consolidated into a single data table, with additional metadata columns, including the file identifier and the corresponding recording ID.

```r
rri_data <- lapply(files, \(i) { 
  rr_interval <- as.double(x = readLines(i))
  not_na <- !is.na(rr_interval)
  rr_interval <- rr_interval[not_na & between(rr_interval, 400, 1200)]
  time <- cumsum(rr_interval)/60000
  data.table(rr_interval = rr_interval[time <= 20],
             time = time[time <= 20],
             file = gsub("data-raw/", "", i, fixed = TRUE))
}) |> rbindlist(idcol = "id")
```

### Exclusion of Incomplete Data

Incomplete recordings, defined as those with less than 10 minutes of cumulative time, were excluded. This threshold was chosen to maintain sufficient data length for robust analysis of the rest-exercise-recovery phases. Files corresponding to incomplete recordings were moved to a separate directory (`data-raw/hrv_data/unused`) for archival purposes. This step ensured that only high-quality data were used for subsequent processing, minimizing the impact of noise and artifacts.

```r
id_to_keep <- rri_data[, max(time) > 10, id][V1 == TRUE, id]
rri_data <- rri_data[id %in% id_to_keep]
```

### Signal Denoising and Filtering

To mitigate noise in the RRi signals, a 10 Hz low-pass Butterworth filter was applied. This filter effectively smooths the data while preserving the key features of the RRi dynamics [@salsekar2012filtering; @fedotov2016selection], such as the onset of exercise-induced changes and post-exercise recovery. A zero-phase filtering technique (`filtfilt`) was used to prevent phase distortion, which is critical for preserving the temporal structure of the RRi signal [@salsekar2012filtering; @fedotov2016selection].

```r
denoise <- function(x, w = 0.1, n = 3) {
  signal::filtfilt(filt = signal::butter(n, w), x)
}
rri_data[, rr_denoised := denoise(rr_interval), id]
```

Edge artifacts introduced by the filtering process were mitigated by trimming the first and last five data points of each signal. This ensures that only the central, unperturbed portions of the signal are retained for analysis.

```r
trim <- function(x, abs) {
  x_len <- length(x)
  ind <- c(1:abs, (x_len - abs):x_len)
  x[ind] <- NA
  x
}
rri_data[, rr_denoised := trim(rr_denoised, abs = 5), id]
```

### Visual Inspection for Manual Noise Removal

Visual inspection of denoised signals was conducted to identify recordings dominated by noise. This manual step allowed the identification of spurious data that automated filtering might miss, such as signals with irregular, non-physiological patterns. Any identified noisy recordings were excluded from further analysis and archived for transparency.

### Standardization of RRi Data

To facilitate comparison across individuals, RRi values were standardized within each recording. The mean and standard deviation of denoised RRi data were calculated for each individual. Standardized RRi values (z-scores) were then computed to ensure that inter-individual variability in baseline RRi values did not confound subsequent analyses.

```r
rri_data[, rri_mean := mean(rr_denoised, na.rm = TRUE), by = id]
rri_data[, rri_sd := sd(rr_denoised, na.rm = TRUE), by = id]
rri_data[, rri_std := (rr_denoised - rri_mean) / rri_sd, by = id]
```

## Sobol Sensitivity Analysis

This section provides a detailed explanation of the Sobol sensitivity analysis conducted to evaluate the influence of model parameters on RRi dynamics. The analysis quantifies the variance contributions of individual parameters, enabling insight into their relative importance in driving model output.

### Sampling Parameter Space

To evaluate the sensitivity of model parameters, random uniform samples were drawn for each parameter from the 95% CI estimated physiological ranges. These ranges were derived from posterior parameter estimates and reflect plausible RRi dynamics during exercise and recovery compatible with the observed data. For each Monte Carlo simulation, 1,000 parameter sets were sampled to ensure robust coverage of the parameter space.

```r
sobol_sample <- function(n_samples) {
  out <- data.frame(
    alpha = runif(n_samples, min = 850.57, max = 872.85), 
    beta = runif(n_samples, min = -359.81, max = -330.97), 
    c = runif(n_samples, min = 0.82, max = 0.86), 
    lambda = runif(n_samples, min = -3.16, max = -2.94), 
    phi = runif(n_samples, min = -2.71, max = -2.48),
    tau = runif(n_samples, min = 6.61, max = 6.81), 
    delta = runif(n_samples, min = 3.05, max = 3.44)
  )
}
```

### Sobol Variance Decomposition

The Sobol analysis decomposes the total variance of the model output into contributions from individual parameters. For each parameter, a perturbed dataset was generated by varying the selected parameter while averaging others. The variance of the perturbed model output was compared to the total variance to compute the first-order Sobol indices:

$$
S_i = \frac{\text{Var}_{\text{perturbed}}(\text{RRi})}{\text{Var}_{\text{total}}(\text{RRi})}.
$$

These indices provide a measure of each parameter's individual contribution to output variability.

```r
sobol_variance <- function(model_function, samples, time_points) {
  total_output <- sapply(1:nrow(samples), function(i) {
    model_function(as.numeric(samples[i, ]), time_points)
  })

  total_variance <- apply(total_output, 1, var)

  sobol_indices <- matrix(0, nrow = length(time_points), ncol = ncol(samples),
                          dimnames = list(NULL, names(samples)))

  for (param in names(samples)) {
    perturbed_output <- sapply(1:nrow(samples), function(i) {
      modified_params <- as.numeric(samples[i, ])
      modified_params[names(samples) != param] <- apply(samples[names(samples) != param], 2, mean)
      model_function(modified_params, time_points)
    })

    variance_contrib <- apply(perturbed_output, 1, var)
    sobol_indices[, param] <- variance_contrib / total_variance
  }

  return(data.frame(time = time_points, sobol_indices))
}
```

### Monte Carlo Simulations and Bootstrapping

To ensure the robustness of sensitivity estimates, 1,000 Monte Carlo simulations were performed, each with 1,000 parameter sets. Bootstrapping was applied to compute confidence intervals for Sobol indices, quantifying the uncertainty of parameter sensitivity across simulations.

```r
n_boot <- 1000
boot <- vector(mode = "list", length = n_boot)

for (i in seq_len(n_boot)) {
  boot[[i]] <- sobol_variance(
    model_function = rr_interval_model, 
    samples = sobol_sample(n_samples = 1000), 
    time_points = seq(0, 20, by = 0.1)
  )
}
```

### Results and Interpretation

The code that generated the summary table of Sobol indices is provided below:

```r
sobol_results[
  j = list(estimate = mean(mean), se = mean(se)) |> lapply(round, 5), 
  by = variable
  ][, `:=`(
    lower = round(estimate - se * qnorm(.975), 5),
    upper = round(estimate + se * qnorm(.975), 5))
  ] |> knitr::kable()
```

## Parameter Estimation

### Individual-Level Parameter Estimation

This section describes the step-by-step methodology for performing individual-level parameter estimation in the hierarchical modeling of RRi dynamics during exercise and recovery. The process utilizes Bayesian posterior sampling with pre-defined priors to estimate subject-specific parameters, enabling detailed insights into personalized cardiovascular responses.

#### Loading and Preparing the Data

RRi data for individual subjects were filtered to remove missing or invalid standardized RRi values ($\text{rri\_std}$). The input data were structured such that each subject's measurements were indexed by their unique identifier ($\text{id}$), ensuring traceability during individual-level analysis.

```r
m_data <- rri_data[!is.na(rri_std)]
```

#### Model Initialization and Posterior Sampling

The prior-only model, previously defined and fitted, served as the initial framework for sampling individual-level posteriors. Unique subject IDs were identified, and a loop was employed to sequentially fit the model for each subject's data. For computational efficiency, the pipeline checks if a posterior sample already exists for a given subject to avoid redundant computation.

```r
list_models[[i]] <- update(
  object = m_model,
  newdata = m_data[id == i],
  seed = 1234,
  iter = 10000, warmup = 5000,
  chains = 5, cores = 5,
  control = list(adapt_delta = 0.99, max_treedepth = 50),
  sample_prior = "no",
  file = file
)
```

The posterior samples from all subjects were combined into a single data table, facilitating downstream analysis and parameter transformation.

#### Transforming Parameters to Original Scale

Parameters such as $\alpha$ (baseline RRi) and $\beta$ (exercise-induced decline) were transformed back to their original scale using the subject-specific mean and standard deviation of RRi. This step ensures that parameter estimates are physiologically interpretable and directly comparable across individuals.

```r
df_models[, b_alpha_Intercept := b_alpha_Intercept * rri_sd + rri_mean, id]
df_models[, b_beta_Intercept := b_beta_Intercept * rri_sd, id]
df_models[, sigma := sigma * rri_sd, id]
```

#### Summarizing Parameter Estimates

For each subject, the posterior means and standard deviations of parameters ($\alpha, \beta, c, \lambda, \phi, \tau, \delta, \sigma$) were computed to create summary statistics. These metrics provide central tendencies and variability for each estimated parameter.

```r
mu_data <- df_models[, lapply(.SD, mean), id, .SDcols = b_alpha_Intercept:sigma]
se_data <- df_models[, lapply(.SD, sd), id, .SDcols = b_alpha_Intercept:sigma]
```

#### Validation Against Observed Data

Estimated parameters were used to simulate RRi dynamics for each subject. Observed RRi data were compared to the simulated predictions to calculate error metrics, including mean absolute percentage error (MAPE), root mean squared error (RMSE), and $R^2$. These metrics quantify the accuracy of the individual-level model in capturing the observed RRi dynamics.

```r
long_data[, error := rr_denoised - pred_rr, id]
long_data[, mape := mean(abs(error/rr_denoised), na.rm = T), id]
long_data[, rmse := sqrt(mean((error^2), na.rm = TRUE)), id]
long_data[, rsquared := 1-sum(error^2)/sum((rr_denoised-rr_bar)^2), id]
```

### Group-Level Parameter Estimation

This section describes the methodology used to perform group-level parameter estimation for modeling RR interval (RRi) dynamics across subjects. The analysis aggregates individual-level posterior estimates into a hierarchical framework, leveraging Bayesian modeling to derive population-level parameters while preserving individual variability.

#### Data and Model Setup

The dataset for this stage consists of posterior means and standard errors ($\mu$ and SE) for parameters estimated at the individual level. Each parameter is modeled hierarchically using a univariate Bayesian framework, allowing for random effects across subjects.

The response variable for each parameter-specific model ($Y_{i,j}$) is expressed as:

$$
Y_{i,j} \sim \mathcal{N}(\mu_j + b_{i,j}, \sigma_j^2),
$$

where $\mu_j$ represents the fixed effect (population mean), $b_{i,j}$ the random effect for subject $i$, and $\sigma_j^2$ the residual variance. This hierarchical specification captures variability both between and within subjects.

#### Model Specification

Each parameter ($\alpha, \beta, c, \lambda, \phi, \tau, \delta, \sigma$) is modeled separately, using the posterior means and standard errors from the individual-level analysis as inputs. For each parameter-specific model:

(1) The response variable ($\mu_{\text{param}}$) represents the posterior mean. (2) The standard error ($\text{SE}_{\text{param}}$) is incorporated using the `se()` function in **brms**. (3) Subject-level random effects ($(1 | \text{id})$) capture between-subject variability.

```r
m_alpha_model <- bf(alphaMu | se(alphaSE, sigma = TRUE) ~ 1 + (1 | id))
m_beta_model <- bf(betaMu | se(betaSE, sigma = TRUE) ~ 1 + (1 | id))
m_c_model <- bf(cMu | se(cSE, sigma = TRUE) ~ 1 + (1 | id))
m_lambda_model <- bf(lambdaMu | se(lambdaSE, sigma = TRUE) ~ 1 + (1 | id))
m_phi_model <- bf(phiMu | se(phiSE, sigma = TRUE) ~ 1 + (1 | id))
m_tau_model <- bf(tauMu | se(tauSE, sigma = TRUE) ~ 1 + (1 | id))
m_delta_model <- bf(deltaMu | se(deltaSE, sigma = TRUE) ~ 1 + (1 | id))
m_sigma_model <- bf(sigmaMu | se(sigmaSE, sigma = TRUE) ~ 1 + (1 | id))
```

#### Prior Specification

Informative priors were specified for each parameter to reflect physiological constraints and ensure robust estimation. For instance, $\alpha$ (baseline RRi) was given a prior centered around 850 ms with a standard deviation of 10 ms, while $\beta$ (exercise-induced decline) was constrained to negative values. These priors guide the model while allowing sufficient flexibility for parameter estimation.

#### Fitting and Post-Processing

The models were fit using **brms**, leveraging Hamiltonian Monte Carlo for efficient sampling. Each model was initialized with 10,000 iterations, including 5,000 warmup steps, and run with five chains to ensure convergence.

```r
fit_model <- function(model, priors) {
  brms::brm(formula = model,
            data = id_data,
            prior = priors,
            family = gaussian(),
            chains = 5, cores = 5, seed = 1234,
            iter = 10000, warmup = 5000,
            control = list(adapt_delta = .99,
                           max_treedepth = 50),
            file = paste0("models/group_level/",
                          deparse(substitute(model)),
                          ".RDS")
  )
}
```

Posterior distributions for all parameters were extracted and compiled into a single dataset for summary and visualization.

```r
m_fits_df <- lapply(models, as_draws_df) |> 
  rbindlist(idcol = "Parameter")
```

The final dataset includes posterior estimates for population-level means ($\mu_j$), subject-level variability ($b_{i,j}$), and residual variance ($\sigma_j^2$) for each parameter. These estimates are stored in `group_level_posterior.csv`, providing a comprehensive view of group-level dynamics and inter-individual differences.

# References

<div id="refs"></div>

### 2025-01-15 tables.qmd
state of the WD when submitted to SR

--- 2025-01-15 paper.qmd
@@
 ---
-
-**Title**: Enhancing Cardiovascular Monitoring: A Non-Linear Approach to RR Interval Dynamics in Exercise and Recovery.
-
-**Authors**: Matías Castillo-Aguilar^1,2^, Diego Mabe-Castro^1,2^, David Medina^4,5^, Cristian Núñez-Espinosa^1,3^\*.
-
-::: {custom-style="Afiliation"}
-^1^ Centro Asistencial Docente e Investigación (CADI-UMAG), University of Magallanes, Punta Arenas, Chile.
-
-^2^ Kinesiology Department, University of Magallanes, Punta Arenas, Chile.
-
-^3^ School of Medicine, University of Magallanes, Punta Arenas, Chile.
-
-^4^ Departamento de Ingeniería en Computación, Universidad de Magallanes, Punta Arenas, Chile.
-
-^5^ Centre for Biotechnology and Bioengineering, CeBiB, Universidad de Chile, Santiago, Chile.
-
-:::
-
-### \*Correspondence:
-
-Cristian Núñez-Espinosa, School of Medicine, University of Magallanes, Punta Arenas, Chile.
-Centro Asistencial de Docencia e Investigación CADI-UMAG, Chile.
-e-mail: [cristian.nunez\@umag.cl](mailto:cristian.nunez@umag.cl).
-Address: Avenida Bulnes 01855, Box 113-D.
-Phone: +56 61 2201411.
-
-{{< pagebreak >}}
-
-## Abstract
-
-This work aimed to develop and validate a novel non-linear model to characterize RR interval (RRi) dynamics throughout a rest-exercise-recovery protocol, offering a more precise and physiologically relevant representation of cardiac autonomic responses than traditional HRV metrics or linear approaches.
-Using data from a cohort of 272 elderly participants, the model employs logistic functions to capture the non-stationary and transient nature of RRi dynamics, with parameter estimation achieved via Hamiltonian Monte Carlo.
-Sobol sensitivity analysis identified baseline RRi ($\alpha$) and recovery proportion ($c$) as the primary drivers of variability, underscoring their critical roles in autonomic regulation and resilience.
-Validation against real-world RRi data demonstrated robust model performance, accurately reflecting autonomic recovery and exercise-induced fluctuations.
-By advancing real-time cardiovascular assessments, this framework holds significant potential for clinical applications in rehabilitation and cardiovascular monitoring in athletic contexts to optimize performance and recovery.
-These findings highlight the model's ability to provide precise, physiologically relevant assessments of autonomic function, paving the way for its use in personalized health monitoring and performance optimization across diverse populations.
-
-**Keywords**: Heart Rate Variability, Exercise Physiology, Autonomic Nervous System, Cardiovascular System, Models, Theoretical, Logistic Models.
-
-{{< pagebreak >}}
-
-# Introduction
-
-Current research has extensively examined the mechanisms underlying cardiac autonomic dynamics in response to exercise and their links to health-related quality of life and cardiovascular disease risk [@eser2022exercise;@castillo2023validity;@mabe2024associations].
-
-In this context, the study of R-R intervals (RRi), defined as the time intervals between heartbeats and their link with exercise, has emerged as an important research area, given its relevance to cardiovascular health, athletic performance, and physiological adaptation [@kristal1995heart; @thayer2010relationship; @dong2016role; @lundstrom2023practices].
-Hence, analyzing the temporal dynamics of RRi in response to exercise can provide valuable insights into how the cardiovascular system adapts to physical stressors, such as exercise-induced fatigue and competition-related strain [@castillo2021cardiac; @eser2022exercise].
-
-Unlike heart rate variability (HRV), which aggregates autonomic responses over time, RRi analysis provides a more granular, direct view of cardiac electrical activity during or immediately following exercise, particularly in older adults [@mongin2022hrv; @castillo2023validity; @mabe2024associations].
-
-Thus, understanding RRi fluctuations in response to exercise is particularly relevant during dynamic exercise periods, where the autonomic nervous system (ANS) shifts between parasympathetic withdrawal and sympathetic activation [@boettger2010heart].
-Therefore, directly modeling RRi dynamics, rather than relying on broader HRV metrics, allows for a direct assessment of physiological markers of autonomic adaptation to stress [@hautala2003short].
-This approach is valuable for identifying recovery patterns and understanding cardiovascular reactivity across individuals with various fitness levels [@mongin2022hrv].
-
-Modeling the RRi behavior has traditionally been approached by leveraging linear regression and time-series analysis [@lian2011simple].
-However, this oppose significant challenges, like oversimplifying overall exercise dynamics without capturing the intricacies of exercise-induced intricate transitions in RRi, especially under intense exertion and recovery phases [@gronwald2019effects].
-More recently, advanced non-linear approaches have been developed to address the limitations of linear methods like decision tree-based ensemble algorithms and convolutional neural networks [@wang2020automated;@lee2023real;@berrahou2024arrhythmia].
-However, many more advanced alternatives fail to generate physiologically meaningful model parameters without directly linking to biological processes [@gronwald2019non].
-
-Alongside this line of inquiry, recent studies have begun exploring non-linear models for RRi dynamics, recognizing their potential to capture the complexity of cardiovascular response to exercise [@da2024post].
-Exponential decay models, for example, have been proposed to describe RRi recovery [@sahoo2024unanticipated], while logistic functions have been used to model the gradual return to baseline after high-intensity exercise [@molkkari2020online; @gronwald2019non].
-These models offer advantages over traditional HRV metrics by providing a more detailed understanding of the cardiovascular system's response to exercise [@wu2003nonlinear].
-However, despite these advancements, few models are specifically designed to capture real-time RRi fluctuations, and even fewer consider physiologically meaningful parameters that allow modeling individual variability across a myriad of exercise dimensions [@kanniainen2023estimation].
-
-Considering the current state of the art, there is a clear need to develop a non-linear model that accurately represents RRi's non-linear transitions during exercise and recovery, where the estimation of clinically significant parameters is compelling.
-Such a model would need to offer a more physiologically relevant representation of the heart's behavior compared to the broader HRV indices commonly used in research [@bacopoulou2021entropy].
-This model must be complex enough to capture the non-linear, exercise-induced cardiovascular dynamics.
-But, simultaneously, it has to be simple enough to provide practical and significant model parameters related to observed physiological processes.
-
-Hence, the primary objective of this paper is to present a novel non-linear model that characterizes continuous RRi transitions from rest to exercise and recovery.
-This model is designed to capture the non-linear changes in RRi, providing meaningful parameters that can enhance our understanding of the physiological processes underlying cardiovascular adaptation to exercise.
-
-# Methods
-
-## Data collection and preprocessing
-
-To further assess the performance of the proposed model, real-world RRi data were analyzed in addition to the synthetic data generated through simulation.
-This dataset was derived from a cohort participating in the FONDECYT Project No.
-11220116, funded by the Chilean National Association of Research and Development (ANID).
-Ethical approval was granted by the Ethics Committee of the University of Chile (ACTA No.
-029-18/05/2022) and the Ethics Committee of the University of Magallanes (No.
-008/SH/2022).
-
-The dataset consisted of 272 participants who underwent a validated exercise protocol encompassing rest, exercise, and recovery phases within a single, continuous measurement session [@castillo2023validity].
-Continuous heart rate data, including RRi, were collected using the Polar Team2 system (Polar®) application, which is capable of capturing dynamic fluctuations associated with varying exercise intensities and recovery.
-
-Preprocessing steps were conducted to remove artifacts and ectopic heartbeats, with less than 3% of data excluded following established guidelines [@malik1996heart].
-The preprocessed RRi data were then aggregated into time intervals to facilitate analysis, allowing the examination of acute exercise responses and post-exercise recovery patterns.
-
-This real-world dataset provided a critical context for validating the model's predictive capability against observed physiological responses, offering a robust foundation for understanding RRi dynamics under physical activity conditions.
-
-## Parameter Estimation
-
-Parameter estimation was performed using Hamiltonian Monte Carlo (HMC) with the No-U-Turn Sampler (NUTS) to explore the parameter space [@hoffman2014no].
-The parameters $\alpha$, $\beta$, $c$, $\lambda$, $\phi$, $\tau$, and $\delta$ were estimated by sampling from the posterior distribution, which was constructed from observed RRi data and model predictions.
-
-The gradient of the log-likelihood function for each parameter was computed during estimation using the brms R package (v2.21.0), which employs the Stan probabilistic programming language.
-Convergence of the HMC chains was assessed using standard diagnostics, including R-hat values, kept below 1.01 for all parameters [@vehtari2019rank], and effective sample sizes, targeted at a minimum of 1,000 for each parameter [@brms].
-Trace plots were inspected to confirm stable mixing.
-These diagnostics collectively confirmed reliable posterior estimates for each parameter.
-
-The fitting process utilized five Markov Chain Monte Carlo (MCMC) chains, each consisting of 10,000 iterations with a burn-in period of 5,000 iterations, resulting in 25,000 post-warmup samples.
-
-To enhance the exploration of parameter space, we performed a two-stage analysis: We assessed parameter values at the individual level, which we then used to estimate  population-level parameters.
-
-### Individual-level analysis
-
-Firstly, each subject's RRi data $\text{RRi}_{i,t}$ was standardized against his mean $\bar{\text{RRi}_i}$ and standard deviation $S_{\text{RRi}_i}$ to improve convergence and exploration of the posterior distribution.
-The standardized RRi data $y_{i,t}$ for each time point $t$ was computed as:
-
-$$
-y_{i,t} = \frac{\text{RRi}_{i,t} - \bar{\text{RRi}_i}}{S_{\text{RRi}_i}}
-$${#eq-ind-std}
-This standardization allowed the model to focus on relative changes in RRi dynamics, independent of individual baseline differences.
-
-The model for each subject $i$ was then specified in terms of standardized RRi data $y_{i,t}$:
-
-$$
-y_{i,t} = \alpha_i + \frac{\beta_i}{1 + e^{\lambda_i \cdot (t - \tau_i)}} + \frac{-c_i \cdot \beta_i}{1 + e^{\phi_i \cdot (t - \tau_i - \delta_i)}} + \epsilon_{i,t}
-$${#eq-indiv-level}
-where $\alpha_i$, $\beta_i$, $c_i$, $\lambda_i$, $\phi_i$, $\tau_i$, $\delta_i$ are the individual-specific model parameters and $\epsilon_{i,t} \sim \mathcal{N}(0, \sigma^2)$ is the residual error term at each time point $t$.
-
-Afterwards, we transformed the estimated $\alpha$ and $\beta$ parameters back to the original RRi scale, ensuring a physiologically meaningful interpretation.
-The transformation for each subject $i$ is given by:
-
-$$
-\begin{aligned}
-\alpha_i^{\text{RRi}} &= \alpha_i \cdot S_{\text{RRi}_i} + \bar{\text{RRi}}_i \\
-\beta_i^{\text{RRi}} &= \beta_i \cdot S_{\text{RRi}_i}
-\end{aligned}
-$${#eq-ind-unstd}
-
-### Group-level analysis
-
-After obtaining the posterior distribution for each subject's parameters, each parameter's mean ($\theta^{\text{ obs}}$) and standard error ($\epsilon$) were calculated.
-These estimates were then used as input data to create a univariate hierarchical model, capturing variability at both the subject and group levels.
-The modeling process is described as follows:
-
-For each subject $i$, we estimated an interdependent stochastic process in which the true parameter $\theta_{k,i}$, with $k \in \{\alpha, \beta, c, \lambda, \phi, \tau, \delta\}$ with their corresponding standard error $\epsilon_{k,i}$ was used to model the observed parameter $\theta^{\text{ obs}}_{k,i}$ as:
-
-$$
-\theta^{\text{ obs}}_{k,i} \sim \mathcal{N}(\theta_{k,i}, \epsilon_{k,i})
-$${#eq-gp-observed}
-
-Then, the true parameter $\theta_{k,i}$ was further modeled as:
-
-$$
-\theta_{k,i} \sim \mathcal{N}(\mu_{k} + b_{k,i}, \sigma_k^2)
-$${#eq-gp-true}
-
-where $\mu_{k}$ is the group-level mean for parameter $k$, $b_{k,i}$ represents the subject-level random effect for the subject $i$ on parameter $k$ and $\sigma_k^2$ is the residual variance for the parameter $k$.
-The subject-level effects $b_{k,i}$ were assumed to be distributed as $b_{k,i} \sim \mathcal{N}(0,\sigma^2)$, with $\sigma$ being the standard error of the subject-level effect.
-
-This hierarchical structure enables us to capture individual variability through subject-level random effects while estimating group-level effects across all parameters, thus providing estimates of subject- and population-level model parameters.
-
-## Model Performance
-
-The primary performance metrics, estimated for each subject, included R^2^, root mean square error (RMSE), and mean absolute percentage error (MAPE), estimated for each subject.
-Bootstrap resampling across each metric was performed to estimate the mean performance of the model and corresponding quantile-based 95% CI.
-
-Also, residual analysis were conducted to evaluate the model's accuracy in capturing RRi dynamics.
-Residuals were defined as the difference between observed and predicted RRi values.
-These residuals were analyzed for temporal structure and partial autocorrelation to ensure that no systematic patterns remained in the errors.
-This indicates that the model has sufficiently captured the underlying dynamics of the RRi response to exercise.
-
-## Model parameters sensitivity
-
-Once a model that described RRi behavior in response to exercise was obtained, an assessment of the proportion of the variance explained by each model parameter was then computed.
-
-We implemented a Sobol sensitivity analysis using Monte Carlo simulations to assess the sensitivity of model parameters influencing RRi over time.
-Sobol index ($S_{\text{ind}}$) provide a measure of the proportion of the contribution of each parameter to the variance in RRi at each time point, and it was selected for its robustness in handling non-linear and non-monotonic relationships, which are intrinsic to RRi dynamics in response to exercise [@zhang2015sobol].
-
-To compute $S_{\text{ind}}$, 1000 Monte Carlo simulations were conducted, each involving 1000 randomly sampled parameter sets (1,000,000 model runs).
-For each set of parameters, RRi were calculated at each time point $t$ across a range from 0 to 20 minutes at intervals of 0.1 minutes.
-The 95% CI parameter values estimated from HMC-NUTS were then used as input ranges for $S_{\text{ind}}$ computation.
-Finally, the mean values of $S_{\text{ind}}$ over the 20-minute time span for each model parameter were estimated and reported, with their corresponding 95% CI using normal approximation based on estimated standard errors (SE).
-
-# Results
-
-## Problem characterization
-
-RRi dynamics in response to exercise tends to follow a U-shaped form.
-The initial decrease in RRi is associated with exercise onset, related to an increased heart rate.
-After exercise cessation, an opposite increase in RRi is observed, which is associated with the cardiovascular recovery phase.
-In both cases, the drop and recovery phases occurs at different rates; some individuals experience a quick recovery in RRi after exercise; however, in some others, this slope is less steep.
-Additionally, the new baseline reached following exercise cessation is often below the RRi baseline before exercise.
-
-These hallmarks of RRi dynamics in response to exercise, highlight the complex and non-linear behavior of the cardiovascular response in the context of both rest and exercise conditions.
-[Figure 1](#fig-rri-example) shows an example of RRi record data.
-
-<!---- FIGURE 1 ---->
-
-## Model construction
-
-The process of deriving the final equation for modeling RRi fluctuations was guided by an iterative exploration of mathematical functions capable of capturing the observed dynamics.
-Initially, exponential and logarithmic functions were considered due to their simplicity and wide applicability in describing temporal changes.
-Exponential functions were hypothesized to capture the rapid initial adaptations of RRi post-exercise onset.
-In contrast, logarithmic functions were explored for their capacity to describe asymptotic behaviors observed in some physiological variables.
-
-However, neither approach successfully reproduced the non-linear and bidirectional nature of the RRi fluctuations.
-While effective at modeling monotonic decay or growth, exponential functions could not account for the observed sigmoidal transitions.
-Similarly, logarithmic functions, with their inherent monotonicity, failed to represent the plateauing behavior seen in real-world data.
-
-We shifted to logistic functions to address these limitations, which inherently model sigmoidal transitions.
-Logistic functions introduce parameters for growth rate and inflection point, allowing for precise control over the shape and timing of the transition between dynamic states.
-By using two coupled logistic functions, one to represent the initial decrease in RRi and a second inverted logistic function describing the recovery phase, we achieved a model structure that could flexibly reproduce the observed non-linear variations.
-
-This approach provided a biologically plausible representation, with parameters that directly correspond to identifiable physiological features, such as the rate of adaptation and recovery, the time to peak response, and the extent of deviation from baseline.
-The logistic function framework emerged as the optimal solution after systematic testing and evaluation against empirical data, ensuring that the model accurately captured the qualitative and quantitative aspects of RRi dynamics.
-
-The mathematical model proposed to characterize the RRi response to exercise and recovery is defined by @eq-main-model.
-
-$$
-\begin{aligned}
-\text{RRi}(t) = \alpha + \frac{\beta}{1 + e^{\lambda(t - \tau)}} + \frac{-c \cdot \beta}{1 + e^{\phi(t - \tau - \delta)}}
-\end{aligned}
-$${#eq-main-model}
-
-This model includes two logistic functions representing the RRi dynamics across exercise and recovery phases.
-The first logistic term models the decrease in RRi during exercise, where the parameter $\beta$ denotes the magnitude of this decline.
-The rate of decrease is governed by $\lambda$, while $\tau$ represents the onset of the RRi decrease or the time the physiological shift begins.
-
-The second logistic term accounts for RRi recovery post-exercise.
-Here, $c$ scales the magnitude of recovery relative to the initial decline represented by $\beta$, capturing the proportion of the decline regained during recovery.
-The rate at which RRi returns to baseline is controlled by $\phi$, and $\delta$ indicates the lag following the cessation of exercise, marking the beginning of recovery, respectively.
-
-Additionally, the dynamics of RRi in response to physical exertion can be represented as a linear combination of a baseline RRi $\alpha$ and two logistic functions denoted as $f_1(t)$ and $f_2(t)$.
-The function $f_1(t)$ models the initial decay in RRi following the initiation of exercise while $f_2(t)$ characterizes the recovery phase after exercise cessation.
-
-Essentially, the fundamental structure of both logistic functions can be expressed as:
-
-$$
-\begin{aligned}
-f(t) = \frac{a_1}{1 + e^{a_2(t - a_3)}}
-\end{aligned}
-$${#eq-logistic-structure}
-
-In this equation, $a_1$ represents the asymptotic value the logistic function approaches, which can be either positive (indicating an increase) or negative (indicating a decrease).
-For $f_1(t)$, this parameter is specified as $\beta$, indicating the absolute change in RRi at the onset of exercise.
-In contrast, for $f_2(t)$, $a_1$ is parametrized as $-c \cdot \beta$, where $c$ denotes the proportion of change relative to the initial drop indicated by $\beta$.
-This parametrization ensures that, after the initial decline, the second logistic function facilitates the return of RRi toward the baseline value $\alpha$.
-
-```{r}
-#| eval: false
-ln_to_perc <- function(ln) 1 - exp(ln)
-perc_to_ln <- function(perc) log(1 - perc)
-
-perc_to_ln(0.9) |> # log(1 - 0.9) ==> Log-scale
-  ln_to_perc()     # 1 - exp(-2.302585) ==> Percentage-change
-```
-
-The parameter $a_2$ defines the rate at which the specified increase or decrease occurs.
-This rate parameter is expressed on a logarithmic scale; to convert it to a percentage change per unit of time, it can be scaled as $1 - \exp(a_2)$.
-<!-- For instance, a 90% decrease per unit time corresponds to $a_2 = \log(1 - 0.9)$, resulting in an approximate value of -2.302585.
--->
-
-The parameter $a_3$ serves as an activation threshold, causing the value within the exponential function, and consequently, the value in the denominator, to increase significantly until reaching $a_3$.
-Beyond this point, the denominator approaches 1, allowing the logistic function to attain the asymptotic level determined by the numerator.
-[Figure 2](#fig-linear-constituents) illustrates the behavior of the model constituents.
-
-<!---- FIGURE 2 ---->
-
-## Sample characteristics
-
-The sample used to assess RRi dynamics consists of 272 subjects selected from a local community of elderly individuals. The sample characteristics can be seen in [Table 1](#tbl-sample-characteristics)
-
-<!---- TABLE 1 ---->
-
-An initial graphical exploration of RRi dynamics (see [Figure 3](#fig-2d-kernel-density)) indicates a clear drop in RRi around the 5-7 minutes mark, associated with exercise-induced cardiovascular stress.
-However, greater variability across individuals in post-exercise recovery can be observed.
-
-<!---- FIGURE 3 ---->
-
-## Parameter estimation
-
-### Priors
-
-Given the parameters that reproduced the observed RRi patterns in exercise and rest conditions, priors were chosen based on physiological constraints and the graphical visualization of standardized RRi data.
-Hence, ensuring the identifiability of model parameters by constraining the parameter space to plausible values to improve model convergence and parameter exploration.
-The prior distributions were defined as follows:
-
-$$
-\begin{aligned} 
-\alpha &\sim \mathcal{N}(1, 0.5) \\
-\beta &\sim \mathcal{N}(-2.5, 0.5) \text{ with } \beta \leq 0 \\
-c &\sim \mathcal{N}(0.8, 0.2) \text{ with } c \geq 0 \\
-\lambda &\sim \mathcal{N}(-2, 0.5) \text{ with } \lambda \leq 0 \\
-\phi &\sim \mathcal{N}(-2, 0.5) \text{ with } \phi \leq 0 \\
-\tau &\sim \mathcal{N}(5, 0.5) \text{ with } \tau \geq 0 \\
-\delta &\sim \mathcal{N}(5, 0.5) \text{ with } \delta \geq 0
-\end{aligned}
-$${#eq-pe-priors}
-
-Simulated standardized RRi dynamics based on prior parameter distributions are shown in [Figure 4](#fig-prior-sim).
-
-<!---- FIGURE 4 ---->
-
-### Parameter estimates
-
-Once subject-level RRi data was fitted using the proposed model in @eq-indiv-level, a population-parameter value was estimated based on the proposed group-level methodology.
-The estimated parameter values can be seen in [Table 2](#tbl-parameters)
-
-<!---- TABLE 2 ---->
-
-In [Figure 5](#fig-group-level-estimates), the model parameter's posterior distribution can be observed.
-
-<!---- FIGURE 5 ---->
-
-## Model evaluation
-
-### Model performance
-
-Estimated through bootstrapped resampling, relative performance metrics suggest that the model tends to deviate a 3.4% (CI~95%~[3.06, 3.81]) from the observed RRi data.
-This is equivalent to a 32.6 ms in the RRi scale (CI~95%~[30.01, 35.77]).
-Additionally, the bootstrapped R^2^ indicates that the model explains 0.868 (CI~95%~[0.834, 0.895]) of the total variance observed in RRi.
-
-Residuals analysis showed that the estimated partial correlation function (ACF) from the model residuals indicates a correlation among non-explained errors greater than 0.1 up to the 5th lag.
-However, the partial ACF is significant (CI-wise) and strictly positive or negative until the second lag.
-Correlations among model residuals against other time indices remained insignificant (see [Figure 6](#fig-error-rates)).
-
-<!---- FIGURE 6 ---->
-
-### Model parameters sensitivity
-
-Sobol sensitivity analysis reveals that the parameter $\alpha$ exerts the most substantial influence on the model's output, followed by parameters $c$ and $\delta$.
-In contrast, parameters $\beta$, $\lambda$, and $\phi$ demonstrate relatively minor effects, with some values crossing zero, indicating negligible influence within the tested parameter ranges.
-
-Individual perturbation of each parameter highlighted that RRi dynamics are sensitive to the baseline RRi parameter, $\alpha$.
-Conversely, the rate parameters for the initial decay during exercise, $\lambda$, and the recovery post-exercise, $\phi$, show lower sensitivity, suggesting that they are not primary sources of variation in predicted RRi trajectories when assessed in isolation.
-The results of the sensitivity analysis are in [Table 3](#tbl-sens-params).
-
-<!---- TABLE 3 ---->
-
-# Discussion
-
-To the best of our knowledge, this study represents the first attempt to develop a non-linear model specifically designed to capture RRi dynamics across a complete rest-exercise-rest protocol continuously.
-Previous studies have either focused on aggregate HRV indices or utilized simplified linear or exponential models, which are insufficient to describe the complex, non-stationary transitions observed during and after exercise [@molkkari2020online].
-By employing a combination of logistic functions, our model uniquely accounts for the gradual shifts in autonomic regulation denoted by RRi dynamics, offering a detailed and physiologically relevant representation of cardiac dynamics.
-This continuous modeling framework integrates exercise-induced RRi decline and post-exercise recovery within a single unified structure, bridging a critical gap in the current literature.
-Such an approach advances our understanding of cardiovascular responses and opens new avenues for real-time monitoring and intervention in both clinical and athletic settings.
-
-The proposed model demonstrates a precise capacity to reproduce RRi dynamics.
-With its combination of logistic functions, it captures the key transitions of cardiac response, the initial decline during exercise, and the subsequent recovery.
-This design accommodates the inherent non-linearity and non-stationarity of RRi dynamics, overcoming the limitations of linear models and exponential functions commonly used in prior studies [@silva2019exponential; @mongin2022decrease].
-
-Compared to previous research, our findings align with efforts to capture nonlinear dynamics in HRV to understand cardiac responses during exercise [@gronwald2019effects].
-Similarly, previous studies have shown that dynamic fluctuations in RRi can serve as critical indicators of cardiorespiratory fitness.
-This supports the need for models to address the complexity of cardiovascular responses during physical stress [@mongin2022hrv].
-However, while many existing models focus primarily on linear metrics or aggregate HRV measures, our study provides a high-resolution analysis of RRi dynamics that enhances interpretability and application across diverse fitness levels and exercise intensities.
-
-The flexibility of the logistic components allows for physiologically interpretable parameters, such as baseline RRi ($\alpha$) and recovery proportion ($c$), which directly correlate with intrinsic cardiac function and autonomic recovery capacity, respectively.
-These features position the model as a robust framework for investigating the cardiovascular system's dynamic adaptation to physical stressors.
-For example, prior studies have highlighted the inadequacy of linear HRV metrics in capturing transient autonomic shifts [@gregoire2023autonomic]; our results align with this critique, demonstrating the advantages of modeling RRi directly.
-
-Unlike prior research that aggregates HRV measures or applies simple decay models, our approach directly models RRi changes, offering richer physiological insight.
-For instance, commonly utilized exponential decay models for post-exercise recovery are used but fail to incorporate the transition dynamics observed during exercise itself [@sahoo2024unanticipated].
-By integrating exercise and recovery phases, our model provides a more comprehensive view of autonomic regulation.
-
-Moreover, the sensitivity of parameters such as $\lambda$ (decay rate) and $(\phi$ (recovery rate) was found to be relatively low, suggesting that the model is robust to variability in these rates while remaining sensitive to key physiological parameters ($\alpha$ and $c$).
-This robustness makes it suitable for individualized monitoring and population-level analyses, offering versatility in its application across different use cases.
-
-The Sobol sensitivity analysis revealed that baseline RRi ($\alpha$) and recovery proportion ($c$) are the primary drivers of model output variance, emphasizing their physiological importance.
-These findings are consistent with prior research, which identified baseline cardiac function as a determinant of cardiovascular health and recovery proportion as a marker of autonomic resilience [@hautala2003short].
-
-However, the Sobol method assumes parameter independence, which may overlook interactions that are common in biological systems [@harenberg2017uncertainty; @cheng2017mixed; @herman2013method].
-For example, the interplay between $\lambda$ and $c$, which dictates the rate and magnitude of recovery, is likely critical but remains unexplored in the current framework.
-Future studies could explore Bayesian sensitivity analysis or variance decomposition methods that account for parameter interdependence [@bornn2010efficient; @xue2021bayesian].
-
-This model demonstrates significant potential for practical applications in clinical and athletic settings.
-In clinical contexts, it could aid in tailoring cardiovascular rehabilitation protocols by monitoring autonomic recovery in real-time, ensuring safe and effective exercise regimens for at-risk populations [@carrasco2024hrv].
-This aligns with previous research, which highlights the importance of individualizing rehabilitation programs to optimize recovery [@takahashi2020signs; @rizvi2023enhancing; @carrasco2024hrv].
-
-The model could guide training strategies in athletic settings, particularly for interval training, where determining optimal recovery periods is crucial.
-Similar findings suggest that precise monitoring of RRi dynamics can prevent overtraining and enhance performance [@hebisz2022heart; @nuuttila2024monitoring].
-The model's ability to integrate real-time data from wearable devices further enhances its applicability in dynamic, uncontrolled environments, enabling field-based monitoring and feedback [@zimatore2020recurrence].
-
-While the model presents substantial advances, it has limitations that warrant consideration.
-First, the assumption of uniform parameter sampling in sensitivity analysis, while practical, may not fully capture the variability observed in populations with extreme autonomic profiles [@kristal1995heart].
-Empirical distributions or Bayesian priors, could improve parameter estimation and enhance the model's applicability to diverse populations [@xue2021bayesian].
-
-Another limitation lies in the demographic composition of the sample, which consisted exclusively of elderly individuals.
-While this population provides valuable insights into age-specific cardiovascular dynamics, the findings may not fully generalize to younger populations, whose autonomic responses to exercise and recovery differ significantly due to higher baseline vagal tone, greater cardiac plasticity, and distinct metabolic profiles [@castillo2021cardiac; @garavaglia2021effect].
-Previous studies have demonstrated that younger individuals exhibit faster autonomic recovery and greater adaptability during physical exertion compared to older populations, suggesting that the parameter estimates derived from this model may vary across age groups [@choi2020declining; @garavaglia2021effect].
-Future research should validate the model in more diverse cohorts, including younger adults and athletes, to ensure broader applicability and to explore potential age-dependent modifications of the model's parameters.
-This would enhance its utility in clinical and athletic contexts, where age diversity is a critical factor [@choi2020declining; @garavaglia2021effect].
-
-Additionally, this study did not explicitly consider environmental and psychological factors, such as temperature, stress, or sleep quality.
-Future work could integrate these variables into the model, enhancing its robustness and applicability across varied real-world scenarios.
-This aligns with calls for more integrative modeling approaches in cardiovascular research [@takahashi2020signs; @hebisz2022heart; @nuuttila2024monitoring].
-
-# Conclusion
-
-In summary, this study presents a novel non-linear model for RRi dynamics, capturing the complex and transient autonomic responses during rest-exercise-recovery protocols, overcoming the limitations of traditional autonomic metrics.
-The model emphasizes their critical roles in reflecting autonomic regulation and resilience by identifying baseline RRi and recovery proportion as the dominant contributors to variability.
-Validated across a cohort of elderly participants, the model demonstrates robust performance in real-time cardiovascular assessments, offering significant potential for clinical applications such as rehabilitation and monitoring in at-risk populations and athletic contexts like fatigue management and performance optimization.
-While the model's applicability is currently constrained by its focus on elderly individuals, future validation in younger cohorts and under diverse environmental conditions will enhance its generalizability and utility.
-This work establishes a foundational framework for advancing personalized cardiovascular health monitoring and intervention.
-
-# References
-
-<div id="refs"></div>
-
-# Author Contributions
-
-Conceptualization, MC-A; Data curation, MC-A; Investigation, MC-A, DM-C; Methodology, MC-A, NMD; Supervision, CN-E; Formal analysis, MC-A; Visualization, MC-A; Writing–original draft, MC-A, CN-E, DM-C; Writing–review & editing, MC-A, CN-E, DM.
-All authors have read and agreed to the published version of the manuscript.
-
-# Data Availability Statement
-
-The data supporting the conclusions of this article will be available from the authors without reservation.
-
-# Competing Interests Statement
-
-The authors declare that this research was conducted without any commercial or financial relationships that could be construed as potential conflicts of interest.
-
-# Funding
-
-This work was funded by ANID Proyecto Fondecyt Iniciación Nº11220116.
-
-{{< pagebreak >}}
-
-# Figures
-
-![](figures/example-data.png)
-[**Figure 1**]{#fig-rri-example}.
-Example data of RRi recordings of 6 subjects over a 20-minute rest-exercise-recovery protocol in a sample of elderly individuals.
-The subject-level data shows the inter-individual variability of RRi dynamics in response to exercised-induced cardiovascular stress, with similar behavior and recovery trajectories over time.
-
-{{< pagebreak >}}
-
-![](figures/model-constituents.png)
-[**Figure 2**]{#fig-linear-constituents}.
-The RRi dynamics in response to exercise are expressed as a linear combination of model constituents based on the baseline RRi $\alpha$ and two logistic functions, denoted $f_1(t)$ and $f_2(t)$, respectively.
-The vertical dashed lines represent the time at which the exercise and recovery onset given by $\tau = 5$ and $\delta = 2$.
-
-{{< pagebreak >}}
-
-![](figures/2d-density-kernel.png)
-[**Figure 3**]{#fig-2d-kernel-density}.
-(A) Mean and SD from each subject's RRi recordings were used for the standardization process.
-(B) 2D kernel density of standardized RRi dynamics over time from a sample of individuals subjected to the rest-exercise-rest protocol.
-Darker colors indicate greater probability density.
-The contrary can be said about lighter colors.
-
-{{< pagebreak >}}
-
-![](figures/prior-model.png)
-[**Figure 4**]{#fig-prior-sim}.
-(A) Simulated standardized RRi dynamics based on prior parameter distributions, illustrating predicted RRi responses to exercise.
-Shaded areas represent 95%, 80%, and 60% quantile CI, offering insight into expected physiological variability across parameters.
-(B) Prior distributions and 95% CI were used to generate prior predictions based on physiological constraints and graphical visualization of standardized RRi data.
-
-{{< pagebreak >}}
-
-![](figures/group-level-estimates.png)
-[**Figure 5**]{#fig-group-level-estimates}.
-(A) Posterior probability distributions of the expectation for each population-parameter estimate ($E[\theta]$) with quantile-based 95% CI.
-(B) Transformed rate parameters into a percentage scale using the $1 - \exp(\theta)$ transformation.
-
-{{< pagebreak >}}
-
-![](figures/error-estimates.png)
-[**Figure 6**]{#fig-error-rates}.
-Individual-level performance metrics.
-(A) Bootstrapped MAPE and RMSE are metrics of relative and absolute model deviance from observed RRi.
-(B) Individual-level estimates of model performance and the relationship between them.
-(C) Partial autocorrelation function (ACF) of model residuals with corresponding quantile-based CI.
-
-{{< pagebreak >}}
 

### 2025-01-15 R/1-prior-rri-model.R

@@ -119,7 +119,9 @@ figure <- ggpubr::ggarrange(
   labels = c("A.", "B."), font.label = list(size = 12)
 )
 
-ggsave("figures/prior-model.pdf", figure, "pdf", 
+ggsave("figures/figure-4.pdf", figure, "pdf", 
        width = 7, height = 4, units = "in")
-ggsave("figures/prior-model.png", figure, "png", 
+ggsave("figures/figure-4.eps", figure, "eps", 
+       width = 7, height = 4, units = "in")
+ggsave("figures/figure-4.png", figure, "png", 
        width = 7, height = 4, units = "in", dpi = 500)

### 2025-01-15 R/2-model-constituents.R

@@ -32,7 +32,9 @@ figure <- ggplot(sim_data, aes(t, RRi)) +
         axis.text = element_text(size = rel(.8))) +
   labs(x = "Time (minutes)", y = "RRi (ms)")
 
-ggsave("figures/model-constituents.pdf", figure, "pdf", 
+ggsave("figures/figure-2.pdf", figure, "pdf", 
        width = 7, height = 4, units = "in")
-ggsave("figures/model-constituents.png", figure, "png", 
+ggsave("figures/figure-2.eps", figure, "eps", 
+       width = 7, height = 4, units = "in")
+ggsave("figures/figure-2.png", figure, "png", 
        width = 7, height = 4, units = "in", dpi = 500)

### 2025-01-15 R/3-sobol-sensitivity.R

@@ -34,8 +34,8 @@ sobol_sample <- function(n_samples) {
     c = runif(n_samples, min = 0.82, max = 0.86), 
     lambda = runif(n_samples, min = -3.16, max = -2.94), 
     phi = runif(n_samples, min = -2.71, max = -2.48),
-    tau = runif(n_samples, min = 3.05, max = 3.44), 
-    delta = runif(n_samples, min = 1, max = 3)
+    tau = runif(n_samples, min = 6.61, max = 6.81), 
+    delta = runif(n_samples, min = 3.05, max = 3.44)
   )
   
   return(out)
@@ -106,7 +106,7 @@ if (!file.exists("R/sobol-sensitivity.RDS")) { ## Omit computation if already ex
 }
 
 sobol_results[
-  j = list(estimate = median(mean), se = median(se)) |> lapply(round, 5), 
+  j = list(estimate = mean(mean), se = mean(se)) |> lapply(round, 5), 
   by = variable
   ][, `:=`(
     lower = round(estimate - se * qnorm(.975), 5),

### 2025-01-15 R/5-2d-density-kernel.R

@@ -50,7 +50,9 @@ figure <- ggpubr::ggarrange(
   hjust = -0.2, vjust = 2
 )
 
-ggsave("figures/2d-density-kernel.pdf", figure, "pdf", 
+ggsave("figures/figure-3.pdf", figure, "pdf", 
        width = 7, height = 4, units = "in")
-ggsave("figures/2d-density-kernel.png", figure, "png", 
+ggsave("figures/figure-3.eps", figure, "eps", 
+       width = 7, height = 4, units = "in")
+ggsave("figures/figure-3.png", figure, "png", 
        width = 7, height = 4, units = "in", dpi = 500)

### 2025-01-15 R/8-error-estimates.R

@@ -35,7 +35,7 @@ boot_error[, mc_sample := seq_len(.N)]
 # -------------------------------------------------------------------------
 
 f_1 <- ggplot(error_data, aes(mape, rmse)) +
-  geom_point(col = "gray20", alpha = .5) +
+  geom_point(col = "gray20", cex = 3/4) +
   scale_y_continuous(name = "RMSE (log-scale)", expand = c(0,0.05),
                      breaks = c(1,2,4,9,20)*10,
                      transform = "log10") +
@@ -108,7 +108,9 @@ figure <-
 
 
 
-ggsave("figures/error-estimates.pdf", figure, "pdf", 
+ggsave("figures/figure-6.pdf", figure, "pdf", 
        width = 7, height = 5, units = "in")
-ggsave("figures/error-estimates.png", figure, "png", 
+ggsave("figures/figure-6.eps", figure, "eps", 
+       width = 7, height = 5, units = "in")
+ggsave("figures/figure-6.png", figure, "png", 
        width = 7, height = 5, units = "in", dpi = 500)

### 2025-01-15 R/9-group-level-estimates.R

@@ -55,9 +55,11 @@ figure <- ggpubr::ggarrange(
   vjust = c(1.5, 2.5)
 )
 
-ggsave("figures/group-level-estimates.pdf", figure, "pdf", 
+ggsave("figures/figure-5.pdf", figure, "pdf", 
        width = 7, height = 5, units = "in")
-ggsave("figures/group-level-estimates.png", figure, "png", 
+ggsave("figures/figure-5.eps", figure, "eps", 
+       width = 7, height = 5, units = "in")
+ggsave("figures/figure-5.png", figure, "png", 
        width = 7, height = 5, units = "in", dpi = 500)
 
 # Plot data ---------------------------------------------------------------

### 2025-01-15 R/example-data.R

@@ -17,8 +17,10 @@ figure <- ggplot(rri_data[id %in% c(1:3,7:9)], aes(time, rr_denoised)) +
   theme(strip.background = element_blank(), strip.text = element_blank())
   
 ## Save the plot
-ggsave("figures/example-data.pdf", figure, "pdf", 
+ggsave("figures/figure-1.pdf", figure, "pdf", 
        width = 7, height = 5, units = "in")
-ggsave("figures/example-data.png", figure, "png", 
+ggsave("figures/figure-1.eps", figure, "eps", 
+       width = 7, height = 5, units = "in")
+ggsave("figures/figure-1.png", figure, "png", 
        width = 7, height = 5, units = "in", dpi = 500)
   

### 2025-02-12 response_to_reviewers.qmd
2nd round of reviews
- reviewer 2 ask to add a bit more dicussion on stochastic and deterministic models.
- editor and reviewer 1 had no more comments.
- hope this is the final round of reviews!

---
title: ""
format: 
  docx:
    reference-doc: "ref.docx"
bibliography: "misc/ref.bib"
csl: "misc/scientific-reports.csl"
editor_options: 
  chunk_output_type: console
execute: 
  echo: false
  warning: false
  error: false
---

# Response to Reviewers

## Overview

We thank the Reviewers for their valuable feedback and constructive comments up to this point, which have helped us to significantly improve the quality and clarity of our manuscript. Below, we provide a detailed response to the comments from reviewer 2.

## Reviewer 2

**Comment**: "A lot of determinism has been previously demonstrated, where the respiration is spontaneous, e.g. as in Physica A, 283, no. 3-4 (2000): 451-461; and Philos. Trans. R. Soc. A, 371, no. 1997 (2013): 20110622;"

**Response**: We appreciate the reviewer’s suggestion and acknowledge the importance of prior deterministic models in characterizing cardio-respiratory interactions. In response, we have included a discussion in our manuscript that situates our model within this deterministic framework, highlighting how logistic functions effectively capture structured, time-dependent fluctuations in RR intervals. We have also cited the suggested references to acknowledge their contributions to the field.

**Comment**: "Similarly, if a complementary approach is taken, then the fluctuations of the cardio-respiratory interactions have been studied also through a stochastic approach, eg. as in Eur. Phys. J. B, 65 (2008): 425-433"

**Response**: We recognize the significance of stochastic modeling in capturing the inherent variability of cardio-respiratory interactions. While our current model is primarily deterministic, it does account for inter-individual variability, which aligns with aspects of stochastic approaches. We have now included a reference to prior stochastic studies and briefly discussed how future extensions of our model could incorporate stochastic elements to enhance its adaptability in more variable physiological conditions.

**Comment**: "The authors could reflect also these important works, if they have further opportunity in the revision process."

**Response**: We appreciate the reviewer’s insightful suggestion. In response, we have revised the discussion section to include references to both deterministic and stochastic modeling approaches, clarifying how our work aligns with and builds upon these perspectives. We believe this addition strengthens the manuscript by providing a broader context for our modeling approach.

### 2025-02-12 paper.qmd
2nd round of reviews
- reviewer 2 ask to add a bit more dicussion on stochastic and deterministic models.
- editor and reviewer 1 had no more comments.
- hope this is the final round of reviews!

@@
 
-**Title**: Enhancing Cardiovascular Monitoring: A Non-Linear Approach to RR Interval Dynamics in Exercise and Recovery.
+**Title**: Enhancing Cardiovascular Monitoring: A Non-Linear Model for Characterizing RR Interval Fluctuations in Exercise and Recovery.
 
-**Authors**: Matías Castillo-Aguilar^1,2^, Diego Mabe-Castro^1,2^, David Medina^4,5^, Cristian Núñez-Espinosa^1,3^\*.
+**Authors**: Matías Castillo-Aguilar^1^, Diego Mabe-Castro^1,2^, David Medina^4,5^, Cristian Núñez-Espinosa^1,3^\*.
 
 ::: {custom-style="Afiliation"}
-^1^ Centro Asistencial Docente e Investigación (CADI-UMAG), University of Magallanes, Punta Arenas, Chile.
+^1^ Centro Asistencial Docente e Investigación (CADI-UMAG), Universidad de Magallanes, Punta Arenas, Chile.
 
-^2^ Kinesiology Department, University of Magallanes, Punta Arenas, Chile.
+^2^ Departamento de Kinesiología, Universidad de Magallanes, Punta Arenas, Chile.
 
-^3^ School of Medicine, University of Magallanes, Punta Arenas, Chile.
+^3^ Escuela de Medicina, Universidad de Magallanes, Punta Arenas, Chile.
 
@@
 
-Cristian Núñez-Espinosa, School of Medicine, University of Magallanes, Punta Arenas, Chile.
+Cristian Núñez-Espinosa, Escuela de Medicina, Universidad de Magallanes, Punta Arenas, Chile.
 Centro Asistencial de Docencia e Investigación CADI-UMAG, Chile.
@@
 
-This work aimed to develop and validate a novel non-linear model to characterize RR interval (RRi) dynamics throughout a rest-exercise-recovery protocol, offering a more precise and physiologically relevant representation of cardiac autonomic responses than traditional HRV metrics or linear approaches.
-Using data from a cohort of 272 elderly participants, the model employs logistic functions to capture the non-stationary and transient nature of RRi dynamics, with parameter estimation achieved via Hamiltonian Monte Carlo.
+This work aimed to develop and validate a novel non-linear model to characterize RR interval (RRi) time-dependent fluctuations throughout a rest-exercise-recovery protocol, offering a more precise and physiologically relevant representation of cardiac autonomic responses than traditional HRV metrics or linear approaches.
+Using data from a cohort of 272 elderly participants, the model employs logistic functions to capture the non-stationary and transient nature of RRi time-dependent fluctuations, with parameter estimation achieved via Hamiltonian Monte Carlo.
 Sobol sensitivity analysis identified baseline RRi ($\alpha$) and recovery proportion ($c$) as the primary drivers of variability, underscoring their critical roles in autonomic regulation and resilience.
-Validation against real-world RRi data demonstrated robust model performance, accurately reflecting autonomic recovery and exercise-induced fluctuations.
+Validation against real-world RRi data demonstrated robust model performance ($R^2$ = 0.868, CI~95%~[0.834, 0.895] and
+Root Mean Square Error [RMSE] = 32.6 ms, CI~95%~[30.01, 35.77]), accurately reflecting autonomic recovery and exercise-induced fluctuations.
 By advancing real-time cardiovascular assessments, this framework holds significant potential for clinical applications in rehabilitation and cardiovascular monitoring in athletic contexts to optimize performance and recovery.
@@
 
-Current research has extensively examined the mechanisms underlying cardiac autonomic dynamics in response to exercise and their links to health-related quality of life and cardiovascular disease risk [@eser2022exercise;@castillo2023validity;@mabe2024associations].
+The human cardiovascular system exhibits intricate dynamic responses to physical exertion, reflecting the complex interplay between the autonomic nervous system (ANS) and cardiac function.
+Understanding these time-dependent fluctuations is crucial for assessing physiological adaptation to exercise, optimizing athletic performance, and evaluating cardiovascular health [@eser2022exercise; @castillo2023validity; @mabe2024associations].
+R-R intervals (RRi), representing the beat-to-beat time intervals between successive heartbeats, provide a direct, high-resolution reflection of cardiac electrical activity.
+Unlike aggregated measures of heart rate variability (HRV), which summarize autonomic activity over longer periods and can mask transient fluctuations, RRi analysis offers a granular, beat-to-beat perspective on autonomic modulation during exercise and recovery [@kristal1995heart; @thayer2010relationship; @dong2016role; @lundstrom2023practices].
 
-In this context, the study of R-R intervals (RRi), defined as the time intervals between heartbeats and their link with exercise, has emerged as an important research area, given its relevance to cardiovascular health, athletic performance, and physiological adaptation [@kristal1995heart; @thayer2010relationship; @dong2016role; @lundstrom2023practices].
-Hence, analyzing the temporal dynamics of RRi in response to exercise can provide valuable insights into how the cardiovascular system adapts to physical stressors, such as exercise-induced fatigue and competition-related strain [@castillo2021cardiac; @eser2022exercise].
+This granular perspective is particularly relevant in dynamic exercise scenarios, where rapid shifts in autonomic balance occur, and in specific populations such as older adults, where age-related changes in autonomic function may influence cardiac responses [@mongin2022decrease; @castillo2023validity; @mabe2024associations].
+Analyzing RRi allows for examining immediate cardiac responses to exercise-induced stress, providing valuable insights into the efficiency and adaptability of the cardiovascular system.
 
-Unlike heart rate variability (HRV), which aggregates autonomic responses over time, RRi analysis provides a more granular, direct view of cardiac electrical activity during or immediately following exercise, particularly in older adults [@mongin2022hrv; @castillo2023validity; @mabe2024associations].
+While many studies have investigated cardiovascular responses to exercise using quasi-stationary protocols, simplifying analysis by minimizing non-stationarities [@storniolo2021symbolic; @porta2018relevance], these approaches may not fully capture physiological responses' dynamic and continuous nature during real-world activities.
+Although traditional linear methods like time-series analysis and linear regression have been employed to model RRi behavior [@lian2011simple], they often fall short in capturing the complex, non-linear time-dependent fluctuations of RRi transitions, particularly during periods of intense exertion and the subsequent recovery phase [@gronwald2019effects].
+This limitation is significant because the ANS undergoes rapid and non-linear shifts between parasympathetic withdrawal and sympathetic activation during exercise, resulting in intricate RRi fluctuations that linear models cannot adequately represent by their nature [@boettger2010heart].
+These rapid autonomic adjustments, including vagal tone and sympathetic outflow changes, contribute to the non-linear patterns observed in RRi data [@kristal1995heart; @thayer2010relationship].
+Consequently, these simplified models may miss critical physiological information related to cardiovascular adaptation, such as the speed and extent of recovery [@hautala2003short].
 
-Thus, understanding RRi fluctuations in response to exercise is particularly relevant during dynamic exercise periods, where the autonomic nervous system (ANS) shifts between parasympathetic withdrawal and sympathetic activation [@boettger2010heart].
-Therefore, directly modeling RRi dynamics, rather than relying on broader HRV metrics, allows for a direct assessment of physiological markers of autonomic adaptation to stress [@hautala2003short].
-This approach is valuable for identifying recovery patterns and understanding cardiovascular reactivity across individuals with various fitness levels [@mongin2022hrv].
+Model-based approaches, particularly those employing exponential functions, have been widely used to estimate heart rate and RRi recovery time constants after exercise [@imai1994vagally; @borresen2008autonomic; @pierpont2000heart; @pierpont2004assessing; @buchheit2007parasympathetic; @peccanha2014heart].
+While these models provide valuable insights into recovery kinetics, they often focus on specific phases of the exercise-recovery cycle.
+They may not fully capture the continuous transitions in RRi from rest to exercise and back to baseline.
+Furthermore, these models often rely on simplifying assumptions about the underlying physiological mechanisms, which may limit their ability to represent individual variability across different exercise intensities and populations accurately.
+Other models, like advanced non-linear approaches, have been developed to address the limitations of linear methods like decision tree-based ensemble algorithms and convolutional neural networks [@wang2020automated; @lee2023real; @berrahou2024arrhythmia].
+More advanced techniques, such as non-linear mode decomposition [@iatsenko2015nonlinear; @iatsenko2016extraction], dynamical modeling [@kralemann2013vivo; @stankovski2012inference], and the explicit consideration of non-autonomous dynamics [@clemson2014discerning; @lehnertz2024time], have also been applied to analyze physiological time series.
 
-Modeling the RRi behavior has traditionally been approached by leveraging linear regression and time-series analysis [@lian2011simple].
-However, this oppose significant challenges, like oversimplifying overall exercise dynamics without capturing the intricacies of exercise-induced intricate transitions in RRi, especially under intense exertion and recovery phases [@gronwald2019effects].
-More recently, advanced non-linear approaches have been developed to address the limitations of linear methods like decision tree-based ensemble algorithms and convolutional neural networks [@wang2020automated;@lee2023real;@berrahou2024arrhythmia].
-However, many more advanced alternatives fail to generate physiologically meaningful model parameters without directly linking to biological processes [@gronwald2019non].
+However, many of these existing non-linear models' lack of a direct link to underlying physiological processes is a significant limitation.
+While they may provide a better fit to the observed data, they often lack clear physiological interpretability, limiting their clinical utility and hindering a deeper understanding of the mechanisms driving RRi changes [@gronwald2019non; @bacopoulou2021entropy; @da2024post].
+Furthermore, few models are designed to capture the continuous, beat-to-beat transitions in RRi throughout the entire rest-exercise-recovery cycle while simultaneously providing physiologically meaningful parameters that can explain individual variability across diverse exercise intensities, durations, and populations [@kanniainen2023estimation].
+This gap hinders a comprehensive understanding of how individuals adapt to exercise and how these adaptations might differ based on age, fitness level, or underlying health status.
+For example, understanding how RRi time-dependent fluctuations differ between trained athletes and sedentary individuals during and after exercise could provide valuable insights for personalized training programs and rehabilitation strategies.
 
-Alongside this line of inquiry, recent studies have begun exploring non-linear models for RRi dynamics, recognizing their potential to capture the complexity of cardiovascular response to exercise [@da2024post].
-Exponential decay models, for example, have been proposed to describe RRi recovery [@sahoo2024unanticipated], while logistic functions have been used to model the gradual return to baseline after high-intensity exercise [@molkkari2020online; @gronwald2019non].
-These models offer advantages over traditional HRV metrics by providing a more detailed understanding of the cardiovascular system's response to exercise [@wu2003nonlinear].
-However, despite these advancements, few models are specifically designed to capture real-time RRi fluctuations, and even fewer consider physiologically meaningful parameters that allow modeling individual variability across a myriad of exercise dimensions [@kanniainen2023estimation].
-
-Considering the current state of the art, there is a clear need to develop a non-linear model that accurately represents RRi's non-linear transitions during exercise and recovery, where the estimation of clinically significant parameters is compelling.
-Such a model would need to offer a more physiologically relevant representation of the heart's behavior compared to the broader HRV indices commonly used in research [@bacopoulou2021entropy].
-This model must be complex enough to capture the non-linear, exercise-induced cardiovascular dynamics.
-But, simultaneously, it has to be simple enough to provide practical and significant model parameters related to observed physiological processes.
-
-Hence, the primary objective of this paper is to present a novel non-linear model that characterizes continuous RRi transitions from rest to exercise and recovery.
-This model is designed to capture the non-linear changes in RRi, providing meaningful parameters that can enhance our understanding of the physiological processes underlying cardiovascular adaptation to exercise.
+Therefore, this paper introduces a novel non-linear model designed to characterize the continuous RRi transitions from rest to exercise and recovery.
+This model aims to address the limitations of existing approaches by (1) accurately capturing the non-linear time-dependent fluctuations of RRi fluctuations throughout the entire rest-exercise-recovery cycle, providing a more complete picture of cardiovascular responses to exercise, and (2) providing physiologically interpretable parameters that reflect the underlying autonomic mechanisms, allowing for a more mechanistic understanding of individual adaptations.
+By focusing on these key aspects, this model offers a more detailed and physiologically relevant understanding of cardiovascular adaptation to exercise compared to traditional HRV metrics, with potential applications in personalized exercise prescription, performance monitoring, and clinical assessment of cardiovascular health.
 
@@
 
-To further assess the performance of the proposed model, real-world RRi data were analyzed in addition to the synthetic data generated through simulation.
+To further assess the proposed model's performance, real-world RRi data were analyzed in addition to the synthetic data generated through simulation.
+The dataset consisted of 272 participants who underwent a validated exercise protocol encompassing rest, exercise, and recovery phases within a single, continuous measurement session [@castillo2023validity].
+
+### Subjects
+
+Participants were recruited from a local community.
+Subjects were included if (i) they were aged 60 years or older; (ii) were permanently residing in the Magallanes and Chilean Antarctic region; (iii) had a percentage greater than 60% on the Karnofsky Performance Status Scale, which allowed us to work with older people who had a state of autonomy necessary to carry out the study tests; (iv) absence of the following diagnosis: diabetic neuropathy; use of pacemakers; clinical depression; cognitive or motor disability; and dementia.
+The exclusion criteria were: (i) consumption of beta-blockers during the study, (ii) taking drugs or stimulant substances within 12 hours before the cardiac assessment, and (iii) having some degree of motor disability that prevented participants from moving around.
+No participants met the exclusion criteria.
 This dataset was derived from a cohort participating in the FONDECYT Project No.
@@
 
-The dataset consisted of 272 participants who underwent a validated exercise protocol encompassing rest, exercise, and recovery phases within a single, continuous measurement session [@castillo2023validity].
-Continuous heart rate data, including RRi, were collected using the Polar Team2 system (Polar®) application, which is capable of capturing dynamic fluctuations associated with varying exercise intensities and recovery.
+### Exercise protocol
+
+The exercise protocol consisted of the continuous measurement of RRi before, during, and immediately after the 2-minute step test, which is a part of the Senior Fitness Test protocol [@rikli2013senior].
+This functional cardiorespiratory test required each subject to march in place as many times as possible for 2 minutes.
+The participants were monitored throughout the assessment using cardiovascular measures (i.e., heart rate and blood pressure) to prevent adverse events during the exercise protocol.
+The evaluation protocol was estimated to last approximately 20 minutes for each subject.
+None of the participants expressed discomfort during the evaluation.
+Continuous heart rate data, including RRi, were collected using the Polar Team2 system (Polar®) application, capable of capturing dynamic fluctuations associated with varying exercise intensities and recovery.
+
+### Preprocessing of RRi data
 
@@
 
-This real-world dataset provided a critical context for validating the model's predictive capability against observed physiological responses, offering a robust foundation for understanding RRi dynamics under physical activity conditions.
+This real-world dataset provided a critical context for validating the model's predictive capability against observed physiological responses, offering a robust foundation for understanding RRi time-dependent fluctuations under physical activity conditions.
 
@@
 The gradient of the log-likelihood function for each parameter was computed during estimation using the brms R package (v2.21.0), which employs the Stan probabilistic programming language.
-Convergence of the HMC chains was assessed using standard diagnostics, including R-hat values, kept below 1.01 for all parameters [@vehtari2019rank], and effective sample sizes, targeted at a minimum of 1,000 for each parameter [@brms].
+Convergence of the HMC chains was assessed using standard diagnostics, including R-hat values, kept below 1.01 for all parameters [@vehtari2019rank], and effective sample sizes targeted at a minimum of 1,000 for each parameter [@brms].
 Trace plots were inspected to confirm stable mixing.
@@
 To enhance the exploration of parameter space, we performed a two-stage analysis: We assessed parameter values at the individual level, which we then used to estimate  population-level parameters.
+This hierarchical structure enables us to capture individual variability through subject-level random effects while estimating group-level effects across all parameters, thus providing estimates of subject- and population-level model parameters.
 
@@
 $${#eq-ind-std}
-This standardization allowed the model to focus on relative changes in RRi dynamics, independent of individual baseline differences.
+
+This standardization allowed the model to focus on relative changes in RRi time-dependent fluctuations independent of individual baseline differences.
 
@@
 $${#eq-indiv-level}
+
 where $\alpha_i$, $\beta_i$, $c_i$, $\lambda_i$, $\phi_i$, $\tau_i$, $\delta_i$ are the individual-specific model parameters and $\epsilon_{i,t} \sim \mathcal{N}(0, \sigma^2)$ is the residual error term at each time point $t$.
@@
 
-This hierarchical structure enables us to capture individual variability through subject-level random effects while estimating group-level effects across all parameters, thus providing estimates of subject- and population-level model parameters.
-
 ## Model Performance
 
-The primary performance metrics, estimated for each subject, included R^2^, root mean square error (RMSE), and mean absolute percentage error (MAPE), estimated for each subject.
+The primary statistical performance metrics, estimated for each subject, included R^2^, root mean square error (RMSE), and mean absolute percentage error (MAPE), estimated for each subject.
 Bootstrap resampling across each metric was performed to estimate the mean performance of the model and corresponding quantile-based 95% CI.
 
-Also, residual analysis were conducted to evaluate the model's accuracy in capturing RRi dynamics.
+Also, residual analysis were conducted to evaluate the model's accuracy in capturing RRi time-dependent fluctuations.
 Residuals were defined as the difference between observed and predicted RRi values.
 These residuals were analyzed for temporal structure and partial autocorrelation to ensure that no systematic patterns remained in the errors.
-This indicates that the model has sufficiently captured the underlying dynamics of the RRi response to exercise.
+This indicates that the model has sufficiently captured the underlying time-dependent fluctuations of the RRi response to exercise.
 
@@
 We implemented a Sobol sensitivity analysis using Monte Carlo simulations to assess the sensitivity of model parameters influencing RRi over time.
-Sobol index ($S_{\text{ind}}$) provide a measure of the proportion of the contribution of each parameter to the variance in RRi at each time point, and it was selected for its robustness in handling non-linear and non-monotonic relationships, which are intrinsic to RRi dynamics in response to exercise [@zhang2015sobol].
+Sobol index ($S_{\text{ind}}$) provide a measure of the proportion of the contribution of each parameter to the variance in RRi at each time point, and it was selected for its robustness in handling non-linear and non-monotonic relationships, which are intrinsic to RRi time-dependent fluctuations in response to exercise [@zhang2015sobol].
 
@@
 The 95% CI parameter values estimated from HMC-NUTS were then used as input ranges for $S_{\text{ind}}$ computation.
-Finally, the mean values of $S_{\text{ind}}$ over the 20-minute time span for each model parameter were estimated and reported, with their corresponding 95% CI using normal approximation based on estimated standard errors (SE).
+Finally, the mean values of $S_{\text{ind}}$ over the 20-minute time span for each model parameter were estimated and reported, with their corresponding 95% CI using a normal approximation based on estimated standard errors (SE).
 
@@
 
-RRi dynamics in response to exercise tends to follow a U-shaped form.
-The initial decrease in RRi is associated with exercise onset, related to an increased heart rate.
-After exercise cessation, an opposite increase in RRi is observed, which is associated with the cardiovascular recovery phase.
-In both cases, the drop and recovery phases occurs at different rates; some individuals experience a quick recovery in RRi after exercise; however, in some others, this slope is less steep.
+RRi time-dependent fluctuations in response to exercise tend to follow a U-shaped form.
+The initial decrease in RRi is associated with exercise onset and an increased heart rate.
+After exercise cessation, an opposite increase in RRi is observed, associated with the cardiovascular recovery phase.
+In both cases, the drop and recovery phases occur at different rates; some individuals experience a quick recovery in RRi after exercise; however, in some others, this slope is less steep.
 Additionally, the new baseline reached following exercise cessation is often below the RRi baseline before exercise.
 
-These hallmarks of RRi dynamics in response to exercise, highlight the complex and non-linear behavior of the cardiovascular response in the context of both rest and exercise conditions.
+These hallmarks of RRi time-dependent fluctuations in response to exercise highlight the complex and non-linear behavior of the cardiovascular response in the context of rest and exercise conditions.
 [Figure 1](#fig-rri-example) shows an example of RRi record data.
 
+<!---- FIGURE 1 ---->
+![](figures/figure-1.png)
+[**Figure 1**]{#fig-rri-example}.
+Example data of RRi recordings of 6 subjects over a 20-minute rest-exercise-recovery protocol in a sample of elderly individuals.
+The subject-level data shows the inter-individual variability of RRi time-dependent fluctuations in response to exercised-induced cardiovascular stress, with similar behavior and recovery trajectories over time.
 <!---- FIGURE 1 ---->
@@
 
-The process of deriving the final equation for modeling RRi fluctuations was guided by an iterative exploration of mathematical functions capable of capturing the observed dynamics.
-Initially, exponential and logarithmic functions were considered due to their simplicity and wide applicability in describing temporal changes.
+The process of deriving the final equation for modeling RRi fluctuations was guided by an iterative exploration of mathematical functions capable of capturing the observed time-dependent fluctuations.
+Initially, exponential and logarithmic functions were considered due to their simplicity and broad applicability in describing temporal changes.
 Exponential functions were hypothesized to capture the rapid initial adaptations of RRi post-exercise onset.
@@
 Logistic functions introduce parameters for growth rate and inflection point, allowing for precise control over the shape and timing of the transition between dynamic states.
-By using two coupled logistic functions, one to represent the initial decrease in RRi and a second inverted logistic function describing the recovery phase, we achieved a model structure that could flexibly reproduce the observed non-linear variations.
+By using two coupled logistic functions, one representing the initial decrease in RRi and a second inverted logistic function describing the recovery phase, we achieved a model structure that could flexibly reproduce the observed non-linear variations.
 
 This approach provided a biologically plausible representation, with parameters that directly correspond to identifiable physiological features, such as the rate of adaptation and recovery, the time to peak response, and the extent of deviation from baseline.
-The logistic function framework emerged as the optimal solution after systematic testing and evaluation against empirical data, ensuring that the model accurately captured the qualitative and quantitative aspects of RRi dynamics.
+The logistic function framework emerged as the optimal solution after systematic testing and evaluation against empirical data, ensuring that the model accurately captured the qualitative and quantitative aspects of RRi time-dependent fluctuations.
 
@@
 
-This model includes two logistic functions representing the RRi dynamics across exercise and recovery phases.
+This model includes two logistic functions representing the RRi time-dependent fluctuations across exercise and recovery phases.
 The first logistic term models the decrease in RRi during exercise, where the parameter $\beta$ denotes the magnitude of this decline.
@@
 Here, $c$ scales the magnitude of recovery relative to the initial decline represented by $\beta$, capturing the proportion of the decline regained during recovery.
-The rate at which RRi returns to baseline is controlled by $\phi$, and $\delta$ indicates the lag following the cessation of exercise, marking the beginning of recovery, respectively.
+The rate at which RRi returns to baseline is controlled by $\phi$, and $\delta$ indicates the lag following the cessation of exercise, marking the beginning of recovery.
 
-Additionally, the dynamics of RRi in response to physical exertion can be represented as a linear combination of a baseline RRi $\alpha$ and two logistic functions denoted as $f_1(t)$ and $f_2(t)$.
+Additionally, the time-dependent fluctuations of RRi in response to physical exertion can be represented as a linear combination of a baseline RRi $\alpha$ and two logistic functions denoted as $f_1(t)$ and $f_2(t)$.
 The function $f_1(t)$ models the initial decay in RRi following the initiation of exercise while $f_2(t)$ characterizes the recovery phase after exercise cessation.
@@
 <!---- FIGURE 2 ---->
+![](figures/figure-2.png)
+[**Figure 2**]{#fig-linear-constituents}.
+The RRi time-dependent fluctuations in response to exercise are expressed as a linear combination of model constituents based on the baseline RRi $\alpha$ and two logistic functions, denoted $f_1(t)$ and $f_2(t)$, respectively.
+The vertical dashed lines represent the time at which the exercise and recovery onset given by $\tau = 5$ and $\delta = 2$.
+<!---- FIGURE 2 ---->
 
@@
 
-The sample used to assess RRi dynamics consists of 272 subjects selected from a local community of elderly individuals. The sample characteristics can be seen in [Table 1](#tbl-sample-characteristics)
+The sample used to assess RRi time-dependent fluctuations consists of 272 subjects selected from a local community of elderly individuals.
+The sample characteristics can be seen in [Table 1](#tbl-sample-characteristics)
 
 <!---- TABLE 1 ---->
+|Characteristic |Overall        |Female         |Male           |
+|:--------------|--------------:|--------------:|--------------:|
+|Sex            |—              |217 (79.8%)    |55 (20.2%)     |
+|Age            |71.14 ± 6.03   |70.73 ± 6.27   |72.73 ± 4.7    |
+|SBP (mm hg)    |130.23 ± 17.07 |129.58 ± 17.37 |132.8 ± 15.69  |
+|DBP (mm hg)    |77.1 ± 9.58    |76.68 ± 9.83   |78.75 ± 8.4    |
+|MAP (mm hg)    |94.81 ± 10.69  |94.31 ± 10.95  |96.76 ± 9.45   |
+|PP (mm hg)     |53.14 ± 14.07  |52.9 ± 14.26   |54.05 ± 13.38  |
+|BMI            |30.66 ± 5.43   |30.7 ± 5.64    |30.53 ± 4.53   |
+|Weight (kg)    |75.06 ± 14.23  |73.88 ± 14.09  |79.69 ± 13.95  |
+|Height (cm)    |156.56 ± 9.18  |155.29 ± 8.46  |161.55 ± 10.24 |
 
-An initial graphical exploration of RRi dynamics (see [Figure 3](#fig-2d-kernel-density)) indicates a clear drop in RRi around the 5-7 minutes mark, associated with exercise-induced cardiovascular stress.
+[**Table 1**]{#tbl-sample-characteristics}.
+Sample characteristics from which continuous RRi monitoring data was collected during the rest-exercise-rest protocol.
+Data is presented as Mean ± standard deviation (SD).
+SBP, systolic blood pressure; DBP, diastolic blood pressure; MAP, mean arterial pressure; PP, pulse pressure; BMI, body mass index.
+<!---- TABLE 1 ---->
+
+An initial graphical exploration of RRi time-dependent fluctuations (see [Figure 3](#fig-2d-kernel-density)) indicates a clear drop in RRi around the 5-7 minutes, associated with exercise-induced cardiovascular stress.
 However, greater variability across individuals in post-exercise recovery can be observed.
 
+<!---- FIGURE 3 ---->
+![](figures/figure-3.png)
+[**Figure 3**]{#fig-2d-kernel-density}.
+(A) Mean and SD from each subject's RRi recordings were used for the standardization process.
+(B) 2D kernel density of standardized RRi dynamics over time from a sample of individuals subjected to the rest-exercise-rest protocol.
+Darker colors indicate greater probability density.
+The contrary can be said about lighter colors.
 <!---- FIGURE 3 ---->
@@
 
-Simulated standardized RRi dynamics based on prior parameter distributions are shown in [Figure 4](#fig-prior-sim).
+Simulated standardized RRi time-dependent fluctuations based on prior parameter distributions are shown in [Figure 4](#fig-prior-sim).
 
+<!---- FIGURE 4 ---->
+![](figures/figure-4.png)
+[**Figure 4**]{#fig-prior-sim}.
+(A) Simulated standardized RRi time-dependent fluctuations based on prior parameter distributions, illustrating predicted RRi responses to exercise.
+Shaded areas represent 95%, 80%, and 60% quantile CI, offering insight into expected physiological variability across parameters.
+(B) Prior distributions and 95% CI were used to generate predictions based on physiological constraints and graphical visualization of standardized RRi data.
 <!---- FIGURE 4 ---->
@@
 <!---- TABLE 2 ---->
+| Parameter | Estimate^1^ | SE^1^ | Lower^2^ | Upper^2^ |
+|-----|---:|---:|---:|---:|
+| $\alpha$  |  861.78 | 5.73 |  850.57 |  872.85 |
+| $\beta$   | -345.49 | 7.41 | -359.81 | -330.97 |
+| $c$       |    0.84 | 0.01 |    0.82 |    0.86 |
+| $\lambda$ |   -3.05 | 0.06 |   -3.16 |   -2.94 |
+| $\phi$    |   -2.60 | 0.06 |   -2.71 |   -2.48 |
+| $\tau$    |    6.71 | 0.05 |    6.61 |    6.81 |
+| $\delta$  |    3.24 | 0.10 |    3.05 |    3.44 |
+| $\sigma$  |   27.57 | 0.57 |   26.45 |   28.70 |
+
+[**Table 2**]{#tbl-parameters}.
+Population-parameter values estimated from group-level analysis.
+^1^ Estimates and SE are computed as the posterior distribution's median and median absolute deviation, respectively; ^2^ Lower and Upper bounds from the quantile-based CI~95%~ of the posterior distribution.
+<!---- TABLE 2 ---->
 
@@
 
+<!---- FIGURE 5 ---->
+![](figures/figure-5.png)
+[**Figure 5**]{#fig-group-level-estimates}.
+(A) Posterior probability distributions of the expectation for each population-parameter estimate ($E[\theta]$) with quantile-based 95% CI.
+(B) Transformed rate parameters into a percentage scale using the $1 - \exp(\theta)$ transformation.
 <!---- FIGURE 5 ---->
@@
 
-Estimated through bootstrapped resampling, relative performance metrics suggest that the model tends to deviate a 3.4% (CI~95%~[3.06, 3.81]) from the observed RRi data.
+Estimated through bootstrapped resampling, relative statistical performance metrics suggest that the model tends to deviate by 3.4% (CI~95%~[3.06, 3.81]) from the observed RRi data.
 This is equivalent to a 32.6 ms in the RRi scale (CI~95%~[30.01, 35.77]).
@@
 <!---- FIGURE 6 ---->
+![](figures/figure-6.png)
+[**Figure 6**]{#fig-error-rates}.
+Individual-level performance metrics.
+(A) Bootstrapped MAPE and RMSE are statistical metrics of relative and absolute model deviance from observed RRi.
+(B) Individual-level estimates of model performance and the relationship between them.
+(C) Partial autocorrelation function (ACF) of model residuals with corresponding quantile-based CI.
+(D) Example data with model estimates of RRi fluctuations and corresponding quantile-based CI initially displayed.
+<!---- FIGURE 6 ---->
 
@@
 
-Individual perturbation of each parameter highlighted that RRi dynamics are sensitive to the baseline RRi parameter, $\alpha$.
+Individual perturbation of each parameter highlighted that RRi time-dependent fluctuations are sensitive to the baseline RRi parameter, $\alpha$.
 Conversely, the rate parameters for the initial decay during exercise, $\lambda$, and the recovery post-exercise, $\phi$, show lower sensitivity, suggesting that they are not primary sources of variation in predicted RRi trajectories when assessed in isolation.
@@
 <!---- TABLE 3 ---->
+|Parameter | Estimate^1^|  SE^1^ | Lower^2^| Upper^2^|
+|:---------|-----------:|-------:|--------:|--------:|
+|$\alpha$  |     0.61329| 0.01756|  0.57887|  0.64771|
+|$\beta$   |     0.06651| 0.00286|  0.06090|  0.07212|
+|$c$       |     0.18939| 0.00815|  0.17342|  0.20536|
+|$\lambda$ |     0.00147| 0.00007|  0.00133|  0.00161|
+|$\phi$    |     0.00160| 0.00008|  0.00144|  0.00176|
+|$\tau$    |     0.04982| 0.00172|  0.04645|  0.05319|
+|$\delta$  |     0.07896| 0.00239|  0.07428|  0.08364|
+
+[**Table 3**]{#tbl-sens-params}.
+Estimated $S_{\text{ind}}$ of model parameters.
+^1^ Estimates and SE are computed as mean and standard deviation of Monte Carlo samples, respectively.
+Each parameter's $S_{\text{ind}}$ reflects a uniform variation from the 95% CIs of the estimated parameter values.
+<!---- TABLE 3 ---->
 
@@
 
-To the best of our knowledge, this study represents the first attempt to develop a non-linear model specifically designed to capture RRi dynamics across a complete rest-exercise-rest protocol continuously.
+To our knowledge, this study represents the first attempt to develop a non-linear model specifically designed to capture RRi time-dependent fluctuations continuously across a complete rest-exercise-rest protocol.
 Previous studies have either focused on aggregate HRV indices or utilized simplified linear or exponential models, which are insufficient to describe the complex, non-stationary transitions observed during and after exercise [@molkkari2020online].
-By employing a combination of logistic functions, our model uniquely accounts for the gradual shifts in autonomic regulation denoted by RRi dynamics, offering a detailed and physiologically relevant representation of cardiac dynamics.
+By employing a combination of logistic functions, our model uniquely accounts for the gradual shifts in autonomic regulation denoted by RRi time-dependent fluctuations, offering a detailed and physiologically relevant representation of cardiac dynamics.
 This continuous modeling framework integrates exercise-induced RRi decline and post-exercise recovery within a single unified structure, bridging a critical gap in the current literature.
-Such an approach advances our understanding of cardiovascular responses and opens new avenues for real-time monitoring and intervention in both clinical and athletic settings.
+Such an approach advances our understanding of cardiovascular responses and opens new avenues for real-time monitoring and intervention in clinical and athletic settings.
 
 The proposed model demonstrates a precise capacity to reproduce RRi dynamics.
-With its combination of logistic functions, it captures the key transitions of cardiac response, the initial decline during exercise, and the subsequent recovery.
-This design accommodates the inherent non-linearity and non-stationarity of RRi dynamics, overcoming the limitations of linear models and exponential functions commonly used in prior studies [@silva2019exponential; @mongin2022decrease].
+Its combination of logistic functions captures the key transitions of cardiac response, the initial decline during exercise, and the subsequent recovery.
+This design accommodates the inherent non-linearity and non-stationarity of RRi time-dependent fluctuations, overcoming the limitations of linear models and exponential functions commonly used in prior studies [@silva2019exponential; @mongin2022decrease].
 
 Compared to previous research, our findings align with efforts to capture nonlinear dynamics in HRV to understand cardiac responses during exercise [@gronwald2019effects].
-Similarly, previous studies have shown that dynamic fluctuations in RRi can serve as critical indicators of cardiorespiratory fitness.
-This supports the need for models to address the complexity of cardiovascular responses during physical stress [@mongin2022hrv].
-However, while many existing models focus primarily on linear metrics or aggregate HRV measures, our study provides a high-resolution analysis of RRi dynamics that enhances interpretability and application across diverse fitness levels and exercise intensities.
+Similarly, previous studies have shown that dynamic fluctuations in RRi can serve as critical indicators of cardiorespiratory fitness [@lundstrom2023practices; @mongin2022decrease].
+This supports the need for models to address the complexity of cardiovascular responses during physical stress [@mongin2022decrease].
+However, while many existing models focus primarily on linear metrics or aggregate HRV measures, our study provides a high-resolution analysis of RRi time-dependent fluctuations that enhances interpretability and application across diverse fitness levels and exercise intensities.
+Critically, many model-based approaches, particularly those employing exponential functions, have been used to estimate time constants of heart rate and RRi recovery after exercise [@borresen2008autonomic; @imai1994vagally; @pierpont2000heart; @pierpont2004assessing; @buchheit2007parasympathetic; @peccanha2014heart].
+These models often focus on characterizing the recovery phase and may not capture the continuous transitions from rest to peak exercise and subsequent recovery.
+Our model, by contrast, provides a unified framework for modeling the entire rest-exercise-recovery cycle, allowing for the estimation of parameters that reflect both the exercise-induced changes in RRi and the subsequent recovery dynamics.
+This continuous modeling approach provides a more comprehensive picture of cardiovascular response to exercise than models focusing solely on recovery kinetics.
 
@@
 
+Prior studies have examined cardiorespiratory interactions using both deterministic and stochastic approaches.
+Deterministic models have demonstrated that respiration-driven heart rate fluctuations exhibit structured, predictable behavior, suggesting an underlying regulatory mechanism of autonomic control [@lotrivc2000synchronization; @iatsenko2013evolution].
+Conversely, stochastic models emphasize the role of random variability in these interactions, accounting for inherent physiological fluctuations [@kenwright2008effect].
+Our non-linear model aligns with the deterministic perspective by employing logistic functions to characterize time-dependent RRi fluctuations while also incorporating inter-individual variability.
+Although this model does not explicitly integrate stochastic noise, it captures structured autonomic responses.
+Future work could explore the incorporation of stochastic elements to further enhance its applicability in more variable physiological conditions.
+
 Unlike prior research that aggregates HRV measures or applies simple decay models, our approach directly models RRi changes, offering richer physiological insight.
@@
 By integrating exercise and recovery phases, our model provides a more comprehensive view of autonomic regulation.
+Furthermore, it's important to note that the traditional "sympathovagal balance" hypothesis, which posits a reciprocal relationship between sympathetic and parasympathetic activity, may be oversimplified, especially during exercise [@storniolo2021symbolic].
+Recent evidence suggests that parasympathetic control can remain active even during periods of high sympathetic activation.
+By capturing the continuous time-dependent fluctuations of RRi, our model may provide insights into these complex interactions, potentially revealing nuances in autonomic control that are not captured by simpler models that assume a strict sympathovagal balance.
 
-Moreover, the sensitivity of parameters such as $\lambda$ (decay rate) and $(\phi$ (recovery rate) was found to be relatively low, suggesting that the model is robust to variability in these rates while remaining sensitive to key physiological parameters ($\alpha$ and $c$).
+Moreover, the sensitivity of parameters such as $\lambda$ (decay rate) and $\phi$ (recovery rate) was found to be relatively low, suggesting that the model is robust to variability in these rates while remaining sensitive to key physiological parameters ($\alpha$ and $c$).
 This robustness makes it suitable for individualized monitoring and population-level analyses, offering versatility in its application across different use cases.
@@
 
-However, the Sobol method assumes parameter independence, which may overlook interactions that are common in biological systems [@harenberg2017uncertainty; @cheng2017mixed; @herman2013method].
+However, the Sobol method assumes parameter independence, which may overlook interactions common in biological systems [@harenberg2017uncertainty; @cheng2017mixed; @herman2013method].
 For example, the interplay between $\lambda$ and $c$, which dictates the rate and magnitude of recovery, is likely critical but remains unexplored in the current framework.
 Future studies could explore Bayesian sensitivity analysis or variance decomposition methods that account for parameter interdependence [@bornn2010efficient; @xue2021bayesian].
+Furthermore, more advanced techniques, such as non-linear mode decomposition [@iatsenko2015nonlinear; @iatsenko2016extraction], dynamical modeling [@kralemann2013vivo; @stankovski2012inference], and the explicit consideration of non-autonomous dynamics [@clemson2014discerning; @lehnertz2024time], offer powerful tools for analyzing physiological time series.
+While these methods can capture complex dynamics, our model provides a more direct link to physiological interpretation through its parameters related to specific aspects of autonomic control.
+Future work could investigate how these approaches could be combined or compared to enhance our understanding of RRi time-dependent fluctuations.
 
@@
 In clinical contexts, it could aid in tailoring cardiovascular rehabilitation protocols by monitoring autonomic recovery in real-time, ensuring safe and effective exercise regimens for at-risk populations [@carrasco2024hrv].
-This aligns with previous research, which highlights the importance of individualizing rehabilitation programs to optimize recovery [@takahashi2020signs; @rizvi2023enhancing; @carrasco2024hrv].
+This aligns with previous research, highlighting the importance of individualizing rehabilitation programs to optimize recovery [@takahashi2020signs; @rizvi2023enhancing; @carrasco2024hrv].
 
 The model could guide training strategies in athletic settings, particularly for interval training, where determining optimal recovery periods is crucial.
-Similar findings suggest that precise monitoring of RRi dynamics can prevent overtraining and enhance performance [@hebisz2022heart; @nuuttila2024monitoring].
+Similar findings suggest that precise monitoring of RRi time-dependent fluctuations can prevent overtraining and enhance performance [@hebisz2022heart; @nuuttila2024monitoring].
 The model's ability to integrate real-time data from wearable devices further enhances its applicability in dynamic, uncontrolled environments, enabling field-based monitoring and feedback [@zimatore2020recurrence].
@@
 First, the assumption of uniform parameter sampling in sensitivity analysis, while practical, may not fully capture the variability observed in populations with extreme autonomic profiles [@kristal1995heart].
-Empirical distributions or Bayesian priors, could improve parameter estimation and enhance the model's applicability to diverse populations [@xue2021bayesian].
+Empirical distributions, or Bayesian priors, could improve parameter estimation and enhance the model's applicability to diverse populations [@xue2021bayesian].
+Bayesian inference could be a valuable extension of this work, particularly dynamic Bayesian inference [@duggento2012dynamical; @stankovski2014tutorial; @lukarski2022variability], specifically designed to model time-evolving dynamics.
+This approach could allow for the incorporation of prior knowledge about individual physiological characteristics and provide more robust estimates of the model parameters.
 
 Another limitation lies in the demographic composition of the sample, which consisted exclusively of elderly individuals.
-While this population provides valuable insights into age-specific cardiovascular dynamics, the findings may not fully generalize to younger populations, whose autonomic responses to exercise and recovery differ significantly due to higher baseline vagal tone, greater cardiac plasticity, and distinct metabolic profiles [@castillo2021cardiac; @garavaglia2021effect].
-Previous studies have demonstrated that younger individuals exhibit faster autonomic recovery and greater adaptability during physical exertion compared to older populations, suggesting that the parameter estimates derived from this model may vary across age groups [@choi2020declining; @garavaglia2021effect].
+While this population provides valuable insights into age-specific cardiovascular time-dependent fluctuations, the findings may not fully generalize to younger populations, whose autonomic responses to exercise and recovery differ significantly due to higher baseline vagal tone, greater cardiac plasticity, and distinct metabolic profiles [@castillo2021cardiac; @garavaglia2021effect].
+Previous studies have demonstrated that younger individuals exhibit faster autonomic recovery and greater adaptability during physical exertion compared to older populations [@choi2020declining; @garavaglia2021effect].
+This suggests that the parameter estimates derived from this model may vary across age groups [@choi2020declining; @garavaglia2021effect].
 Future research should validate the model in more diverse cohorts, including younger adults and athletes, to ensure broader applicability and to explore potential age-dependent modifications of the model's parameters.
@@
 
-Additionally, this study did not explicitly consider environmental and psychological factors, such as temperature, stress, or sleep quality.
+Furthermore, the uneven sex ratio in our sample (79.8% female, 20.2% male) is another limitation that should be addressed in future studies.
+Sex differences in autonomic control have been reported [@castillo2021cardiac], and this imbalance could have influenced our results.
+Future research should strive for a more balanced sex ratio to minimize potential bias and explore sex-specific differences in RRi time-dependent fluctuations during exercise and recovery.
+This study did not explicitly consider environmental and psychological factors like temperature, stress, or sleep quality.
 Future work could integrate these variables into the model, enhancing its robustness and applicability across varied real-world scenarios.
@@
 
-In summary, this study presents a novel non-linear model for RRi dynamics, capturing the complex and transient autonomic responses during rest-exercise-recovery protocols, overcoming the limitations of traditional autonomic metrics.
+In summary, this study presents a novel non-linear model for RRi time-dependent fluctuations, capturing the complex and transient autonomic responses during rest-exercise-recovery protocols, overcoming the limitations of traditional autonomic metrics.
 The model emphasizes their critical roles in reflecting autonomic regulation and resilience by identifying baseline RRi and recovery proportion as the dominant contributors to variability.
@@
 This work was funded by ANID Proyecto Fondecyt Iniciación Nº11220116.
-
-{{< pagebreak >}}
-
-# Figures
-
-![](figures/example-data.png)
-[**Figure 1**]{#fig-rri-example}.
-Example data of RRi recordings of 6 subjects over a 20-minute rest-exercise-recovery protocol in a sample of elderly individuals.
-The subject-level data shows the inter-individual variability of RRi dynamics in response to exercised-induced cardiovascular stress, with similar behavior and recovery trajectories over time.
-
-{{< pagebreak >}}
-
-![](figures/model-constituents.png)
-[**Figure 2**]{#fig-linear-constituents}.
-The RRi dynamics in response to exercise are expressed as a linear combination of model constituents based on the baseline RRi $\alpha$ and two logistic functions, denoted $f_1(t)$ and $f_2(t)$, respectively.
-The vertical dashed lines represent the time at which the exercise and recovery onset given by $\tau = 5$ and $\delta = 2$.
-
-{{< pagebreak >}}
-
-![](figures/2d-density-kernel.png)
-[**Figure 3**]{#fig-2d-kernel-density}.
-(A) Mean and SD from each subject's RRi recordings were used for the standardization process.
-(B) 2D kernel density of standardized RRi dynamics over time from a sample of individuals subjected to the rest-exercise-rest protocol.
-Darker colors indicate greater probability density.
-The contrary can be said about lighter colors.
-
-{{< pagebreak >}}
-
-![](figures/prior-model.png)
-[**Figure 4**]{#fig-prior-sim}.
-(A) Simulated standardized RRi dynamics based on prior parameter distributions, illustrating predicted RRi responses to exercise.
-Shaded areas represent 95%, 80%, and 60% quantile CI, offering insight into expected physiological variability across parameters.
-(B) Prior distributions and 95% CI were used to generate prior predictions based on physiological constraints and graphical visualization of standardized RRi data.
-
-{{< pagebreak >}}
-
-![](figures/group-level-estimates.png)
-[**Figure 5**]{#fig-group-level-estimates}.
-(A) Posterior probability distributions of the expectation for each population-parameter estimate ($E[\theta]$) with quantile-based 95% CI.
-(B) Transformed rate parameters into a percentage scale using the $1 - \exp(\theta)$ transformation.
-
-{{< pagebreak >}}
-
-![](figures/error-estimates.png)
-[**Figure 6**]{#fig-error-rates}.
-Individual-level performance metrics.
-(A) Bootstrapped MAPE and RMSE are metrics of relative and absolute model deviance from observed RRi.
-(B) Individual-level estimates of model performance and the relationship between them.
-(C) Partial autocorrelation function (ACF) of model residuals with corresponding quantile-based CI.
-
-{{< pagebreak >}}
-
-# Tables
-
-|Characteristic |Overall        |Female         |Male           |
-|:--------------|--------------:|--------------:|--------------:|
-|Sex            |—              |217 (79.8%)    |55 (20.2%)     |
-|Age            |71.14 ± 6.03   |70.73 ± 6.27   |72.73 ± 4.7    |
-|SBP (mm hg)    |130.23 ± 17.07 |129.58 ± 17.37 |132.8 ± 15.69  |
-|DBP (mm hg)    |77.1 ± 9.58    |76.68 ± 9.83   |78.75 ± 8.4    |
-|MAP (mm hg)    |94.81 ± 10.69  |94.31 ± 10.95  |96.76 ± 9.45   |
-|PP (mm hg)     |53.14 ± 14.07  |52.9 ± 14.26   |54.05 ± 13.38  |
-|BMI            |30.66 ± 5.43   |30.7 ± 5.64    |30.53 ± 4.53   |
-|Weight (kg)    |75.06 ± 14.23  |73.88 ± 14.09  |79.69 ± 13.95  |
-|Height (cm)    |156.56 ± 9.18  |155.29 ± 8.46  |161.55 ± 10.24 |
-
-[**Table 1**]{#tbl-sample-characteristics}.
-Sample characteristics from which continuous RRi monitoring data was collected during a rest-exercise-rest protocol.
-Data is presented as Mean ± standard deviation (SD).
-SBP, systolic blood pressure; DBP, diastolic blood pressure; MAP, mean arterial pressure; PP, pulse pressure; BMI, body mass index.
-
-{{< pagebreak >}}
-
-| Parameter | Estimate^1^ | SE^1^ | Lower^2^ | Upper^2^ |
-|-----|---:|---:|---:|---:|
-| $\alpha$  |  861.78 | 5.73 |  850.57 |  872.85 |
-| $\beta$   | -345.49 | 7.41 | -359.81 | -330.97 |
-| $c$       |    0.84 | 0.01 |    0.82 |    0.86 |
-| $\lambda$ |   -3.05 | 0.06 |   -3.16 |   -2.94 |
-| $\phi$    |   -2.60 | 0.06 |   -2.71 |   -2.48 |
-| $\tau$    |    6.71 | 0.05 |    6.61 |    6.81 |
-| $\delta$  |    3.24 | 0.10 |    3.05 |    3.44 |
-| $\sigma$  |   27.57 | 0.57 |   26.45 |   28.70 |
-
-[**Table 2**]{#tbl-parameters}.
-Population-parameter values estimated from group-level analysis.
-^1^ Estimates and SE are computed as median and mean absolute deviation of the posterior distribution, respectively; ^2^ Lower and Upper bounds from the quantile-based CI~95%~ of the posterior distribution.
-
-{{< pagebreak >}}
-
-|Parameter | Estimate^1^|  SE^1^ | Lower^2^| Upper^2^|
-|:---------|-----------:|-------:|--------:|--------:|
-|$\alpha$  |     0.61329| 0.01756|  0.57887|  0.64771|
-|$\beta$   |     0.06651| 0.00286|  0.06090|  0.07212|
-|$c$       |     0.18939| 0.00815|  0.17342|  0.20536|
-|$\lambda$ |     0.00147| 0.00007|  0.00133|  0.00161|
-|$\phi$    |     0.00160| 0.00008|  0.00144|  0.00176|
-|$\tau$    |     0.04982| 0.00172|  0.04645|  0.05319|
-|$\delta$  |     0.07896| 0.00239|  0.07428|  0.08364|
-
-[**Table 3**]{#tbl-sens-params}. Estimated $S_{\text{ind}}$ of model parameters. Each parameter's $S_{\text{ind}}$ reflects a uniform variation from the 95% CIs of the estimated parameter values.

### 2025-02-12 R/6-fit-model-to-data.R

@@ -93,7 +93,7 @@ fwrite(id_data, file = "models/id_level_posterior.csv")
 ## Generate data with estimated parameters and observed data
 long_data <- merge.data.table(
   m_data[, .(id, time, rr_denoised)],
-  id_data[, .SD, .SDcols = id:deltaMu],
+  id_data[, .SD, .SDcols = id:sigmaMu],
   by = "id",
 )
 
@@ -104,6 +104,50 @@ long_data[, pred_rr := RRi_sim(
   lambda = lambdaMu, phi = phiMu,
   tau = tauMu, delta = deltaMu
 ), id]
+
+
+# Re-fit problematic observations -----------------------------------------
+
+check_id <- logical(length = 272)
+for(i in seq_along(unique(error_phase$id))) {
+  long_data[id == i, j = {
+    plot(time, rr_denoised, type = "l", lwd = 1/2)
+    lines(time, pred_rr, type = "l", lwd = 1/2)
+    abline(v = c(tauMu, tauMu + deltaMu), lwd = 1/2, lty = 2)
+  }]
+  cat("Actual: ", i)
+  if (askYesNo(msg = "Todo bien?")) {
+    next
+  } else {
+    check_id[i] <- TRUE
+  }
+}
+
+## Problematic recordings
+ids <- c(19L, 21L, 26L, 27L, 34L, 50L, 52L, 86L, 100L, 102L, 105L, 132L, 
+         135L, 139L, 150L, 154L, 159L, 233L, 246L, 258L)
+
+models <- vector(mode = "list", length = length(ids))
+
+for (i in ids) {
+  print(paste0("Now on id ", i))
+  file <- paste0("models/id_level/m_id_",i,".RDS")
+  unlink(file)
+  
+  models[[i]] <- update(
+    object = m_model,
+    newdata = m_data[id == i],
+    seed = 12345,
+    iter = 10000, warmup = 5000,
+    chains = 5, cores = 5,
+    control = list(adapt_delta = 0.99,
+                   max_treedepth = 50),
+    sample_prior = "no",
+    file = file
+  )
+} 
+
+# Compute error metrics ---------------------------------------------------
 
 ## Compute error metrics
 long_data[, error := rr_denoised - pred_rr, id]

### 2025-02-12 R/8-error-estimates.R

@@ -12,7 +12,8 @@ error_data_long <- fread(file = "models/error_performance.csv")
 ## Prepare data ----
 
 ### Unique MAPE and RMSE for each id
-error_data <- error_data_long[, lapply(.SD, unique), id, .SDcols = c(14:16)]
+error_data <- error_data_long[, lapply(.SD, unique), id, .SDcols = c(15:17)]
+
 
 ### PACF for each id
 acf_data <- error_data_long[
@@ -98,7 +99,7 @@ f_3 <- ggplot(acf_data, aes(lag, acf)) +
   theme_classic(base_line_size = 1/4, base_size = 12) +
   theme(axis.text = element_text(size = rel(.8)))
 
-figure <- 
+fig_errors <- 
   ggpubr::ggarrange(
     ggpubr::ggarrange(
       ggpubr::ggarrange(f_2a, f_2b, f_2c, nrow = 1, ncol = 3),
@@ -107,10 +108,103 @@ figure <-
   )
 
 
+# Add predicted RRi -------------------------------------------------------
+
+# ID: id %in% c(1:3,7:9) 1,2,3,7,8,9
+
+f_pred_rr <- ggplot(error_data_long[id %in% c(1:3,7:9)], aes(time, rr_denoised, group = id)) +
+  facet_wrap(~ id, nrow = 2, ncol = 3) +
+  geom_ribbon(aes(ymin = pred_rr - sigmaMu * qnorm(.975), ymax = pred_rr + sigmaMu * qnorm(.975)), 
+              #alpha = .15, 
+              fill = "gray90"
+              ) +
+  geom_ribbon(aes(ymin = pred_rr - sigmaMu * qnorm(.9), ymax = pred_rr + sigmaMu * qnorm(.9)), 
+              #alpha = .15, 
+              fill = "gray80"
+              ) +
+  geom_ribbon(aes(ymin = pred_rr - sigmaMu * qnorm(.75), ymax = pred_rr + sigmaMu * qnorm(.75)), 
+              #alpha = .15, 
+              fill = "gray70"
+              ) +
+  geom_line(lwd = 1/4, na.rm = TRUE, col = "gray40") +
+  geom_line(aes(y = pred_rr), linewidth = 1/2) +
+  labs(x = "Time (min)", y = "RRi (ms)") +
+  theme_classic(base_line_size = 1/4, base_size = 12) +
+  theme(strip.background = element_blank(), 
[…]
